### 2017-06-12 paper.tex
first commit

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
%customized
\linespread{1.5}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%units
\usepackage{siunitx}
\DeclareSIUnit\Molar{\textsc{m}}
\sisetup{scientific-notation = true}
%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Simulating detailed Ca\textsuperscript{2+}-calmodulin-CaMKII network in a new way: biophysical attributes that affect CaMKII activation pattern} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Ximing Li\textsuperscript{1},
William Holmes\textsuperscript{1}
\\
\bigskip
\textbf{1} Department of Biological Sciences, Neuroscience Program, Ohio University, Athens, Ohio, United States of America
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Current address notes
%å\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

% Deceased author note
%\dag Deceased

% % Group/Consortium Author Note
% \textpilcrow Membership list can be found in the Acknowledgments section.

% % Use the asterisk to denote corresponding authorship and provide email address in note below.
% * correspondingauthor@institute.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons and understanding how this occurs has been the focus of many modeling studies. However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature and complex activation kinetics. To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion of the number of species. With this new method, spatial and temporal aspects of the signaling network can be studied without compromising biochemical details. We used this new method to examine how calmodulin molecules, both partially loaded and fully loaded with Ca\textsuperscript{2+}, choose pathways to interact with and activate CaMKII under various Ca\textsuperscript{2+} input conditions. We found that the frequency dependence of the CaMKII phosphorylation with Ca\textsuperscript{2+} input is not primarily determined by subunit number, but is intrinsic to the network and can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion. Depending on whether Ca\textsuperscript{2+} input is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns. The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca2+ channel nanodomains and cytoplasm.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}

Ca\textsuperscript{2+} signals are commonly used by cells for various types of activities. In neurons, Ca\textsuperscript{2+} can regulate gene expression and dendritic spine enlargement, strengthen synaptic connectivity, and promote neural growth or even death. These strikingly contrasting cell processes are possible because Ca\textsuperscript{2+} signals of varying strengths and temporal patterns are interpreted differently by intracellular signaling pathways. In our study, we focus on one particular pathway that involves calmodulin and Calcium/calmodulin-dependent protein kinase II (CaMKII). We have developed a new computational method that gets around the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods. Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca2+ input conditions, calmodulin availability and the Ca2+ diffusion rate.

\linenumbers
% Use "Eq" instead of "Equation" for equation citations.
% Camkii binding with F-actin is mostly reported in spines studies, but also seen in cytoplasm as Li Boxing 2016
%Bhalla's paper CaMKII model with diffusion, but simplified CaM states;
% setup with spine geometry; camca4 only; no diffusion effect; stochastic? hybrid? deterministic
\section*{Introduction}
Calcium/calmodulin-dependent protein kinase II (CaMKII) is an important enzyme widely distributed in the central nervous system and other tissues including cardiac muscles~\cite{Wayman:2008gla,Lisman:2002ki,Erickson:2014fs}. CaMKII is involved in intracellular Ca\textsuperscript{2+} signaling as an effector of the Ca\textsuperscript{2+} sensor protein calmodulin (CaM)~\cite{Herring:2016bh,Hell:2014bd,Coultrap:2012ip}. In neurons, CaMKII molecules are especially involved in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}. In spines, activated CaMKII molecules interact with postsynaptic density proteins, facilitating actins to reorganize, leading to spine enlargement and upregulation of $alpha$-amino-3-hydroxy-5-methyl-4-isoxazolepropionic acid (AMPA) receptor numbers~\cite{Herring:2016bh}. CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}. In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus. Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors. Therefore CaMKII molecules play a key role in excitation-transcription coupling.

The CaMKII molecule is a holoenzyme that consists of 12 subunits. Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase. Activities such as CaM binding and auto-phosphorylation result in various subunits states. For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site. Once phosphorylated, the CaM unbinding rate decreases dramatcially, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}. Even when the CaM molecule unbinds, the CaMKII subunit stays activated, entering an autonomous state. Finally, Thr305/Thr306 can also be phosphorylated. This theronine site overlaps with the CaM binding site and its phosphorylation blocks the binding of CaM molecules. In this case, the subunit becomes "capped".

Researchers have long been interested in modeling and simulating CaMKII~\cite{Zhabotinsky:2000fp,Holmes:2000uk,Kubota:2001ul,Dupont:2003vq,Bhalla:2004cu,Lucic:2008gt,Zeng:2010bq,Pepke:2010ju,Michalski:2012ds}. However, the structural complexity and multi-state nature of CaMKII present a technical challenge. A major problem is combinatorial explosion~\cite{Stefan:2014gl}. One CaMKII subunit has only a countable number of states to consider, but with 6 or 12 subunits on a holoenzyme, the number of combinations of states for a holoenzyme becomes extremely large. This is a common problem in systems biology since large proteins usually consist of multiple subunits. Software that adopts a rule-based approach such as BioNetGen~\cite{Hlavacek:2006iq} can be used to expand the network based on a set of given reaction rules. But for CaMKII holoenzyme, using BioNetGen to generate the network is still computationally intensive. After expansion with BioNetGen, a 6-subunit holoenzyme with four states for each subunit contains 700 unique species and 12192 unique reactions. The size of the network grows substantially when more detailed kinetics are involved. For example, each CaM has 4 Ca\textsuperscript{2+}-binding sites, giving rise to 9 distinct binding states; each state has its own distinct kinetics to interact with a CaMKII subunit. Then each CaMKII subunit exhibits a distinct phosphorylation rate depending on the state of CaM being bound. For simplicity, consider a 6-subunit CaMKII holoenzyme. Each subunit would potentially have 20 states. Then for the holoenzyme, the total number of unique species reaches over $\frac{6^{20}}{6}=609.36\times10^{12}$. It would be extremely time-consuming and not practical for BioNetGen to generate this network. To overcome this problem, most previous modeling studies of CaMKII simplified the Ca\textsuperscript{2+}-CaM-CaMKII reaction network by either modeling CaMKII as monomers~\cite{Bhalla:2004cu,Pepke:2010ju} or allowing only CaM fully-loaded with Ca\textsuperscript{2+} to interact with CaMKII subunits~\cite{Zhabotinsky:2000fp,Kubota:2001ul,Dupont:2003vq,Michalski:2012ds}.

Another complication is related to the intracellular environment. Conventionally, biochemical reactions are modeled in a deterministic manner using Ordinary Differential Equations (ODE). The underlying assumptions are that reactions occur in a spatially homogenous environment, reactants are abundant and not subject to stochasticity, and molecules diffuse sufficiently fast. The majority of previous modeling studies belong to this category except for a few that are stochastic or hybrid models ~\cite{Bhalla:2004cu, Zeng:2010bq, Holmes:2000uk}. However, intracellular space is highly heterogeneous and compartmentalized~\cite{LubyPhelps:2000uj,Dix:2008gy}. Reactions are often restricted to a small space. Structures such as the cytoskeleton, scaffold proteins or endoplasmic reticulum often act as diffusion barriers to slow down molecules. In addition, most previous modeling studies focus on the dendritic spine, where NMDA receptors-channels are the main Ca\textsuperscript{2+} providers.

In the present work, we focuse on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which servers for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus. Specifically, we examined the Ca\textsuperscript{2+}-CaM-CaMKII network activities near L-type Ca\textsuperscript{2+} channels. This is an area less studied previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}. To study this Ca\textsuperscript{2+}-CaM-CaMKII signaling network and deal with the problem of combinatorial explosion, we modified the published and freely available simulator Smoldyn~\cite{Andrews:2004fs}. Smoldyn is a particle-based stochastic simulator and has been used for simulating reaction and diffusion processes in cells. It works well for relatively simple reaction networks but not for holoenzymes such as CaMKII. We added new data structures to describe the reaction network in a compact way. CaMKII holoenzymes are modeled as a collection of subunits. Each subunit has a set of binding sites. Subunits react independently and diffuse collectively. Reactions are defined between binding sites. The reaction network is stored in a hashtable to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network. We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}. Moreover, we added detailed components to the network and examined the frequency dependence of CaMKII activation. We found that the number of subunits does not play a significant role in CaMKII activation. Instead, slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation. In the presence of a limited number of CaM molecules, CaM must make an intricate choice between binding Ca\textsuperscript{2+} first or a CaMKII subunit first. Interestingly, a change in the decision pattern leads to an altered frequency dependence of the network.

% % Results and Discussion can be combined.
\section*{Results}
\subsection*{Testing of the simulator modifications}
We first tested the modifications to Smoldyn used in this paper (see Methods) using a Ca\textsuperscript{2+}-CaM network (Fig~\ref{fig1}A). The 4 Ca\textsuperscript{2+} binding sites on CaM gives rise to a total of 9 different binding states of a CaM molecule. For the reaction volume we use a $500\times500\times500$ \SI{}{\cubic\nm} cube with all sides being reflective. The size of this volume mimics that of a Ca\textsuperscript{2+} channel nanodomain or a small dendritic spine head. Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.8671}{\micro\Molar} and \SI{9.3023}{\micro\Molar} respectively. The concentrations are chosen to produce an observable amount of 4-Ca\textsuperscript{2+} bound CaM at the steady state. All molecules are initially uniformly distributed. We tested reactions using two different Ca\textsuperscript{2+} diffusion constants \SI{0.0000022}{\square\cm\per\s}~\cite{Keller:2008ez} and half of this value \SI{0.0000011}{\square\cm\per\s}.

We first characterized the reactions in the network to see if results of our stochastic model could reasonably be compared to results with standard ODE methods. A second order chemical reaction in the solvent phase consists of two steps: molecules encountering each other by diffusion followed by the molecules reacting with each other. If the encountering step takes a much longer time to occur than the reacting step, the reaction is diffusion-limited; otherwise, the reaction is activation-limited. Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{},
\begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx \k_{on}$ as diffusion is sufficiently fast. Given the diffusion coefficients for Ca\textsuperscript{2+} and calmodulin and the parameter values for Ca\textsuperscript{2+} and calmodutlin reaction kinetics in Table 1., we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime. This was also true if diffusion constants were reduced 10 fold. ? Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.

We coded a deterministic ODE system in COPASI~\cite{} and compared results with our stochastic model using the same kinetics. We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{2}. Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used. These results suggest that our modifications of Smoldyn are working properly.

\begin{figure}[!h]
	\caption{{\bf}
	A: In this example, N1 and N2 are the two sites of N-lobe; C1 and C2 are the two sites of C-lobe. The binding and unbinding rate constants are labeled at the end of reactions. Between lobes reactions are independent. Within each lobe bindings are cooperative. Thus N2 binding cannot happen without N1 being bound; a similar rule applies to C sites. ‘+’ represents a binding reaction; ‘~’ represents molecules that are bound. ‘{}’ on the left-hand side specifies the binding sites states of reactants or the conditions for a reaction to occur. On the right-hand side, binding sites involved in the reaction are assigned to new values to change state. Notice the ‘==‘ sign on the left-hand side, and the ‘=‘ sign on the right-hand side are different. The ‘==‘ represents True or False of equality, whereas the ‘=‘ denotes an assigning operation. B: Ca\textsuperscript{2+} and CaM interactions, resulting in 9 different states of CaM. Arrows represent the direction of binding reactions. C: The complete Ca\textsuperscript{2+}-CaM-CaMKII network. The layer in blue represents Ca\textsuperscript{2+}-CaM interactions the same as in figure B. Edges are labeled to indicate binding reactions. For example, N1 means binding of a Ca\textsuperscript{2+} ion on CaM N1 site. KN1 means CaM attached to a CaMKII subunit binds a Ca\textsuperscript{2+} ion on N1 site. KNxCy means CaMKII subunit binds a CaM has $x$ Ca\textsuperscript{2+} ions bound on N-lobe and $y$ Ca2+ on C-lobe, $x$ and $y$ are from 0 to 2. KpNxCy means phosphorylated CaMKII subunit binds a CaM. Red edges mark the preferred pathway. D: Reaction scheme1. From top to bottom, the blue edges represent autophosphorylations of CaM N0C2, N1C2 and N2C2 respectively. The layer colored yellow represents Ca\textsuperscript{2+}-CaM interactions, the same as in panel B. E. Reaction scheme22. From left to right, the blue edges represent autophosphorylations of CaM N2C0, N2C1 and N2C2 respectively. The front layer(Layer2) colored yellow represents interactions between Ca\textsuperscript{2+} and CaMKII attached CaM molecules. Layer3 represents interactions between Ca\textsuperscript{2+} and phosphorylated CaMKII attached CaM molecules.}
\label{fig1}
\end{figure}

\begin{figure}[!h]
	\caption{{\bf}
	Time courses of 9 states of CaM molecules. The smooth lines are simulated using COPASI, and the discontinous lines are from the simple stochastic models with two different diffusion coefficients.}
\label{fig2}
\end{figure}

\subsection*{Analysis of the Ca\textsuperscript{2+}-CaM-CaMKII network}

We set up a prototype model to identify the major network branches that contribute to the phosphorylation of CaMKII subunits. The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig1}. In this network diagram molecule species are represented as vertices and reactions are represented as the network edges. CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in the $1 \times 1 \times 2$\SI{}{\cubic\um} box. As an initial contion, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII. The initial state is at an equilibrium, which we intentionally computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules (Fig~\ref{fig}). Ca\textsuperscript{2+} ions were released as a single source from the top $1 \times 1$ \SI{}{\square\um}. A previously generated input file having a total of 6 Ca2+ bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods). We ran the simulation up to \SI{2.5}{\s}, recorded the molecule numbers at every \SI{1}{\ms} and logged all reaction events. Using the events log, we counted the accumulated occurrences of each reaction type at every \SI{10}{\ms}, starting from the initial steady-state until the Ca\textsuperscript{2+} bursts were over.

For a particular reaction, the number of occurrences during a time span is considered to be the net number of molecule state changes along the corresponding edge, i.e., the number of unbinding events is subtracted from the number of binding events. A negative number means that unbinding occurs more often than binding within the given time span.

We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{}). The network consists of three layers (Fig~\ref{}). The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2) as shown separately in Fig~\ref{}. The front layer (Layer2) is for reactions of Ca\textsuperscript{2+} with CaM attached to unphosphorylated CaMKII (KNxCy). The middle layer (Layer3) is for reactions of Ca\textsuperscript{2+} with CaM attached to phosphorylated CaMKII (KpNxCy). Starting in Layer1, there are at most 4 possible ways for each CaM species to change state: binding a Ca\textsuperscript{2+} at a C site, binding Ca\textsuperscript{2+} at an N site, binding to an unphosphorylated CaMKII subunit or binding to a phosphorylated CaMKII subunit. However, binding to a phosphorylated CaMKII subunit requires a bound subunit to be phosphorylated first and then lose the bound CaM. This type of reaction rarely occurs during the early stage of the simulation since phosphorylation is slow and CaM unbinding from a phosphorylated subunit is uncommon. Thus we focus on the first 3 types of reactions.

\begin{figure}[!h]
	\caption{{\bf}
	Accumulated reaction occurrences over time for nine states of CaM molecules. For each type of CaM, three types of reactions can occur: binding of Ca\textsuperscript{2+} at N site, at C site or binding a CaMKII subunit. N1 represents Ca\textsuperscript{2+} binding on the N1 site; N2 means Ca\textsuperscript{2+} binding on the N2 site; C1 means Ca\textsuperscript{2+} binding on the C1 site; C2 means Ca\textsuperscript{2+} binding on the C2 site; K-NxCy means binding between CaMKII and CaM NxCy. Green arrows indicate the path consists of most likely chosen reactions starting from N0C0. Following the green arrows, labeled reaction leads to constructing the predominant pathway.}
\label{fig3}
\end{figure}
***
For each CaM state, we counted the accumulated occurrences of the possible reaction types. Results are shown in Fig~\ref{}. The plot reveals a predominant pathway for CaM and CaMKII state transitions. Starting as apoCaM, a CaM molecule tends to bind a Ca\textsuperscript{2+} ion on the C1 site and then on the C2 site, entering the N0C2 state. CaM in the N0C2 state has a strong preference to enter layer 2 by binding to a CaMKII subunit, after which it quickly binds Ca\textsuperscript{2+} ions to the N1 and N2 sites to become KN2C2 Fig~\ref{}. [Also evident from Fig~\ref{} is that once CaM has two Ca\textsuperscript{2+} ions on either lobe, the preference is to bind to a CaMKII subunit before binding additional Ca\textsuperscript{2+} ions. Once bound to a CaMKII subunit, additional Ca\textsuperscript{2+} ions bind to CaM more quickly. ?] In this scenario, it is rare for CaM to become fully bound with Ca\textsuperscript{2+} ions prior to binding with a CaMKII subunit. Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{}. This preferred pathway is consistent with a hypothetical major pathway that leads to CaMKII autophosphorylation shown in Pepke et al.~\cite{Pepke:2010ju}. The observation is also consistent with ~\cite{} in showing that partially loaded CaM molecules may be very important in activating CaMKII subunits. We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities. For example, even though N sites bind Ca\textsuperscrtip{2+} faster, the C sites have higher affinity making Ca\textsuperscript{2+} binding to C sites preferred.

To confirm the critical role of NxC2 (x=0,1,2) CaM in activation and phosphorylation of CaMKII, we set up two modified reaction schemes (Fig~\ref{}). In Scheme1, only CaMKII subunits bound with NxC2 are allowed to become phosphorylated. In Scheme2, phosphorylation is allowed only for subunits bound with N2Cx. In particular, the phosphorylation rates in the two schemes are equivalent, i.e., $k_{on}$ of reaction KpN2C0_rxn is the same as that of KpN0C2_rxn, and the $k_{on}$ of KN2C1_rxn equals that of KN1C2_rxn. Not surprisingly, the two schemes give rise to different phosphorylation levels as shown in Fig~\ref{}. Scheme 1 performs slightly worse than the whole network, whereas Scheme 2 produces much lower phosphorylation levels. Therefore, edges of the network are not equally involved in CaMKII phosphorylation. Phosphorylation from KNxC2 is especially important.

\begin{comment}
\subsection*{CaMKII activation frequency dependence is not primarily determined by CaMKII holoenzyme structures}

A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzymes activation are sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses. A recent study \textit{in vivo} also noticed glutate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}. It has been hypothesized that the holoenzyme structure requires cooperative phosphorylation between neighboring subunits and thus the structure contributes to the observed frequency dependence. However several other studies~\cite{Kubota:2001ul,Pepke:2010ju,Michalski:2012ds} suggest that the number of subunits of a holoenzyme does not result in significant difference in CaMKII activation pattern, therefore should not be related to the frequency dependence. Since multi-subunit CaMKII has not been modeled in fine details, we used the customized simulator to test the effects of subunits number by varying the number from 2, 6 to 12. For each subunit number, we used 5 Hz and 10 Hz Ca\textsuperscript{2+} input and simulated 10 trials for each frequency. The Ca\textsuperscript{2+} input files are generated from a 5 Hz and a 10 Hz theta-burst voltage file respectively, simulated using a stochastic Ca\textsuperscript{2+} channel model as described in Methods. 
%In particular, 40 trials of Ca\textsuperscript{2+} input files were generated for each voltage file and matched by total Ca\textsuperscript{2+} ions. The total Ca\textsuperscript{2+} amount difference between matched files is less than 10 Ca\textsuperscript{2+} ions.

For a given subunit number, we consistently observed significantly higher phosphorylation level with 10 Hz than 5 Hz (Fig~\ref{fig4}B,C). Typically, 10 Hz input allows more accumulation of NxC2 type CaM, leading to more bound CaMKII subunits and thus more phosphorylations (Fig~\ref{fig4}A). The number of subunits affects the phosphorylation level but does not significantly contribute to the observed frequency effect(Fig~\ref{fig4}C). Specifically, the 2-subunits condition makes it easier for a holoenzyme to propagate phosphorylation between neighbors, thus resulting in the highest phosphorylation level under 10 Hz and 5 Hz. However, the fewer-subunits-easier-propagation advantage is not consistent between the 6-subunits and 12-subunits conditions.% The 6-subunits condition results in a higher phosphorylation level than the 12-subunits condition under 10 Hz but not consistently so under 5 Hz.

\begin{figure}[!h]
	\caption{{\bf}
	A: Time courses of CaMKII subunits and CaM molecules when given 5 Hz Ca\textsuperscript{2+} influx generated from 16 channels. ‘camkii\_b’ means CaM bound CaMKII subunits. ‘camkii\_p’ means phosphorylated CaMKII subunits. ‘KN2C2 + KpNxC2’ represents CaM molecules bound to either CaMKII or phosphorylated CaMKII and are fully Ca\textsuperscript{2+} loaded on C sites. B: Time courses of phosphorylated CaMKII subunits for 10 Hz (smooth lines) and 5 Hz input (dashed lines) with the complete network, reaction scheme1 and reaction scheme2 respectively. C: Time courses of phosphorylated CaMKII subunits for 10 Hz(fast rising) and 5 Hz input (slow rising) when one holoenzyme contains 2, 6 and 12 subunits respectively. D: With 5 Hz input, time courses of molecule number for subunits that are activated themselves (bound or phosphorylated) and together their neighbors. E: The same as in D except that 10 Hz input is used.}
\label{fig4}
\end{figure}
\end{comment}

\subsection*{CaMKII activation frequency dependence and the detailed Ca2+-CaM-CaMKII network mechanisms invovled}

A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzymes activation are sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses. A recent study \textit{in vivo} also noticed glutamate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}. Numerous modeling studies have also reported a dependence of CaMKII activation on the frequency of calcium signals~\cite{}. To confirm this CaMKII activation pattern, we tested our 6-subunit CaMKII model using the previously generated 10 Hz and 5 Hz Ca\textsuperscript{2+} influx files (see Methods). The total number of influx Ca\textsuperscript{2+} ions are comparable (40433 ions in 5 Hz vs 40435 ions in 10 Hz). For each frequency, we simulated 10 trials and presented the time-varying mean +/- standard deviation of phosphorylated subunits. As shown in Fig~\ref{fig}, we observed significantly higher phosphorylation levels with 10 Hz than with 5 Hz input.

The observed frequency preference for 10 Hz is inherent to the Ca\textsuperscript{2+}-CaM-CaMKII binding. Through examining the reaction occurrences over time, we noticed that the 10 Hz input results in more bindings between NxC2 CaM and CaMKII, and consequently more autophosphorylations (Fig~\ref{}). Furthermore, the frequency effect can be modulated by slowing down Ca\textsuperscript{2+} diffusion and limiting the total amount of CaM. In the presence of moderate amount of Ca\textsuperscript{2+}, CaM molecules follow the predominant pathway by choosing reaction paths according to affinities. As a result, CaMKII interactions with CaM and Ca\textsuperscript{2+} binding to CaM are relatively independent. In contrast, when Ca\textsuperscript{2+} influx becomes saturating, the flush of Ca\textsuperscript{2+} quickly takes up the CaM, leaving CaMKII temporarily lack of binding partners. This temporary shortage of CaM results in an alternative pathway decision and ultimately a suboptimal phosphorylation level. Given the same total amount of Ca\textsuperscript{2+}, 10 Hz is easier to saturate CaM in a short time than 5 Hz. To demonstrate the effect of saturating Ca\textsuperscript{2+}, we increased the Ca\textsuperscript{2+} channel number from 16 channels as in the prototype model to allow more Ca\textsuperscript{2+} ions to flow in per action potential (Fig~\ref{fig5}). In the mean time, the total Ca\textsuperscript{2+} influx is kept consistent for 5 Hz and 10 Hz. As Ca\textsuperscript{2+} level increases, the network produces more phosphorylated CaMKII subunits, and the phosphorylation level difference between 5 Hz and 10 Hz conditions diminishes. Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz stimulus (Fig~\ref{fig5}F).

\begin{figure}[!h]
	\caption{{\bf}
	A: Using Ca\textsuperscript{2+} input generated with double times Ca2+\textsuperscript{2+} amount, accumulated reaction occurrences for autophosphorylation of CaMKII bound with N0C2, N1C2 and N2C2 respectively. Thick lines are for 10 Hz input; thin lines are for 5 Hz. B-E: The same as in A with Ca\textsuperscript{2+} input generated with $3\times$. $3.5\times$, $4\times$ and $6\times$ channels respectively. F: Summary plot of phosphorylation level with 10 Hz and 5 Hz input when various numbers of Ca\textsuperscript{2+} channels are involved in Ca\textsuperscript{2+} influx generation.}
\label{fig5}
\end{figure}

We examined the reaction occurrences profile over time on Layer1 edges (Fig~\ref{fig6}) and noticed that the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} level becomes saturating. This trend can be revealed from the pathway choice for CaM N0C2 (Fig~\ref{fig6}A-C). At nonsaturating Ca\textsuperscript{2+} level, N0C2 molecules are more likely to bind with CaMKII (reaction KN0C2) than binding a Ca\textsuperscript{2+} ion on the N1 site (reaction N1). When Ca\textsuperscript{2+} becomes saturating, N0C2 molecules tends to choose N1 binding over reaction KN0C2. This switch of reaction choice is observed for 10 Hz but not for 5 Hz. Likewise, we analyzed the profile of all possible CaMKII-CaM binding reactions. For 5 Hz conditions, binding of N0C2 consistently dominates over other CaMKII-CaM binding reactions, regardless of the amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig6}D-F). However for 10 Hz, as Ca\textsuperscript{2+} influx increases, binding of N2C2 (reaction KN2C2) gradually replaces binding of N0C2 (reaction KN0C2) and becomes the dominant choice for CaM to enter Layer2 (Fig~\ref{fig6}G-I). In breif, this switch of dominant pathway occurs for 10 Hz, correlating with the suboptimal phosphorylation level observed for 10 Hz.

\begin{figure}[!h]
	\caption{{\bf}
	A-C: Reaction occurrences of N0C0 binding Ca\textsuperscript{2+} at the N1 site and reaction occurrences of N0C0 directly binding with CaMKII. Thick lines are for 10 Hz input, and thin lines are for 5 Hz input. D-F: Reaction occurrences for CaMKII binding with all 9 CaM states. 5 Hz input generated with $1\times$, $3\times$ and $6\times$ Ca\textsuperscript{2+} influx are used. N0C2 are the dominant CaM state binding with CaMKII. Colored lines are for frequently occurred reactions. Grey lines are for infrequently occurred reactions. G-I: The same as in D-F except that 10 Hz input are used. 
	}
\label{fig6}
\end{figure}

The saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by slowing Ca\textsuperscript{2+} diffusion and limiting the available CaM in the system (Fig~\ref{fig7}). A slowed Ca\textsuperscript{2+} diffusion allows more interactions between Ca\textsuperscript{2+} and CaM, effectively mimicking a saturating level Ca\textsuperscript{2+}. To demonstrate, we set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network. Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} channels and input frequencies. Also, the frequency preference for 10 Hz is reversed at $3\times$ of $6\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig7}C,D). In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference. On the other hand, the same effect can be artificially created by limiting the total amount of CaM in the system (Fig~\ref{fig7}A,B). To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}). Even though the Ca\textsuperscript{2+} amount does not change, by limiting CaM we noticed a reverse of frequency preference from 10 Hz to 5 Hz. It is also worth noting is that by combining slow diffusion and limited CaM, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig7}E,F).

\begin{figure}[!h]
	\caption{{\bf}
	A: Using $3.5\times$ Ca\textsuperscript{2+} input, accumulated reaction occurrences for autophosphorylation of CaMKII bound with N0C2, N1C2 and N2C2 respectively. Thick lines are for 10 Hz input; thin lines are for 5 Hz. B: The same as in panel A except that the total CaM available in the system is decreased to half amount. C: Reaction occurrences using Ca\textsuperscript{2+} input generated with $3\times$ Ca\textsuperscript{2+} input. D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{0.0000011}{\square\cm\per\s}. E: Reaction occurrences using twice Ca\textsuperscript{2+} input. F: The same as in C except with slowed Ca\textsuperscript{2+} diffusion and half amount of CaM.}
\label{fig7}
\end{figure}

\subsection*{CaM changing dominant pathway can be demonstrated using a reduced network }

To better understand what happens during the process, we used a reduced reaction network to capture the observed frequency prefrence reversion (Fig~\ref{fig8}B). The network is derived from the initial steps in the whole Ca\textsuperscript{2+}-CaM-CaMKII network. It is simple enough to be simulated using the original Smoldyn. We used 5 pulses of instantaneous Ca\textsuperscript{2+} release as the input instead of realistic Ca\textsuperscript{2+} influx and varied the amount of Ca\textsuperscript{2+} influx per pulse. The CaMKII molecules are modeled as monomers with a phosphorylation rate of \SI{1}{\per\s}. The 10 Hz stimulus becomes saturating and fails to generates more phosphorylation than 5 Hz when Ca\textsuperscript{2+} amount reaches 40000 ions per pulse. We separated the reduced network into two paths based on the sequence of Ca\textsuperscript{2+}-CaM-CaMKII binding (Fig~\ref{fig8}A). Simulations were carried out using each path alone to compare with the reduced network. When Ca\textsuperscript{2+} amount is not saturating, the reduced network shows a comparable phosphorylation level with that of Path2 (Fig~\ref{fig7}B,C). This pathway choice is consistent with the "follow-the-affinity" principle. Also the phosphorylation level increases linearly with the Ca\textsuperscript{2+} input amount. However when Ca\textsuperscript{2+} becomes saturating, the growth of phosphorylation level with incrasing Ca\textsuperscript{2+} amount slows down for both 10 Hz and 5 Hz (Fig~\ref{fig8}A). But the trend is considerable for 10 Hz and only slight for 5 Hz. Such an effect is mostly contributed from Path2 (Fig~\ref{fig8}B). When Path2 alone is present, the phosphorylation level as function of increasing Ca\textsuperscript{2+} input shows a faster increase for 5 Hz than 10 Hz.

\begin{figure}[!h]
	\caption{{\bf}
	A: The reduced reaction network derived from the whole network. CaMKII are modeled as monomers and can undergo phosphorylation at an arbitrary rate \SI{1}{\per\s} as long as bound with a CaM N1C2. B: Summary plot of phosphorylation level by varying Ca\textsuperscript{2+} ions amount per pulse from 10000 ions/pulse to 50000 ions/pulse. Dashed lines represent the slow of phosphorylation level growth as a function of Ca\textsuperscript{2+} influx. C: Summary plot of phosphorylation level with the various amount of Ca\textsuperscript{2+} ions delivered at either 5 Hz or 10 Hz. Phosphorylation levels are generated through Path1 or Path2 alone. D: In the presence of non-saturating Ca\textsuperscript{2+}(ca10000), time courses of N0C2 molecules when Path1 or Path2 alone is present, or both are present. E: The same as in D except that in the presence of saturating Ca\textsuperscript{2+}(ca50000).}
\label{fig8}
\end{figure}

The two paths branch from the node N0C2. That is, CaM N0C2 molecules need to make choices for binding parterners. How choices are made depend on the amount of each binding partner. When Ca\textsuperscript{2+} amount is scarce, Path2 dominates; when Ca\textsuperscript{2+} becomes saturating, Path1 competes with Path2. Hence inherently, there is a competition invovled between the two paths, which can alter the time course of N0C2 and eventually the phosphorylation level. To visualize the competition, we compared the time courses of N0C2 in the reduced network and those when Path1 or Path2 are recruited alone. Hypotheticall, when competitions exhist, the N0C2 molecules are in high demand. Hence the consumption of N0C2 increases when both Path1 and Path2 are present. In other words, the amount of N0C2 decreases more than when either Path is present alone. This is indeed the case when we examined the N0C2 amount in nonsaturating (10000 ions per pulse, Fig~\ref{fig8}D) and saturating Ca\textsuperscript{2+} influx (50000 ions per pulse, Fig~\ref{fig8}E). In the nonsaturating case, the time course of N0C2 of the reduced network matches well with that of the Path2 alone condition. N0C2 molecules in Path2 are not affected by the presence of Path 1. In contrast, in the saturating case, N0C2 amount becomes considerably lower when both paths are recruited together than when either path is present alone. This implies that the two paths negatively affect each other. Hence confirms the competition.

\section*{Discussion}
In the study, we presented an efficient approach to simulate multi-subunit molecules and reaction network in detailed kinetics, as well as the effects of diffusion on the reaction network. We also introduced an intuitive approach to analyzing the pathway choices of a network based on the reaction history of a simulation, allowing us to grasp insights of a large network quickly.

Our findings agree with observations from previous studies. First, under physiological conditions when Ca\textsuperscript{2+} influx is moderate, CaM molecules partially loaded with Ca\textsuperscript{2+} are important for CaMKII activation. In particular, the pathway through N0C1, N0C2, and KN0C2 plays a crucial role. Second, when studying the effects of Ca\textsuperscript{2+} input frequency, we noticed that the number of subunits does not significantly affect the patterns of CaMKII. The mechanism should reside within the network kinetics. These notions are consistent with previous modeling studies ~\cite{Kubota:2001ul,Pepke:2010ju,Michalski:2012ds}. Third, CaM availablility is factor for CaMKII frequency response, as suggested in ~\cite{DeKoninck:1998wh}. Given that free diffusible CaM molecules are limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}, this conclusion further implies the regulatory role of of many endogenous CaM binding proteins ~\cite{Skene:1990kf}. Fourth, factors such as crowding and spatial homogeneity can slow down molecule diffusion and thus may have substantial effects on CaMKII phosphorylation. In our simulations, Ca\textsuperscript{2+} diffusion slowed down by half could lead to a dramatic increase of phosphorylation level. A similar effect of slow diffusion on CaMKII activation has also been shown in a spine model~\cite{Bhalla:2004cu}. In addition, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.

%chao etal, autoinhibited through dimerization;
%stratton etal, subunits exchange 
Our model does not contain Thr305/Thr306 phosphorylation and some newly discovered mechanisms of CaMKII, which may explain the special multi-subunit strucutre of the holoenzyme. For example, it has been found that there exists a compact autoinhibition state, which occurs through dimerization of adjacent subunits from top and bottom rings. Once Ca\textsuperscript{2+} bound to a dimerized subunit, the dimer dissembles and the two subunits are released. Another recent study indicated that phosphorylated CaMKII subunits can undergo subunits exchange to faciliate proprogating the activation triggered by Ca\textsuperscript{2+}-CaM~\cite{Stratton:2014ct}.

%Our finding indicates the CaM amount is important for CaMKII frequency dependence. It is known that free diffusible apoCaM molecules are limited in cells~\cite{}. Although there are abundant intracellular CaM molecules, most of them are captured by CaM binding proteins. The distribution of CaM is spatially heterogeneous as well. It is estimated that CaM molecules are enriched upto mM level concentration in areas close to the Ca\textsuperscript{2+} channels~\cite{Mori:2004fb}. These extra details making CaM regulation a complicated process and the output of Ca\textsuperscript{2+} signaling network necessarily depends on the specific biophysical context in which the network is located.

%Our approach to model multi-subunit macromolecules allows binding sites to act independently, therefore potentially different from simply expanding reaction rules and loading the reactions for simulation. A molecule can undergo multiple state changes within a time step as long as the binding sites involved are independent of each other. For example, apoCaM N0C0 can become N1C1 within a time step, since N1 and C1 act independently; but cannot become N0C2 or N2C0, because N2 and C2 act cooperatively on N1 and C1 respectively. But since our time step is sufficiently small, the discrepancy between the implementations stay conceptual and does not lead to different steady states.

One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of high molecule concentrations. One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM are acting very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in modeling studies. These binding kinetics are in the diffusion-limited regime, rendering traditional mass action based methods inaccurate. However for these fast kinetics, there currently lacks an efficient method to simulate. If to simulate with the original Smoldyn algorithm, the simulation time step has to be considerablly reduced to obtain the correct steady state~\cite{Andrews:2015}. Another software package using an enhanced Green's Function Algorithm~\cite{vanZon:2005jd} can handle the high concentration diffusion-limited reactions accurately, but it takes an impractically long time to simulate. It is interesting to examine the Ca\textsuperscript{2+}-CaM-CaMKII using the fast kinetics parameters and we anticipate to address this question in future development.

\section*{Methods}
\subsection*{Simulator modifications}

We expanded the molecule data structure in Smoldyn to include complexes, molecules and binding sites. A complex may contain multiple molecules and a molecule may contain multiple binding sites. Reactions are specified between binding sites. Each binding site has binary states. For example, bound is coded as 1 and unbound as 0; phosphorylated as 1 and unphosphorylated as 0. Each molecule has a vector to store the states of binding sites. All reactions are stored in a hashtable with reactants and binding states as entry keys. As an example, binding reactions involving the N and C lobes of a CaM molecule can be coded as in Fig~\ref{fig}. CaMKII holoenzymes are two 6-subunits rings. But for simplicity, we modeled each CaMKII holoenzyme as only one ring at a radius of \SI{8}{\nm}. Each subunit is separated from its direct neighbors at a fixed distance \SI{8}{\nm} (estimated from ~\cite{Gaertner:2004jk}) and has a distinct physical location.

In the original version of Smoldyn, each reaction generates new molecules as products and removes reactants. In our case, since one molecule can have multiple binding sites and is potentially associated with multiple partners, entirely removing a molecule is not practical because other attached molecules would also be affected. Also, removing and generating new molecules makes it difficult to track the reaction history of a molecule. Therefore, during reactions we do not remove molecules, but merely change molecule identities and positions. Molecules bound together physically overlap and synchronize their locations automatically. When they diffuse, the diffusion coefficient is determined by the dominant molecule. For example, when Ca\textsuperscript{2+} and CaM are bound, Ca\textsuperscript{2+} molecules diffuse with CaM.

Macromolecules usually have multiple binding sites, and sometimes these sites compete for the same ligand. For example, CaM has 4 Ca\textsuperscript{2+} binding sites. Since the N and C sites act independently, the N1 and C1 sites compete for binding Ca\textsuperscript{2+} and an apoCaM (N0C0) can become either N1C0 or N0C1, resulting in a branching reaction scheme. Thus a decision process is needed to choose a reaction path when a binding event occurs.

As an example, consider the following two reactions
\begin{quote}
\begin{verbatim}
	rxn1: camN0C0 + ca <-> camN1C0
	rxn2: camN0C0 + ca <-> camN0C1
\end{verbatim}
\end{quote}

Rxn1 has a forward rate constant $k_{f1}$ in \SI{}{\per\micro\Molar\per\s} and a backward rate constant $k_{b1}$ \SI{}{\per\s}. Rxn2 has rate constants $k_{f2}$ and $k_{b2}$. The two reactions can be viewed together as rxn3, which has an overall kinetics $k_{f3}$ and $k_{b3}$. According to the law of mass action, the two reactions can be written in to differential equations and the following equations hold:
\begin{equation}
	k_{f3}=k_{f1}+k_{f2}
\end{equation}
\begin{equation}\frac{k_{f3}}{k_{b3}}=\frac{k_{f2}}{k_{b2}} + \frac{k_{f1}}{k_{b1}}\end{equation}
\begin{equation}k_{b3}=pk_{b1} + (1-p)k_{b2}\end{equation}

where $p$ represents the proportion in $k_{b3}$ is due to unbinding of camN1C0. Knowing that $\frac{p}{1-p}=\frac{[camN1C0]}{[camN0C1]}=\frac{k_{f1}/k_{b1}}{k_{f2}/k_{b2}}$, we obtain $k_{f3}=k_{f1}+k_{f2}$. Smoldyn uses binding radii to implement second order reactions. If two molecules are spatially separated by a distance smaller than the corresponding binding radius, then the reaction proceeds. In Smoldyn, a special algorithm is used to calculate the binding radius, which depends on the kinetic rate constant, simulation time step and total diffusion rate of reactants. In case of a branched binding scheme sharing common reactants, we first calculate a binding radius $r_3$ based on $k_{f3}$. If the distance between a molecule pair is smaller than $r_3$, binding happens. To make a reaction choice, we generated a uniformly distributed random number from 0 to 1. If the number falls in the range $(0,\frac{kf_1}{kf_3}]$, then rxn1 is chosen; instead if the number falls in the range $(\frac{kf_1}{kf_3}, 1]$, we pick rxn2. Following this approach, the steady state of the reaction network can be kept consistent with the prediction by the mass action law.

\subsection*{Reaction network}
We focus on the interactions among Ca\textsuperscript{2+}, CaM, and CaMKII. We first used a Ca\textsuperscript{2+}-CaM network for testing to confirm that modifications on the simulator are valid. Then we added CaMKII holoenzymes to study the signaling network in detailes (Fig~\ref{fig1}C). Our rule-based approach allows reactions to be described at the binding site level and perform simulations without compromising the resolution of chemical kinetics. We set up a cube-shaped model to represent a chunk from a cell body. The cell is assumed as a sphere with radius \SI{5}{\micro\meter}. The cube has a dimension \SI{1}{\um} in width and \SI{2}{\um} in depth (Fig~\ref{fig9}A). The top surface of the cube is the cell membrane, reflective to all molecules. The four sides are also reflective to prevent molecul1es from escaping laterally through membrane surface of side compartments. The reflective boundary is essentially equivalent to periodic boundary in terms of conservation of molecules in lateral direction. The bottom surface is partially absorbing to Ca\textsuperscript{2+} ions but reflective to CaM. Reflective to CaM guarantees a steady state initial condition. The partial absorption is built-in in Smoldyn to resemble an unbounded diffusion situation~\cite{Andrews:2009dr}.

\begin{figure}[!h]
	\caption{{\bf}
	A: Dimension of the cube-shaped model. Ca\textsuperscript{2+} channels are located on the top surface. B: Find the number channels by fitting the single channel current to the GHK equation. N equals 6 by fitting. C: $n_{inf}(v)$ and $\tau_n(v)$ used in the voltage-gated Ca\textsuperscript{2+} channels.}
\label{fig9}
\end{figure}

Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type CaL) are located on the top surface to provide Ca\textsuperscript{2+} influx. For simplicity, these channels are clustered at the center of the membrane. The channels open and close depending on a time-varying membrane voltage file generated from a neuron model. CaMKII subunits ($\beta$-subunits) are uniformly present at a concentration of \SI{10}{\umol}. They are also immobilized presumably by attaching to actins~\cite{Li:2016cq}. Free diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}. This is consistent with the notion that at the cell resting level free diffusible CaM is considerably limited~\cite{Tran:2003fs,2008BpJ....95.6002S,LubyPhelps:1995kl}. Table S1 lists all the reactions with corresponding kinetic parameters involved in the network. Kinetic parameters are integrated from various sources and are adjusted to satisfy microscopic reversibility. 
%However, they are highly concentrated (\SI{245}{\umol}) in the nanodomain attached to Ng molecules. The high concentration of CaM is consistent with the finding~\cite{Mori:2004fb} that L-type Ca\textsuperscript{2+} entry is enriched with CaM. Besides,

\subsection*{Ca\textsuperscript{2+} input conditions}
We used a previously built multi-compartment neuron model to generate the somatic action potential trains at frequencies of 5 Hz and 10 Hz~\cite{Li:2014fv}. The action potentials are activated by synaptic stimulations, following a series of presynaptic release of neurotransmitters which was modeled with a probability profile measured in experiments~\cite{Grover:2009hb}. Presynaptic stimulations activate spine AMPA receptors, elevate the membrane potential and open NMDA receptors, allowing Ca\textsuperscript{2+} ions to enter the spines. Accumulated Ca\textsuperscript{2+} influx further depolarizes the dendrite tree and rapidly propagate toward the soma, initiating somatic action potentials. The pattern of action potentials follows the synaptic stimulation. We varied the intervals of synaptic release to generate action potentials in the form of theta-burst with different inter-bursts intervals, which provide different Ca\textsuperscript{2+} influx frequencies for the signaling network.

The cell level neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances. Its membrane voltage output was applied to the Ca\textsuperscript{2+} signaling model (Fig~\ref{fig10}A,C). Since L-type Ca\textsuperscript{2+} channels are relatively fast, the membrane potentials are unaffected by the channel activities, and the neuron model and the signaling model can be safely decoupled. The L-type channels in the signaling model are implemented as single channels with voltage-depdendent stochastic opening and closing. The rates of open and close are functions of membrane voltage and calculated using varialbles of $n$ and $\tau$ (Fig~\ref{fig9}C) similar to Hudgkin-Huxley equations~\cite{Tuckwell:2012tt}.

\begin{quote}
 \verb|CaL{gate==0} <-> CaL[gate=1]|.
\end{quote}
Then the following set of equations are used to decribe voltage-gated Ca\textsuperscript{2+} channels:
\begin{equation}rate_{open}(V)=\frac{n_{\inf}}{\tau_n}\end{equation}
\begin{equation}rate_{close}(V)=\frac{1-n_{\inf}}{\tau_n}\end{equation}
\begin{equation}n_{\inf}=\frac{1}{1+exp(\frac{V+V_{half}}{slope})}\end{equation}
\begin{equation}\tau_n=\tau_0+\frac{4\tau_a\sqrt{\tau_g(1-\tau_g)}}{exp(-D_V(1-\tau_g)k_f/s)+exp(D_V\tau_gk_f/s)}\end{equation}

The $rate_{open}$ and $rate_{close}$ are used to calculate conditional probabilities to determine the state of a channel for the next time step in the following way
\begin{equation}P(C|O)=1-exp(-rate_{close}(V)dt)\end{equation}
\begin{equation}P(O|C)=1-exp(-rate_{open}(V)dt)\end{equation}
\begin{equation}P(O|O)=1-P(C|O)\end{equation}

For each channel, at a given time, a probability is calculated based on the membrane voltage to decided whether a channel opens. If it opens, a varying number of Ca\textsuperscript{2+} ions are generated. To calculate how many ions, we used the following equation~\cite{2013PNAS..11015794T} to obtain the unitary current for a single channel
\begin{equation}i_{ca}=-g(V-V_s)\frac{\exp\frac{-(V-V_s)}{RT/zF}}{1-\exp\frac{-(V-V_s)}{RT/zF}}\end{equation}
where $g$ is chosen as \SI{5}{\pico\siemens} and $RT/zF$ equals \SI{12}{\milli\volt}. A current density is calculated using GHK current equation~\cite{citeulike:306134} assuming an extracellular [Ca\textsuperscript{2+}] of \SI{2}{\m\Molar} and an intracellular concentration of \SI{50}{\nano\Molar}. Since the membrane surface is \SI{1}{\um\square}, the current density is converted to a total current $I_{ghk}$ for the area. Assuming a total number of N channels are present, for each channel $\frac{I_{ghk}}{N}$ should equal $i_{ca}$. By fitting the two expressions, we obtained a N of 16 and a $V_s$ of \SI{-1.90955}{\milli\volt} (Fig~\ref{fig9}B). Using the $i_{ca}$, the number of ions entered during one time step is calcuated as $\frac{i_{ca}}{2e}dt$, where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge. To speed up the simulation, Ca\textsuperscript{2+} influx patterns were generated in a separate simulation using the voltage file (Fig~\ref{fig10}B,D). Then the generated Ca\textsuperscript{2+} influx files were fed into the signaling network. To guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then matched them by total Ca\textsuperscript{2+}.

\begin{figure}[!h]
	\caption{{\bf}
	A: A 5 Hz action potential trains voltage file generated from the NEURON model. B: Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file. Ca\textsuperscript{2+} influx is stochastic. There are spontaneous firings in the absence of action potentials. C: 10 Hz action potential trains voltage file generated from the NEURON model. D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} influx files.}
\label{fig10}
\end{figure}
\section*{Conclusion}
\section*{Supporting information}

% For figure citations, please use "Fig" instead of "Figure".
% Place figure captions after the first paragraph in which they are cited.

% % Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
% \paragraph*{S1 Fig.}
% \label{S1_Fig}
% {\bf Bold the title sentence.} Add descriptive text after the title of the item (optional).

% \paragraph*{S1 File.}
% \label{S1_File}
% {\bf Lorem ipsum.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

% \paragraph*{S1 Appendix.}
% \label{S1_Appendix}
% {\bf Lorem ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\paragraph*{S1 Table.}
\label{S1_Table}
{\bf The total amount of Ca\textsuperscript{2+} influx of selected files for 5 Hz and 10 Hz.} 
\paragraph*{S2 Table.}
\label{S2_Table}
{\bf Kinetic parameters of all reactions.}
\paragraph*{S3 Table.}
\label{S3_Table}
{\bf Reaction characterizations.}

\section*{Acknowledgments}

\nolinenumbers

% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% or
%
% Compile your BiBTeX database using our plos2015.bst
% style file and paste the contents of your .bbl file
% here. See http://journals.plos.org/plosone/s/latex for 
% step-by-step instructions.
% 
\bibliography{paper}{}

\end{document}

### 2017-06-12 paper.tex
reversed frequency preference revised

@@
 A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzymes activation are sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses.
-A recent study \textit{in vivo} also noticed glutamate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}.
+A recent study also showed that \textit{in vivo} glutamate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}.
 Numerous modeling studies have also reported a dependence of CaMKII activation on the frequency of calcium signals~\cite{}.
@@
 
-The observed frequency preference for 10 Hz is inherent to the Ca\textsuperscript{2+}-CaM-CaMKII binding.
-Through examining the reaction occurrences over time, we noticed that the 10 Hz input results in more bindings between NxC2 CaM and CaMKII, and consequently more autophosphorylations (Fig~\ref{}).
-Furthermore, the frequency effect can be modulated by slowing down Ca\textsuperscript{2+} diffusion and limiting the total amount of CaM.
+We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between NxC2 CaM and CaMKII, and consequently more autophosphorylations (Fig~\ref{}).
+Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
+The frequency effect can be reversed by providing saturating amount of Ca\textsuperscript{2+}.
+To demonstrate, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential (Fig~\ref{fig5}).
+In the mean time, the total Ca\textsuperscript{2+} influx is kept consistent for 5 Hz and 10 Hz.
+As Ca\textsuperscript{2+} level increases, the network produces more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminishes.
+Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz input (Fig~\ref{fig5}F).
+
+Equivalently, the saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by limiting the available CaM and slowing Ca\textsuperscript{2+} diffusion (Fig~\ref{fig7}).
+To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
+Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed a reverse of frequency preference from 10 Hz to 5 Hz.
+The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
+We set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
+Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequencies.
+Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig7}C,D).
+In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference.
+It is also worth noting is that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig7}E,F).
+
+To explain what happened, we profiled the reaction occurrences over time on Layer1 edges.
+We noticed that the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} level becomes saturating.
 In the presence of moderate amount of Ca\textsuperscript{2+}, CaM molecules follow the predominant pathway by choosing reaction paths according to affinities.
-As a result, CaMKII interactions with CaM and Ca\textsuperscript{2+} binding to CaM are relatively independent.
-In contrast, when Ca\textsuperscript{2+} influx becomes saturating, the flush of Ca\textsuperscript{2+} quickly takes up the CaM, leaving CaMKII temporarily lack of binding partners.
-This temporary shortage of CaM results in an alternative pathway decision and ultimately a suboptimal phosphorylation level.
+On one hand, bindings between N0C2 and CaMKII (rxn_KN0C2) are more likely to occur than bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules (rxn_N1).
+Similarly, CaMKII binding with N0C2 (rxn_KN0C2) dominates all types of CaMKII-CaM binding reactions (Fig~\ref{fig}).
+However, when \textsuperscript{2+} influx becomes saturating, CaM molecules tend to stay in the Layer1 till they get fully loaded by Ca\textsuperscript{2+}.
+As a result, CaMKII binding with N2C2 (rxn_KN2C2) becomes dominant.
+This decision change gives rise to an altered pathway and eventually a suboptimal phosphorylation level.
 Given the same total amount of Ca\textsuperscript{2+}, 10 Hz is easier to saturate CaM in a short time than 5 Hz.
-To demonstrate the effect of saturating Ca\textsuperscript{2+}, we increased the Ca\textsuperscript{2+} channel number from 16 channels as in the prototype model to allow more Ca\textsuperscript{2+} ions to flow in per action potential (Fig~\ref{fig5}).
-In the mean time, the total Ca\textsuperscript{2+} influx is kept consistent for 5 Hz and 10 Hz.
-As Ca\textsuperscript{2+} level increases, the network produces more phosphorylated CaMKII subunits, and the phosphorylation level difference between 5 Hz and 10 Hz conditions diminishes.
-Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz stimulus (Fig~\ref{fig5}F).
+Correspondingly, the switch of pathway choice is more dramatic for 10 Hz at $4.5\times$ Ca\textsuperscript{2+}, resulting in the revserved frquency perference as observed previously.
 
@@
 
+\begin{comment}
 We examined the reaction occurrences profile over time on Layer1 edges (Fig~\ref{fig6}) and noticed that the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} level becomes saturating.
@@
 In breif, this switch of dominant pathway occurs for 10 Hz, correlating with the suboptimal phosphorylation level observed for 10 Hz.
+\end{comment}
 
@@
 \end{figure}
-
-The saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by slowing Ca\textsuperscript{2+} diffusion and limiting the available CaM in the system (Fig~\ref{fig7}).
-A slowed Ca\textsuperscript{2+} diffusion allows more interactions between Ca\textsuperscript{2+} and CaM, effectively mimicking a saturating level Ca\textsuperscript{2+}.
-To demonstrate, we set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
-Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} channels and input frequencies.
-Also, the frequency preference for 10 Hz is reversed at $3\times$ of $6\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig7}C,D).
-In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference.
-On the other hand, the same effect can be artificially created by limiting the total amount of CaM in the system (Fig~\ref{fig7}A,B).
-To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
-Even though the Ca\textsuperscript{2+} amount does not change, by limiting CaM we noticed a reverse of frequency preference from 10 Hz to 5 Hz.
-It is also worth noting is that by combining slow diffusion and limited CaM, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig7}E,F).
 

### 2017-06-25 paper.tex
methods

@@
 \DeclareSIUnit\Molar{\textsc{m}}
-\sisetup{scientific-notation = true}
+%\sisetup{scientific-notation = true}
 %% Include all macros below
@@
 Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons and understanding how this occurs has been the focus of many modeling studies.
-However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature and complex activation kinetics.
-To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion of the number of species.
+However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled.
+To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion.
 With this new method, spatial and temporal aspects of the signaling network can be studied without compromising biochemical details.
 We used this new method to examine how calmodulin molecules, both partially loaded and fully loaded with Ca\textsuperscript{2+}, choose pathways to interact with and activate CaMKII under various Ca\textsuperscript{2+} input conditions.
-We found that the frequency dependence of the CaMKII phosphorylation with Ca\textsuperscript{2+} input is not primarily determined by subunit number, but is intrinsic to the network and can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion.
-Depending on whether Ca\textsuperscript{2+} input is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns.
-The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca2+ channel nanodomains and cytoplasm.
+We found that the frequency dependence of the CaMKII phosphorylation is intrinsic to the network kinetics and the activation pattern can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion.
+Depending on whether Ca\textsuperscript{2+} influx is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns.
+In addition, the size of the holoenzyme shows a subtle effect on CaMKII activation.
+The more extended the subunits are organized, the easier for CaM molecules to access and activate the subunits.
+The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca\textsuperscript{2+} channel nanodomains and cytoplasm.
 % Please keep the Author Summary between 150 and 200 words
@@
 These strikingly contrasting cell processes are possible because Ca\textsuperscript{2+} signals of varying strengths and temporal patterns are interpreted differently by intracellular signaling pathways.
-In our study, we focus on one particular pathway that involves calmodulin and Calcium/calmodulin-dependent protein kinase II (CaMKII).
-We have developed a new computational method that gets around the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods.
-Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca2+ input conditions, calmodulin availability and the Ca2+ diffusion rate.
+In our study, we focus on one particular pathway that involves calmodulin and Ca\textsuperscript{2+}/calmodulin-dependent protein kinase II (CaMKII).
+We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods.
+Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca\textsuperscript{2+} input conditions, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.
 
@@
 CaMKII is involved in intracellular Ca\textsuperscript{2+} signaling as an effector of the Ca\textsuperscript{2+} sensor protein calmodulin (CaM)~\cite{Herring:2016bh,Hell:2014bd,Coultrap:2012ip}.
-In neurons, CaMKII molecules are especially involved in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}.
+In neurons, CaMKII molecules are essential in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}.
 In spines, activated CaMKII molecules interact with postsynaptic density proteins, facilitating actins to reorganize, leading to spine enlargement and upregulation of $alpha$-amino-3-hydroxy-5-methyl-4-isoxazolepropionic acid (AMPA) receptor numbers~\cite{Herring:2016bh}.
@@
 
-The CaMKII molecule is a holoenzyme that consists of 12 subunits.
+The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings.
 Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase.
@@
 Finally, Thr305/Thr306 can also be phosphorylated.
-This theronine site overlaps with the CaM binding site and its phosphorylation blocks the binding of CaM molecules.
+There theronine sites overlap with the CaM binding site and its phosphorylation blocks the binding of CaM molecules.
 In this case, the subunit becomes "capped".
@@
 Another complication is related to the intracellular environment.
-Conventionally, biochemical reactions are modeled in a deterministic manner using Ordinary Differential Equations (ODE).
+Conventionally, biochemical reactions are modeled in a deterministic manner using Ordinary Differential Equations (ODEs).
 The underlying assumptions are that reactions occur in a spatially homogenous environment, reactants are abundant and not subject to stochasticity, and molecules diffuse sufficiently fast.
@@
 Structures such as the cytoskeleton, scaffold proteins or endoplasmic reticulum often act as diffusion barriers to slow down molecules.
-In addition, most previous modeling studies focus on the dendritic spine, where NMDA receptors-channels are the main Ca\textsuperscript{2+} providers.
+In addition, most previous modeling studies focus on the dendritic spine, where NMDA receptor-channels are the main Ca\textsuperscript{2+} providers.
 
-In the present work, we focuse on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which servers for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus.
-Specifically, we examined the Ca\textsuperscript{2+}-CaM-CaMKII network activities near L-type Ca\textsuperscript{2+} channels.
+In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus.
+Specifically, we examined Ca\textsuperscript{2+}-CaM-CaMKII network activity near L-type Ca\textsuperscript{2+} channels.
 This is an area less studied previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}.
@@
 Moreover, we added detailed components to the network and examined the frequency dependence of CaMKII activation.
-We found that the number of subunits does not play a significant role in CaMKII activation.
-Instead, slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation.
-In the presence of a limited number of CaM molecules, CaM must make an intricate choice between binding Ca\textsuperscript{2+} first or a CaMKII subunit first.
+We found that slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation.
+Also in the presence of a limited number of CaM molecules, CaM must make an intricate choice between binding Ca\textsuperscript{2+} first or a CaMKII subunit first.
 Interestingly, a change in the decision pattern leads to an altered frequency dependence of the network.
@@
 The 4 Ca\textsuperscript{2+} binding sites on CaM gives rise to a total of 9 different binding states of a CaM molecule.
-For the reaction volume we use a $500\times500\times500$ \SI{}{\cubic\nm} cube with all sides being reflective.
+For the reaction volume we use a $\SI{500}{\nm}\times\SI{500}{\nm}\times\SI{500}{\nm}$ cube with all sides being reflective.
 The size of this volume mimics that of a Ca\textsuperscript{2+} channel nanodomain or a small dendritic spine head.
-Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.8671}{\micro\Molar} and \SI{9.3023}{\micro\Molar} respectively.
+Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.867}{\micro\Molar} and 9.302\SI{}{\micro\Molar} respectively.
 The concentrations are chosen to produce an observable amount of 4-Ca\textsuperscript{2+} bound CaM at the steady state.
 All molecules are initially uniformly distributed.
-We tested reactions using two different Ca\textsuperscript{2+} diffusion constants \SI{0.0000022}{\square\cm\per\s}~\cite{Keller:2008ez} and half of this value \SI{0.0000011}{\square\cm\per\s}.
+We tested reactions using two different Ca\textsuperscript{2+} diffusion constants \SI{2.2e-6}{\square\cm\per\s}~\cite{Keller:2008ez} and half of this value \SI{1.1e-6}{\square\cm\per\s}.
 
@@
 If the encountering step takes a much longer time to occur than the reacting step, the reaction is diffusion-limited; otherwise, the reaction is activation-limited.
-Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{},
+Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{Bamford198505},
 \begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
-For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx \k_{on}$ as diffusion is sufficiently fast.
-Given the diffusion coefficients for Ca\textsuperscript{2+} and calmodulin and the parameter values for Ca\textsuperscript{2+} and calmodutlin reaction kinetics in Table 1., we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime.
+For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx k_{on}$ as diffusion is sufficiently fast.
+Given the diffusion coefficients for Ca\textsuperscript{2+} and CaM and the parameter values for Ca\textsuperscript{2+} and CaM reaction kinetics in Table S2, we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime.
 This was also true if diffusion constants were reduced 10 fold.
-?
 Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.
 
-We coded a deterministic ODE system in COPASI~\cite{} and compared results with our stochastic model using the same kinetics.
-We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{2}.
+We coded a deterministic ODE system in COPASI~\cite{Hoops:2006gy} and compared results with our stochastic model using the same kinetics.
+We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{fig2}.
 Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used.
@@
 \begin{figure}[!h]
-	\caption{{\bf}
-	A: In this example, N1 and N2 are the two sites of N-lobe; C1 and C2 are the two sites of C-lobe.
+	\caption{{\bf Ca\textsuperscript{2+}-CaM interaction network.}
+	A: Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM.
+CaM has two lobes, an N lobe and a C lobe, each of which has 2 Ca\textsuperscript{2+} binding sites.
+Binding sites on the lobes are designated N1, N2, C1 and C2.
+The N lobe and C lobe bind Ca\textsuperscript{2+} independently of each other, but Ca\textsuperscript{2+} binding within a lobe is cooperative.
+Thus site N2 cannot bind Ca\textsuperscript{2+} unless N1 has already bound Ca\textsuperscript{2+} and a similar rule applies to sites C2 and C1.
+In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" implies binding sites N1 and N2 are both bound with Ca\textsuperscript{2+} and "C2" implies similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" imply only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound.
+Arrows represent the direction of binding reactions.
+	B: The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator.
 The binding and unbinding rate constants are labeled at the end of reactions.
-Between lobes reactions are independent.
-Within each lobe bindings are cooperative.
-Thus N2 binding cannot happen without N1 being bound; a similar rule applies to C sites.
-‘+’ represents a binding reaction; ‘~’ represents molecules that are bound.
-‘{}’ on the left-hand side specifies the binding sites states of reactants or the conditions for a reaction to occur.
-On the right-hand side, binding sites involved in the reaction are assigned to new values to change state.
-Notice the ‘==‘ sign on the left-hand side, and the ‘=‘ sign on the right-hand side are different.
+‘+’ represents a binding reaction; ‘$\sim$’ represents molecules that are bound.
+‘\{\}’ on the left-hand side specifies the states of binding sites of reactants or the conditions for a reaction to occur.
+On the right-hand side, binding sites involved in the reaction are assigned to new values.
+Notice the ‘==‘ sign is used on the left-hand side, but the ‘=‘ sign is used on the right-hand side.
 The ‘==‘ represents True or False of equality, whereas the ‘=‘ denotes an assigning operation.
-B: Ca\textsuperscript{2+} and CaM interactions, resulting in 9 different states of CaM.
-Arrows represent the direction of binding reactions.
-C: The complete Ca\textsuperscript{2+}-CaM-CaMKII network.
-The layer in blue represents Ca\textsuperscript{2+}-CaM interactions the same as in figure B.
-Edges are labeled to indicate binding reactions.
-For example, N1 means binding of a Ca\textsuperscript{2+} ion on CaM N1 site.
-KN1 means CaM attached to a CaMKII subunit binds a Ca\textsuperscript{2+} ion on N1 site.
-KNxCy means CaMKII subunit binds a CaM has $x$ Ca\textsuperscript{2+} ions bound on N-lobe and $y$ Ca2+ on C-lobe, $x$ and $y$ are from 0 to 2.
-KpNxCy means phosphorylated CaMKII subunit binds a CaM.
-Red edges mark the preferred pathway.
-D: Reaction scheme1.
-From top to bottom, the blue edges represent autophosphorylations of CaM N0C2, N1C2 and N2C2 respectively.
-The layer colored yellow represents Ca\textsuperscript{2+}-CaM interactions, the same as in panel B.
-E.
-Reaction scheme22.
-From left to right, the blue edges represent autophosphorylations of CaM N2C0, N2C1 and N2C2 respectively.
-The front layer(Layer2) colored yellow represents interactions between Ca\textsuperscript{2+} and CaMKII attached CaM molecules.
-Layer3 represents interactions between Ca\textsuperscript{2+} and phosphorylated CaMKII attached CaM molecules.}
+	}
 \label{fig1}
@@
 \begin{figure}[!h]
-	\caption{{\bf}
-	Time courses of 9 states of CaM molecules. The smooth lines are simulated using COPASI, and the discontinous lines are from the simple stochastic models with two different diffusion coefficients.}
+	\caption{{\bf Test of Smoldyn modifications with a simple Ca\textsuperscript{2+}-CaM reaction network.}
+	Plots compare the number of CaM molecules in each of the 9 possible states as computed with the modified stochastic simulator and COPASI.
+Thick smooth lines are results using COPASI and thin black and red lines are results with the modified simulator for two different values of the calcium diffusion coefficient.
+	}
 \label{fig2}
 \end{figure}
+
+As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complex.
+As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system, assuming saturating amount of fully loaded CaM molecules are provided and subunit phosphorylations are allowed only from neighbors that are CaM bound but not those phosphorylated, at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme.
+In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numebrs, the fraction approaches $1-e^{-1}$.
+We built a model using a closed system with $\SI{1}{\um}\times\SI{1}{\um}\times\SI{1}{\um}$ geomemtry.
+6000 CaMKII subunits and 6020 fully loaded CaM were included in the system.
+Only the following reactions were allowed: bindings and unbindings between CaM and CaMKII, as well as phosphorylations by neighbor subunits that were bound but not phosphorylated.
+As expected, we obtained steady state phosphorylation levels that match the Michalski and Loew's formula.
+Finally, using the same closed system, we tested the complete network and compared with results obtained from a simulation implemented with spatial Gillespie algorithm by Zeng and Holmes~\cite{Zeng:2010bq}.
+The two simulations showed comparable results.
 
@@
 We set up a prototype model to identify the major network branches that contribute to the phosphorylation of CaMKII subunits.
-The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig1}.
+The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig3}A.
 In this network diagram molecule species are represented as vertices and reactions are represented as the network edges.
-CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in the $1 \times 1 \times 2$\SI{}{\cubic\um} box.
-As an initial contion, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII.
-The initial state is at an equilibrium, which we intentionally computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules (Fig~\ref{fig}).
-Ca\textsuperscript{2+} ions were released as a single source from the top $1 \times 1$ \SI{}{\square\um}.
-A previously generated input file having a total of 6 Ca2+ bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods).
+CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in a $\SI{1}{\um}\times\SI{1}{\um}\times\SI{2}{\um}$ box.
+As an initial condition, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII.
+The initial state is at an equilibrium, which we carefully computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules.
+For simplicity, we do not consider the resting level intracellular Ca\textsuperscript{2+} contributed significantly to the steady state initial condition.
+Extracellular Ca\textsuperscript{2+} ions were released as a single source from the top center of box ($\SI{1}{\um}\times$\SI{1}{\square\um}).
+A previously generated input file having a total of 6 Ca\textsuperscript{2+} bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods).
 We ran the simulation up to \SI{2.5}{\s}, recorded the molecule numbers at every \SI{1}{\ms} and logged all reaction events.
 Using the events log, we counted the accumulated occurrences of each reaction type at every \SI{10}{\ms}, starting from the initial steady-state until the Ca\textsuperscript{2+} bursts were over.
-
 For a particular reaction, the number of occurrences during a time span is considered to be the net number of molecule state changes along the corresponding edge, i.e., the number of unbinding events is subtracted from the number of binding events.
@@
 
-We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{}).
-The network consists of three layers (Fig~\ref{}).
-The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2) as shown separately in Fig~\ref{}.
+\begin{figure}[!h]
+	\caption{{\bf The complete Ca\textsuperscript{2+}-CaM-CaMKII network.}
+	A: Species are represented by vertices.
+Edges are labeled to indicate binding reactions.
+For example, "N1\_rxn" means binding of a Ca\textsuperscript{2+} ion on CaM at the N1 site.
+"KN1\_rxn" means CaM attached to a CaMKII subunit binds a Ca\textsuperscript{2+} ion on its N1 site.
+"K-NxCy\_rxn" means a CaMKII subunit binds a CaM that has x Ca\textsuperscript{2+} ions bound on the N-lobe and y Ca\textsuperscript{2+} ions bound on the C-lobe (x and y range from 0 to 2).
+"KpNxCy\_rxn" means a phosphorylated CaMKII subunit binds a CaM with x and y Ca\textsuperscript{2+} ions bound on the N and C lobes respectively.
+Red edges mark the predominant pathway.
+B: Layers shaded with blue colors.
+The back layer(Layer1) represents interactions between Ca\textsuperscript{2+} and CaM.
+The front layer(Layer2) represents interactions between Ca\textsuperscript{2+} and CaM attached to a CaMKII subunit.
+Layer3 represents interactions between Ca\textsuperscript{2+} and CaM attached to a phosphorylated CaMKII subunit.
+	}
+\label{fig3}
+\end{figure}
+
+We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{fig3}A).
+The network consists of three layers (Fig~\ref{fig3}B).
+The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2).
 The front layer (Layer2) is for reactions of Ca\textsuperscript{2+} with CaM attached to unphosphorylated CaMKII (KNxCy).
@@
 
+For each CaM state, we counted the accumulated occurrences of the possible reaction types.
+Results are shown in Fig~\ref{fig4}.
+The plot reveals a predominant pathway for CaM and CaMKII state transitions as labeled in Fig~\ref{fig3}A.
+Starting as apoCaM, a CaM molecule tends to bind a Ca\textsuperscript{2+} ion on the C1 site and then on the C2 site, entering the N0C2 state.
+CaM in the N0C2 state has a strong preference to enter layer 2 by binding to a CaMKII subunit, after which it quickly binds Ca\textsuperscript{2+} ions to the N1 and N2 sites to become KN2C2.
+Also evident from Fig~\ref{fig4} is that once CaM has two Ca\textsuperscript{2+} ions on either lobe, the preference is to bind to a CaMKII subunit before binding additional Ca\textsuperscript{2+} ions.
+Once bound to a CaMKII subunit, additional Ca\textsuperscript{2+} ions bind to CaM more quickly.
+In this scenario, it is rare for CaM to become fully bound with Ca\textsuperscript{2+} ions prior to binding with a CaMKII subunit.
+Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{fig5}.
+This preferred pathway is consistent with a hypothetical major pathway that leads to CaMKII autophosphorylation shown in Pepke et al.~\cite{Pepke:2010ju}.
+The observation is also consistent with Shifman et al.~\cite{Shifman:2006hw} in showing that partially loaded CaM molecules may be very important in activating CaMKII subunits.
+We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S2.
+For example, even though N sites bind Ca\textsuperscript{2+} faster, the C sites have higher affinity making Ca\textsuperscript{2+} binding to C sites preferred.
+
 \begin{figure}[!h]
-	\caption{{\bf}
-	Accumulated reaction occurrences over time for nine states of CaM molecules.
-For each type of CaM, three types of reactions can occur: binding of Ca\textsuperscript{2+} at N site, at C site or binding a CaMKII subunit.
-N1 represents Ca\textsuperscript{2+} binding on the N1 site; N2 means Ca\textsuperscript{2+} binding on the N2 site; C1 means Ca\textsuperscript{2+} binding on the C1 site; C2 means Ca\textsuperscript{2+} binding on the C2 site; K-NxCy means binding between CaMKII and CaM NxCy.
-Green arrows indicate the path consists of most likely chosen reactions starting from N0C0.
-Following the green arrows, labeled reaction leads to constructing the predominant pathway.}
-\label{fig3}
+	\caption{{\bf Accumulated reaction occurrences over time for Layer1 reactions.}
+	For a CaM binding state, 3 types of reactions were shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or binding of CaM to an unphosphorylated CaMKII subunit.
+Green arrows indicate reactions participating in the predominant pathway starting from N0C0.
+	}
+\label{fig4}
 \end{figure}
-***
-For each CaM state, we counted the accumulated occurrences of the possible reaction types.
-Results are shown in Fig~\ref{}.
-The plot reveals a predominant pathway for CaM and CaMKII state transitions.
-Starting as apoCaM, a CaM molecule tends to bind a Ca\textsuperscript{2+} ion on the C1 site and then on the C2 site, entering the N0C2 state.
-CaM in the N0C2 state has a strong preference to enter layer 2 by binding to a CaMKII subunit, after which it quickly binds Ca\textsuperscript{2+} ions to the N1 and N2 sites to become KN2C2 Fig~\ref{}.
-[Also evident from Fig~\ref{} is that once CaM has two Ca\textsuperscript{2+} ions on either lobe, the preference is to bind to a CaMKII subunit before binding additional Ca\textsuperscript{2+} ions.
-Once bound to a CaMKII subunit, additional Ca\textsuperscript{2+} ions bind to CaM more quickly.
-?] In this scenario, it is rare for CaM to become fully bound with Ca\textsuperscript{2+} ions prior to binding with a CaMKII subunit.
-Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{}.
-This preferred pathway is consistent with a hypothetical major pathway that leads to CaMKII autophosphorylation shown in Pepke et al.~\cite{Pepke:2010ju}.
-The observation is also consistent with ~\cite{} in showing that partially loaded CaM molecules may be very important in activating CaMKII subunits.
-We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities.
-For example, even though N sites bind Ca\textsuperscrtip{2+} faster, the C sites have higher affinity making Ca\textsuperscript{2+} binding to C sites preferred.
 
-To confirm the critical role of NxC2 (x=0,1,2) CaM in activation and phosphorylation of CaMKII, we set up two modified reaction schemes (Fig~\ref{}).
+\begin{figure}[!h]
+	\caption{{\bf Accumulated reaction occurrences over time for Layer2 reactions.}
+	For a CaM binding state, 3 types of reactions were shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or phosphorylation of the CaMKII that has the CaM bound.
+	}
+\label{fig5}
+\end{figure}
+
+To confirm the critical role of NxC2 (x=0,1,2) CaM in activation and phosphorylation of CaMKII, we set up two modified reaction schemes (Fig~\ref{fig6}A).
 In Scheme1, only CaMKII subunits bound with NxC2 are allowed to become phosphorylated.
 In Scheme2, phosphorylation is allowed only for subunits bound with N2Cx.
-In particular, the phosphorylation rates in the two schemes are equivalent, i.e., $k_{on}$ of reaction KpN2C0_rxn is the same as that of KpN0C2_rxn, and the $k_{on}$ of KN2C1_rxn equals that of KN1C2_rxn.
-Not surprisingly, the two schemes give rise to different phosphorylation levels as shown in Fig~\ref{}.
+Note the phosphorylation rates in the two schemes are equivalent, i.e., $k_{on}$ of reaction KpN2C0\_rxn is the same as that of KpN0C2\_rxn, and the $k_{on}$ of KpN2C1\_rxn equals that of KpN1C2\_rxn.
+Not surprisingly, the two schemes give rise to different phosphorylation levels as shown in Fig~\ref{fig6}B.
 Scheme 1 performs slightly worse than the whole network, whereas Scheme 2 produces much lower phosphorylation levels.
@@
 
-\begin{comment}
-\subsection*{CaMKII activation frequency dependence is not primarily determined by CaMKII holoenzyme structures}
-
-A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzymes activation are sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses.
-A recent study \textit{in vivo} also noticed glutate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}.
-It has been hypothesized that the holoenzyme structure requires cooperative phosphorylation between neighboring subunits and thus the structure contributes to the observed frequency dependence.
-However several other studies~\cite{Kubota:2001ul,Pepke:2010ju,Michalski:2012ds} suggest that the number of subunits of a holoenzyme does not result in significant difference in CaMKII activation pattern, therefore should not be related to the frequency dependence.
-Since multi-subunit CaMKII has not been modeled in fine details, we used the customized simulator to test the effects of subunits number by varying the number from 2, 6 to 12.
-For each subunit number, we used 5 Hz and 10 Hz Ca\textsuperscript{2+} input and simulated 10 trials for each frequency.
-The Ca\textsuperscript{2+} input files are generated from a 5 Hz and a 10 Hz theta-burst voltage file respectively, simulated using a stochastic Ca\textsuperscript{2+} channel model as described in Methods.
-%In particular, 40 trials of Ca\textsuperscript{2+} input files were generated for each voltage file and matched by total Ca\textsuperscript{2+} ions.
-The total Ca\textsuperscript{2+} amount difference between matched files is less than 10 Ca\textsuperscript{2+} ions.
-
-For a given subunit number, we consistently observed significantly higher phosphorylation level with 10 Hz than 5 Hz (Fig~\ref{fig4}B,C).
-Typically, 10 Hz input allows more accumulation of NxC2 type CaM, leading to more bound CaMKII subunits and thus more phosphorylations (Fig~\ref{fig4}A).
-The number of subunits affects the phosphorylation level but does not significantly contribute to the observed frequency effect(Fig~\ref{fig4}C).
-Specifically, the 2-subunits condition makes it easier for a holoenzyme to propagate phosphorylation between neighbors, thus resulting in the highest phosphorylation level under 10 Hz and 5 Hz.
-However, the fewer-subunits-easier-propagation advantage is not consistent between the 6-subunits and 12-subunits conditions.% The 6-subunits condition results in a higher phosphorylation level than the 12-subunits condition under 10 Hz but not consistently so under 5 Hz.
-
 \begin{figure}[!h]
-	\caption{{\bf}
-	A: Time courses of CaMKII subunits and CaM molecules when given 5 Hz Ca\textsuperscript{2+} influx generated from 16 channels.
-‘camkii\_b’ means CaM bound CaMKII subunits.
-‘camkii\_p’ means phosphorylated CaMKII subunits.
-‘KN2C2 + KpNxC2’ represents CaM molecules bound to either CaMKII or phosphorylated CaMKII and are fully Ca\textsuperscript{2+} loaded on C sites.
-B: Time courses of phosphorylated CaMKII subunits for 10 Hz (smooth lines) and 5 Hz input (dashed lines) with the complete network, reaction scheme1 and reaction scheme2 respectively.
-C: Time courses of phosphorylated CaMKII subunits for 10 Hz(fast rising) and 5 Hz input (slow rising) when one holoenzyme contains 2, 6 and 12 subunits respectively.
-D: With 5 Hz input, time courses of molecule number for subunits that are activated themselves (bound or phosphorylated) and together their neighbors.
-E: The same as in D except that 10 Hz input is used.}
-\label{fig4}
-\end{figure}
-\end{comment}
-
-\subsection*{CaMKII activation frequency dependence and the detailed Ca2+-CaM-CaMKII network mechanisms invovled}
-
-A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzymes activation are sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses.
-A recent study also showed that \textit{in vivo} glutamate uncaging frequency affects CaMKII activation in spine~\cite{Fujii:2013bg}.
-Numerous modeling studies have also reported a dependence of CaMKII activation on the frequency of calcium signals~\cite{}.
-To confirm this CaMKII activation pattern, we tested our 6-subunit CaMKII model using the previously generated 10 Hz and 5 Hz Ca\textsuperscript{2+} influx files (see Methods).
-The total number of influx Ca\textsuperscript{2+} ions are comparable (40433 ions in 5 Hz vs 40435 ions in 10 Hz).
-For each frequency, we simulated 10 trials and presented the time-varying mean +/- standard deviation of phosphorylated subunits.
-As shown in Fig~\ref{fig}, we observed significantly higher phosphorylation levels with 10 Hz than with 5 Hz input.
-
-We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between NxC2 CaM and CaMKII, and consequently more autophosphorylations (Fig~\ref{}).
-Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
-The frequency effect can be reversed by providing saturating amount of Ca\textsuperscript{2+}.
-To demonstrate, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential (Fig~\ref{fig5}).
-In the mean time, the total Ca\textsuperscript{2+} influx is kept consistent for 5 Hz and 10 Hz.
-As Ca\textsuperscript{2+} level increases, the network produces more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminishes.
-Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz input (Fig~\ref{fig5}F).
-
-Equivalently, the saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by limiting the available CaM and slowing Ca\textsuperscript{2+} diffusion (Fig~\ref{fig7}).
-To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
-Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed a reverse of frequency preference from 10 Hz to 5 Hz.
-The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
-We set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
-Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequencies.
-Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig7}C,D).
-In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference.
-It is also worth noting is that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig7}E,F).
-
-To explain what happened, we profiled the reaction occurrences over time on Layer1 edges.
-We noticed that the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} level becomes saturating.
-In the presence of moderate amount of Ca\textsuperscript{2+}, CaM molecules follow the predominant pathway by choosing reaction paths according to affinities.
-On one hand, bindings between N0C2 and CaMKII (rxn_KN0C2) are more likely to occur than bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules (rxn_N1).
-Similarly, CaMKII binding with N0C2 (rxn_KN0C2) dominates all types of CaMKII-CaM binding reactions (Fig~\ref{fig}).
-However, when \textsuperscript{2+} influx becomes saturating, CaM molecules tend to stay in the Layer1 till they get fully loaded by Ca\textsuperscript{2+}.
-As a result, CaMKII binding with N2C2 (rxn_KN2C2) becomes dominant.
-This decision change gives rise to an altered pathway and eventually a suboptimal phosphorylation level.
-Given the same total amount of Ca\textsuperscript{2+}, 10 Hz is easier to saturate CaM in a short time than 5 Hz.
-Correspondingly, the switch of pathway choice is more dramatic for 10 Hz at $4.5\times$ Ca\textsuperscript{2+}, resulting in the revserved frquency perference as observed previously.
-
-\begin{figure}[!h]
-	\caption{{\bf}
-	A: Using Ca\textsuperscript{2+} input generated with double times Ca2+\textsuperscript{2+} amount, accumulated reaction occurrences for autophosphorylation of CaMKII bound with N0C2, N1C2 and N2C2 respectively.
-Thick lines are for 10 Hz input; thin lines are for 5 Hz.
-B-E: The same as in A with Ca\textsuperscript{2+} input generated with $3\times$.
-$3.5\times$, $4\times$ and $6\times$ channels respectively.
-F: Summary plot of phosphorylation level with 10 Hz and 5 Hz input when various numbers of Ca\textsuperscript{2+} channels are involved in Ca\textsuperscript{2+} influx generation.}
-\label{fig5}
-\end{figure}
-
-\begin{comment}
-We examined the reaction occurrences profile over time on Layer1 edges (Fig~\ref{fig6}) and noticed that the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} level becomes saturating.
-This trend can be revealed from the pathway choice for CaM N0C2 (Fig~\ref{fig6}A-C).
-At nonsaturating Ca\textsuperscript{2+} level, N0C2 molecules are more likely to bind with CaMKII (reaction KN0C2) than binding a Ca\textsuperscript{2+} ion on the N1 site (reaction N1).
-When Ca\textsuperscript{2+} becomes saturating, N0C2 molecules tends to choose N1 binding over reaction KN0C2.
-This switch of reaction choice is observed for 10 Hz but not for 5 Hz.
-Likewise, we analyzed the profile of all possible CaMKII-CaM binding reactions.
-For 5 Hz conditions, binding of N0C2 consistently dominates over other CaMKII-CaM binding reactions, regardless of the amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig6}D-F).
-However for 10 Hz, as Ca\textsuperscript{2+} influx increases, binding of N2C2 (reaction KN2C2) gradually replaces binding of N0C2 (reaction KN0C2) and becomes the dominant choice for CaM to enter Layer2 (Fig~\ref{fig6}G-I).
-In breif, this switch of dominant pathway occurs for 10 Hz, correlating with the suboptimal phosphorylation level observed for 10 Hz.
-\end{comment}
-
-\begin{figure}[!h]
-	\caption{{\bf}
-	A-C: Reaction occurrences of N0C0 binding Ca\textsuperscript{2+} at the N1 site and reaction occurrences of N0C0 directly binding with CaMKII.
-Thick lines are for 10 Hz input, and thin lines are for 5 Hz input.
-D-F: Reaction occurrences for CaMKII binding with all 9 CaM states.
-5 Hz input generated with $1\times$, $3\times$ and $6\times$ Ca\textsuperscript{2+} influx are used.
-N0C2 are the dominant CaM state binding with CaMKII.
-Colored lines are for frequently occurred reactions.
-Grey lines are for infrequently occurred reactions.
-G-I: The same as in D-F except that 10 Hz input are used.
+	\caption{{\bf Reaction scheme 1 and 2 adapted from the complete network.}
+	A: Reaction scheme 1 and 2 to verify that CaMKII bound with CaM NxC2 are more invovled in phosphorylation than those bound with N2Cx.
+	B.
+Comparison of phosphorylation levels under complete network, scheme 1 and scheme 2.
+The results of the complete network were summarized from 15 trials of simulations and presented in mean phosphorylation +/- standard deviation.
 	} 
@@
 
+\subsection*{CaMKII activation frequency dependence and the detailed Ca\textsuperscript{2+}-CaM-CaMKII network mechanisms}
+
+A classic experiment by DeKonick and Schulman~\cite{DeKoninck:1998wh} showed that \textit{in vitro} CaMKII holoenzyme activation is sensitive to the frequency of Ca\textsuperscript{2+}-CaM pulses.
+A recent study~\cite{Fujii:2013bg} also showed that \textit{in vivo} glutamate uncaging frequency affects CaMKII activation in dendritic spines.
+Numerous modeling studies have also reported a dependence of CaMKII activation on the frequency of Ca\textsuperscript{2+} signals~\cite{Pepke:2010ju,Dupont:2003vq,Michalski:2012ds,Kubota:2001ul}.
+To confirm this CaMKII activation pattern, we tested our 6-subunit CaMKII model using previously generated 10 Hz and 5 Hz Ca\textsuperscript{2+} influx files.
+The total number of Ca\textsuperscript{2+} ions entering was comparable (40433 ions with 5 Hz vs 40435 ions with 10 Hz).
+As shown in Fig~\ref{fig7}, we observed much higher phosphorylation levels with 10 Hz than with 5 Hz input.
+
 \begin{figure}[!h]
-	\caption{{\bf}
-	A: Using $3.5\times$ Ca\textsuperscript{2+} input, accumulated reaction occurrences for autophosphorylation of CaMKII bound with N0C2, N1C2 and N2C2 respectively.
-Thick lines are for 10 Hz input; thin lines are for 5 Hz.
-B: The same as in panel A except that the total CaM available in the system is decreased to half amount.
-C: Reaction occurrences using Ca\textsuperscript{2+} input generated with $3\times$ Ca\textsuperscript{2+} input.
-D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{0.0000011}{\square\cm\per\s}.
-E: Reaction occurrences using twice Ca\textsuperscript{2+} input.
-F: The same as in C except with slowed Ca\textsuperscript{2+} diffusion and half amount of CaM.}
+	\caption{{\bf Comparison of 10 Hz and 5 Hz on CaMKII activation.}
+	A: Number of bound CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	B: Number of phosphorylated CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	}
 \label{fig7}
@@
 
-\subsection*{CaM changing dominant pathway can be demonstrated using a reduced network }
-
-To better understand what happens during the process, we used a reduced reaction network to capture the observed frequency prefrence reversion (Fig~\ref{fig8}B).
-The network is derived from the initial steps in the whole Ca\textsuperscript{2+}-CaM-CaMKII network.
-It is simple enough to be simulated using the original Smoldyn.
-We used 5 pulses of instantaneous Ca\textsuperscript{2+} release as the input instead of realistic Ca\textsuperscript{2+} influx and varied the amount of Ca\textsuperscript{2+} influx per pulse.
-The CaMKII molecules are modeled as monomers with a phosphorylation rate of \SI{1}{\per\s}.
-The 10 Hz stimulus becomes saturating and fails to generates more phosphorylation than 5 Hz when Ca\textsuperscript{2+} amount reaches 40000 ions per pulse.
-We separated the reduced network into two paths based on the sequence of Ca\textsuperscript{2+}-CaM-CaMKII binding (Fig~\ref{fig8}A).
-Simulations were carried out using each path alone to compare with the reduced network.
-When Ca\textsuperscript{2+} amount is not saturating, the reduced network shows a comparable phosphorylation level with that of Path2 (Fig~\ref{fig7}B,C).
-This pathway choice is consistent with the "follow-the-affinity" principle.
-Also the phosphorylation level increases linearly with the Ca\textsuperscript{2+} input amount.
-However when Ca\textsuperscript{2+} becomes saturating, the growth of phosphorylation level with incrasing Ca\textsuperscript{2+} amount slows down for both 10 Hz and 5 Hz (Fig~\ref{fig8}A).
-But the trend is considerable for 10 Hz and only slight for 5 Hz.
-Such an effect is mostly contributed from Path2 (Fig~\ref{fig8}B).
-When Path2 alone is present, the phosphorylation level as function of increasing Ca\textsuperscript{2+} input shows a faster increase for 5 Hz than 10 Hz.
-
+We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylation (Fig~\ref{fig8}C,D).
+Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
+The frequency effect can be reversed by providing saturating amount of Ca\textsuperscript{2+}.
+As demonstrated in Fig~\ref{fig9}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential.
+The total Ca\textsuperscript{2+} influx was again comparable for 5 Hz and 10 Hz.
+As Ca\textsuperscript{2+} influx was increased, the network produced more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminished.
+Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz input (Fig~\ref{fig9}F).
 \begin{figure}[!h]
-	\caption{{\bf}
-	A: The reduced reaction network derived from the whole network.
-CaMKII are modeled as monomers and can undergo phosphorylation at an arbitrary rate \SI{1}{\per\s} as long as bound with a CaM N1C2.
-B: Summary plot of phosphorylation level by varying Ca\textsuperscript{2+} ions amount per pulse from 10000 ions/pulse to 50000 ions/pulse.
-Dashed lines represent the slow of phosphorylation level growth as a function of Ca\textsuperscript{2+} influx.
-C: Summary plot of phosphorylation level with the various amount of Ca\textsuperscript{2+} ions delivered at either 5 Hz or 10 Hz.
-Phosphorylation levels are generated through Path1 or Path2 alone.
-D: In the presence of non-saturating Ca\textsuperscript{2+}(ca10000), time courses of N0C2 molecules when Path1 or Path2 alone is present, or both are present.
-E: The same as in D except that in the presence of saturating Ca\textsuperscript{2+}(ca50000).}
+	\caption{{\bf Effects of 10 Hz and 5 Hz on reaction occurrences.} 
+	A: Accumulated reaction occurrences for CaM CaMKII binding reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
+	B: The same as in A but for 10 Hz Ca\textsuperscript{2+} input.
+	C: Accumulated reaction occurrences for CaMKII phosphorylation reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
+	D: The same as in C but for 10 Hz Ca\textsuperscript{2+} input.
+	}
 \label{fig8}
@@
 
-The two paths branch from the node N0C2.
-That is, CaM N0C2 molecules need to make choices for binding parterners.
-How choices are made depend on the amount of each binding partner.
-When Ca\textsuperscript{2+} amount is scarce, Path2 dominates; when Ca\textsuperscript{2+} becomes saturating, Path1 competes with Path2.
-Hence inherently, there is a competition invovled between the two paths, which can alter the time course of N0C2 and eventually the phosphorylation level.
-To visualize the competition, we compared the time courses of N0C2 in the reduced network and those when Path1 or Path2 are recruited alone.
-Hypotheticall, when competitions exhist, the N0C2 molecules are in high demand.
-Hence the consumption of N0C2 increases when both Path1 and Path2 are present.
-In other words, the amount of N0C2 decreases more than when either Path is present alone.
-This is indeed the case when we examined the N0C2 amount in nonsaturating (10000 ions per pulse, Fig~\ref{fig8}D) and saturating Ca\textsuperscript{2+} influx (50000 ions per pulse, Fig~\ref{fig8}E).
-In the nonsaturating case, the time course of N0C2 of the reduced network matches well with that of the Path2 alone condition.
-N0C2 molecules in Path2 are not affected by the presence of Path 1.
-In contrast, in the saturating case, N0C2 amount becomes considerably lower when both paths are recruited together than when either path is present alone.
-This implies that the two paths negatively affect each other.
-Hence confirms the competition.
+\begin{figure}[!h]
+	\caption{{\bf Effects of various Ca\textsuperscript{2+} influx on frequency dependence.}
+	A-E: Accumulated reaction occurrences for phosphorylation of CaMKII subunits bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) with different levels of Ca\textsuperscript{2+} influx.
+	F: Summary of phosphorylation levels in the presence of varying amount of Ca\textsuperscript{2+} delivered at 5 Hz and 10 Hz.
+The advantage of 10 Hz on phosphorylation is reversed as Ca\textsuperscript{2+} influx increases.
+	}
+\label{fig9}
+\end{figure}
+
+Equivalently, the saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by limiting the available CaM and slowing Ca\textsuperscript{2+} diffusion (Fig~\ref{fig10}).
+To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
+Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed that the 5 Hz input results in more phosphorylation reactions than 10 Hz(Fig~\ref{fig10}A,B).
+The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
+We set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
+Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequencies.
+Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig10}C,D).
+In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference.
+It is also worth noting that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig10}E,F).
+
+\begin{figure}[!h]
+	\caption{{\bf Effects of CaM availability and Ca\textsuperscript{2+} diffusion on frequency dependence.}
+	A: With $3\times$ Ca\textsuperscript{2+} influx, accumulated reaction occurrences for CaMKII phosphorylation bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	B: The same as in A except that the total CaM available in the system is decreased to half. 
+	C: The same as in A except that $3.5\times$ Ca\textsuperscript{2+} is provided.
+	D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{0.0000011}{\square\cm\per\s}. 
+	E: The same as in A except that $2\times$ Ca\textsuperscript{2+} is provided.
+	F: The same as in E except with slowed Ca\textsuperscript{2+} diffusion and half amount of CaM.
+	}
+\label{fig10}
+\end{figure}
+
+To understand why frequency dependence changed, we examined the reaction occurrences over time on Layer1 edges (Fig~\ref{fig11}).
+We noticed that regardless of the input frequency, the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} levels become saturating.
+In the presence of moderate Ca\textsuperscript{2+} influx (Fig~\ref{fig11}A,B), CaM molecules follow the predominant pathway by choosing reaction paths according to affinities: bindings between N0C2 and CaMKII are more likely to occur than bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules.
+CaMKII binding with N0C2 dominates all types of CaMKII-CaM binding reactions (Fig~\ref{fig11}D,E,G).
+However, when Ca\textsuperscript{2+} influx becomes saturating, bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules dominate over the bindings between N0C2 and CaMKII (Fig~\ref{fig11}C).
+CaM molecules tend to stay in the Layer1 untill they get fully loaded by Ca\textsuperscript{2+}.
+As a result, CaMKII binding with N2C2 becomes dominant over all other CaM CaMKII binding reactions (Fig~\ref{fig11}F,H,I).
+This decision change gives rise to an altered pathway and eventually a suboptimal phosphorylation level.
+Given the same total amount of Ca\textsuperscript{2+}, 10 Hz influx saturates CaM in a shorter time than 5 Hz.
+Correspondingly, the switch of pathway choice is more dramatic for 10 Hz at $4.5\times$ Ca\textsuperscript{2+}, resulting in the reversed frequency perference as observed previously.
+
+\begin{figure}[!h]
+	\caption{{\bf Change of pathway choice as Ca\textsuperscript{2+} influx increases}
+	A: With $2\times$ Ca\textsuperscript{2+} input, reaction occurrences of CaM N0C2 binding Ca\textsuperscript{2+} at the N1 site and directly binding with CaMKII.
+Thick lines are for 10 Hz input, and thin lines are for 5 Hz input.
+B: The same as in A except that $3\times$ Ca\textsuperscript{2+} is given.
+C: The same as in C except that $4.5\times$ Ca\textsuperscript{2+} is given.
+D-F: Accumulated reaction occurrences for CaM CaMKII binding reactions under varying amount of total Ca\textsuperscript{2+} influx from $2\times$, $3\times$ to $4.5\times$.
+Ca\textsuperscript{2+} inputs are delivered at 5 Hz.
+G-I: The same as in D-F except that Ca\textsuperscript{2+} inputs are delivered at 10 Hz.
+	}
+\label{fig11}
+\end{figure}
+
+To validate our conclusion about the pathway decision change, we used a reduced reaction network to capture the observed frequency preference reversal (Fig~\ref{fig12}).
+The reduced network is derived from the initial steps in the whole Ca\textsuperscript{2+}-CaM-CaMKII network.
+We used the original Smoldyn to simulate the simple network, in order to demonstrate that the reversal of frequency preference is not an artifact of our modified simulator but is inherent to the network itself.
+We used 5 pulses of instantaneous Ca\textsuperscript{2+} release as the input and varied the amount of Ca\textsuperscript{2+} influx per pulse.
+The CaMKII molecules are modeled as monomers with an arbirary phosphorylation rate of \SI{1}{\per\s}.
+As expected, the reversal of frequency preference can be qualitatively captured by the simple reaction scheme (Fig~\ref{fig12}A).
+The 10 Hz stimulus becomes saturating and fails to generate more phosphorylation than 5 Hz when Ca\textsuperscript{2+} influx reaches 40000 ions per pulse.
+
+\begin{figure}[!h]
+	\caption{{\bf Change of pathway choice demonstracted using a reduced network}
+	A: The reduced reaction network derived from the complete network.
+CaMKII are modeled as monomers and can undergo phosphorylation at an arbitrary rate \SI{1}{\per\s} as long as bound with a CaM N1C2.
+	B: Summary of phosphorylation levels in the presence of varying Ca\textsuperscript{2+} ions per pulse from 10000 ions/pulse to 50000 ions/pulse delivered at 5 Hz and 10 Hz.
+	}
+\label{fig12}
+\end{figure}
+
+\subsection*{The structural orgnization of CaMKII subunits}
+
+Recent work on CaMKII holoenzyme structure suggests that how subunits are organized can affect the activation of the holoenzyme.
+In particular, whether subunits are arranged in a compact or an extended way can change the accessibility of Ca\textsuperscript{2+}-CaM to CaMKII.
+It is known that the structural arrangement is related to a linker region between a subunit's kinase domain and the holoenzyme central hub.
+Bayer et al.~\cite{Bayer:2002er} examined splice variants of $\beta$-CaMKII.
+These variants have identical kinase domains yet different linker lengths.
+They responded to Ca\textsuperscript{2+} oscillations differently, even though they showed no difference to prolonged Ca\textsuperscript{2+}-CaM input.
+In particular, for Ca\textsuperscript{2+} pulse input, the variants with longer-linker length exhibited a higher autophosphorylation rate.
+Another study by Chao et al.~\cite{Chao:2011iw} also indicated that the linker length affects the capability of a subunit's kinase domain to undock from the central hub.
+Undocking helps the subunit to release from autoinhibited state.
+Thus a longer linker can keep the kinase domain further away from the central hub, allowing a better chance for Ca\textsuperscript{2+}-CaM to access the binding site.
+
+In our prototype model, holoenzyme subunits have distinct physical locations and are separated by \SI{8}{\nm} from its neighbors.
+To examine such a structural effect on CaMKII activation, we varied the radius of an one-ring shaped holoenzyme from \SI{5}{\nm} to \SI{15}{\nm}.
+The holoenzyme consists of 6 euqally spaced subunits.
+Neighbor subunits are spaced the same length as the holoenzyme radius.
+According to the literature~\cite{Chao:2011iw,Gaertner:2004jk}, a typical one-ring holoenzyme radius is within \SI{5}{\nm} to \SI{8}{\nm}.
+For each radius condition, we tested 15 simulation trials using the default Ca\textsuperscript{2+} influx delivered at 5 Hz.
+As shown in Fig~\ref{fig13}, there is a slight but distinguishable increase in phosphorylation level as the radius of the holoenzyme increases.
+
+\begin{figure}[!h]
+	\caption{{\bf Effects of holoenzyme size on CaMKII phosphorylation.}
+	Phophorylation levels as a function of the holoenzyme radius presented in mean+standard deviation. Each radius condition is summarized from 15 trials.
+	}
+\label{fig13}
+\end{figure}
 
@@
 In particular, the pathway through N0C1, N0C2, and KN0C2 plays a crucial role.
-Second, when studying the effects of Ca\textsuperscript{2+} input frequency, we noticed that the number of subunits does not significantly affect the patterns of CaMKII.
-The mechanism should reside within the network kinetics.
-These notions are consistent with previous modeling studies ~\cite{Kubota:2001ul,Pepke:2010ju,Michalski:2012ds}.
-Third, CaM availablility is factor for CaMKII frequency response, as suggested in ~\cite{DeKoninck:1998wh}.
-Given that free diffusible CaM molecules are limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}, this conclusion further implies the regulatory role of of many endogenous CaM binding proteins ~\cite{Skene:1990kf}.
-Fourth, factors such as crowding and spatial homogeneity can slow down molecule diffusion and thus may have substantial effects on CaMKII phosphorylation.
-In our simulations, Ca\textsuperscript{2+} diffusion slowed down by half could lead to a dramatic increase of phosphorylation level.
-A similar effect of slow diffusion on CaMKII activation has also been shown in a spine model~\cite{Bhalla:2004cu}.
-In addition, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.
+Second, factors such as CaM availablility~\cite{DeKoninck:1998wh} and Ca\textsuperscript{2+} diffusion can affect the CaMKII frequency response.
+Experiment studies suggest that free diffusible CaM molecules are higly limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}.
+The limited CaM further implies a regulatory role of of many endogenous CaM binding proteins ~\cite{Skene:1990kf}.
+Third, intracellular crowding and spatial homogeneity can slow down molecule diffusion.
+For example, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.
+This complication in Ca\textsuperscript{2+} diffusion may have substantial effects on CaMKII phosphorylation and frequency response.
+Fourth, we demonstrated that the holoenzyme size can affect the level of phosphorylation.
+This is consistent with the idea that the configuration of a holoenzyme, whether compact or extended, affects CaM molecules accessing the CaMKII subunits.
 
@@
 %stratton etal, subunits exchange 
-Our model does not contain Thr305/Thr306 phosphorylation and some newly discovered mechanisms of CaMKII, which may explain the special multi-subunit strucutre of the holoenzyme.
-For example, it has been found that there exists a compact autoinhibition state, which occurs through dimerization of adjacent subunits from top and bottom rings.
+Our model does not contain Thr305/Thr306 phosphorylation and some newly discovered mechanisms of CaMKII, which may lead to more complicated activation pattern of CaMKII holoenzymes.
+For example, it has been found that there exists a compact autoinhibition state, which occurs through dimerization of adjacent subunits from top and bottom rings~\cite{Chao:2010fn}.
 Once Ca\textsuperscript{2+} bound to a dimerized subunit, the dimer dissembles and the two subunits are released.
@@
 
-%Our finding indicates the CaM amount is important for CaMKII frequency dependence.
-It is known that free diffusible apoCaM molecules are limited in cells~\cite{}.
-Although there are abundant intracellular CaM molecules, most of them are captured by CaM binding proteins.
-The distribution of CaM is spatially heterogeneous as well.
-It is estimated that CaM molecules are enriched upto mM level concentration in areas close to the Ca\textsuperscript{2+} channels~\cite{Mori:2004fb}.
-These extra details making CaM regulation a complicated process and the output of Ca\textsuperscript{2+} signaling network necessarily depends on the specific biophysical context in which the network is located.
-
-%Our approach to model multi-subunit macromolecules allows binding sites to act independently, therefore potentially different from simply expanding reaction rules and loading the reactions for simulation.
-A molecule can undergo multiple state changes within a time step as long as the binding sites involved are independent of each other.
-For example, apoCaM N0C0 can become N1C1 within a time step, since N1 and C1 act independently; but cannot become N0C2 or N2C0, because N2 and C2 act cooperatively on N1 and C1 respectively.
-But since our time step is sufficiently small, the discrepancy between the implementations stay conceptual and does not lead to different steady states.
-
-One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of high molecule concentrations.
+One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of highly concetrated molecules.
 One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM are acting very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in modeling studies.
@@
 However for these fast kinetics, there currently lacks an efficient method to simulate.
-If to simulate with the original Smoldyn algorithm, the simulation time step has to be considerablly reduced to obtain the correct steady state~\cite{Andrews:2015}.
+If using the original Smoldyn algorithm, the simulation time step has to be considerablly reduced to obtain the correct steady state~\cite{Andrews:2015}.
 Another software package using an enhanced Green's Function Algorithm~\cite{vanZon:2005jd} can handle the high concentration diffusion-limited reactions accurately, but it takes an impractically long time to simulate.
@@
 Each molecule has a vector to store the states of binding sites.
-All reactions are stored in a hashtable with reactants and binding states as entry keys.
-As an example, binding reactions involving the N and C lobes of a CaM molecule can be coded as in Fig~\ref{fig}.
-CaMKII holoenzymes are two 6-subunits rings.
-But for simplicity, we modeled each CaMKII holoenzyme as only one ring at a radius of \SI{8}{\nm}.
-Each subunit is separated from its direct neighbors at a fixed distance \SI{8}{\nm} (estimated from ~\cite{Gaertner:2004jk}) and has a distinct physical location.
+All reactions are stored in a hash table with reactants and their binding states as entry keys.
+A hash table is a data structure to store association arrays as well as rapid lookup.
+In our case, the reaction network can be considered as associations between reactants, therefore ideal to be impemented using a hash table.
+CaM is an example of a molecule with multiple binding sites.
+The binding reactions involving the N and C lobes of a CaM can be coded as in Fig~\ref{fig1}B.
+A CaMKII holoenzyme is an example of a complex composed of two 6-subunits rings.
+Each subunit is a molecule containing binding sites for calmodulin and phosphorylation.
+Each ring has a radius of \SI{8}{\nm} and is separated from its direct neighbors at a fixed distance of \SI{8}{\nm} (estimated from~\cite{Gaertner:2004jk}).
+For simplicity, we modeled CaMKII holoenzymes as one ring of 6 subunits.
 
-In the original version of Smoldyn, each reaction generates new molecules as products and removes reactants.
+In the original version of Smoldyn, each reaction generates new molecules as products and reactant molecules are removed.
 In our case, since one molecule can have multiple binding sites and is potentially associated with multiple partners, entirely removing a molecule is not practical because other attached molecules would also be affected.
-Also, removing and generating new molecules makes it difficult to track the reaction history of a molecule.
-Therefore, during reactions we do not remove molecules, but merely change molecule identities and positions.
+In addition, removing and generating new molecules makes it difficult to track the reaction history of a molecule.
+Therefore, during reactions we do not remove molecules, but merely change molecule binding states and positions.
 Molecules bound together physically overlap and synchronize their locations automatically.
-When they diffuse, the diffusion coefficient is determined by the dominant molecule.
-For example, when Ca\textsuperscript{2+} and CaM are bound, Ca\textsuperscript{2+} molecules diffuse with CaM.
+When molecules diffuse, the diffusion coefficient is determined by the dominant molecule.
+For example, when Ca\textsuperscript{2+} and CaM are bound, the attached Ca\textsuperscript{2+} molecule diffuse with the CaM diffusion rate; similarly, CaM bound to a CaMKII subunit will diffuse with the CaMKII diffusion rate.
 
@@
 For example, CaM has 4 Ca\textsuperscript{2+} binding sites.
-Since the N and C sites act independently, the N1 and C1 sites compete for binding Ca\textsuperscript{2+} and an apoCaM (N0C0) can become either N1C0 or N0C1, resulting in a branching reaction scheme.
+Since the N and C sites act independently, the N1 and C1 sites compete for Ca\textsuperscript{2+} and an apoCaM N0C0 can become either N1C0 or N0C1, resulting in a branching reaction scheme.
 Thus a decision process is needed to choose a reaction path when a binding event occurs.
 
-As an example, consider the following two reactions
+To do this, consider the following two reactions
 \begin{quote}
@@
 Rxn1 has a forward rate constant $k_{f1}$ in \SI{}{\per\micro\Molar\per\s} and a backward rate constant $k_{b1}$ \SI{}{\per\s}.
-Rxn2 has rate constants $k_{f2}$ and $k_{b2}$.
-The two reactions can be viewed together as rxn3, which has an overall kinetics $k_{f3}$ and $k_{b3}$.
-According to the law of mass action, the two reactions can be written in to differential equations and the following equations hold:
+Rxn2 has a similar rate constants $k_{f2}$ and $k_{b2}$.
+The two reactions can be viewed together as an equivalent rxn3, which has overall kinetics rates $k_{f3}$ and $k_{b3}$.
+According to the law of mass action, the reactions can be written as differential equations and we can obtain
 \begin{equation}
@@
 \end{equation}
-\begin{equation}\frac{k_{f3}}{k_{b3}}=\frac{k_{f2}}{k_{b2}} + \frac{k_{f1}}{k_{b1}}\end{equation}
-\begin{equation}k_{b3}=pk_{b1} + (1-p)k_{b2}\end{equation}
 
-where $p$ represents the proportion in $k_{b3}$ is due to unbinding of camN1C0.
-Knowing that $\frac{p}{1-p}=\frac{[camN1C0]}{[camN0C1]}=\frac{k_{f1}/k_{b1}}{k_{f2}/k_{b2}}$, we obtain $k_{f3}=k_{f1}+k_{f2}$.
 Smoldyn uses binding radii to implement second order reactions.
@@
 If the number falls in the range $(0,\frac{kf_1}{kf_3}]$, then rxn1 is chosen; instead if the number falls in the range $(\frac{kf_1}{kf_3}, 1]$, we pick rxn2.
-Following this approach, the steady state of the reaction network can be kept consistent with the prediction by the mass action law.
+Following this approach, the network can be kept consistent with the prediction by the mass action law.
 
@@
 We focus on the interactions among Ca\textsuperscript{2+}, CaM, and CaMKII.
-We first used a Ca\textsuperscript{2+}-CaM network for testing to confirm that modifications on the simulator are valid.
-Then we added CaMKII holoenzymes to study the signaling network in detailes (Fig~\ref{fig1}C).
-Our rule-based approach allows reactions to be described at the binding site level and perform simulations without compromising the resolution of chemical kinetics.
-We set up a cube-shaped model to represent a chunk from a cell body.
-The cell is assumed as a sphere with radius \SI{5}{\micro\meter}.
-The cube has a dimension \SI{1}{\um} in width and \SI{2}{\um} in depth (Fig~\ref{fig9}A).
-The top surface of the cube is the cell membrane, reflective to all molecules.
-The four sides are also reflective to prevent molecul1es from escaping laterally through membrane surface of side compartments.
-The reflective boundary is essentially equivalent to periodic boundary in terms of conservation of molecules in lateral direction.
+We first used a Ca\textsuperscript{2+}-CaM network for testing to confirm that modifications to the simulator were working properly.
+Then we added CaMKII holoenzymes to study the reaction network in details.
+We set up a cube-shaped model to represent a portion of a cell body (Fig~\ref{fig14}A).
+The cube has dimensions of \SI{1}{\um} in width and \SI{2}{\um} in depth.
+The top surface of the cube represents the cell membrane, reflective to all molecules.
+The four sides are also reflective.
 The bottom surface is partially absorbing to Ca\textsuperscript{2+} ions but reflective to CaM.
-Reflective to CaM guarantees a steady state initial condition.
-The partial absorption is built-in in Smoldyn to resemble an unbounded diffusion situation~\cite{Andrews:2009dr}.
+This conservation of CaM molecules and CaMKII subunits guarantees a steady state initial condition.
+This partial absorption is a built-in feature in the original Smoldyn to resemble unbounded diffusion~\cite{Andrews:2009dr}.
 
 \begin{figure}[!h]
-	\caption{{\bf}
-	A: Dimension of the cube-shaped model.
-Ca\textsuperscript{2+} channels are located on the top surface.
-B: Find the number channels by fitting the single channel current to the GHK equation.
-N equals 6 by fitting.
-C: $n_{inf}(v)$ and $\tau_n(v)$ used in the voltage-gated Ca\textsuperscript{2+} channels.}
-\label{fig9}
+	\caption{{\bf Model geometry and Ca\textsuperscript{2+} parameters.}
+	A: Dimensions of the cube-shaped model. Ca\textsuperscript{2+} channels are located on the top surface. 
+	B: $n_{inf}(v)$ and $\tau_n(v)$ used in the voltage-gated Ca\textsuperscript{2+} channels.
+	C: The number channels N by fitting the single channel current to the GHK equation. N equals 16 by fitting. 
+	}
+\label{fig14}
 \end{figure}
 
-Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type CaL) are located on the top surface to provide Ca\textsuperscript{2+} influx.
-For simplicity, these channels are clustered at the center of the membrane.
-The channels open and close depending on a time-varying membrane voltage file generated from a neuron model.
-CaMKII subunits ($\beta$-subunits) are uniformly present at a concentration of \SI{10}{\umol}.
-They are also immobilized presumably by attaching to actins~\cite{Li:2016cq}.
-Free diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}.
-This is consistent with the notion that at the cell resting level free diffusible CaM is considerably limited~\cite{Tran:2003fs,2008BpJ....95.6002S,LubyPhelps:1995kl}.
-Table S1 lists all the reactions with corresponding kinetic parameters involved in the network.
-Kinetic parameters are integrated from various sources and are adjusted to satisfy microscopic reversibility.
-%However, they are highly concentrated (\SI{245}{\umol}) in the nanodomain attached to Ng molecules.
-The high concentration of CaM is consistent with the finding~\cite{Mori:2004fb} that L-type Ca\textsuperscript{2+} entry is enriched with CaM.
-Besides,
+Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type) are located on the top surface to provide Ca\textsuperscript{2+} influx.
+For simplicity, these channels are located at the center of the membrane.
+The channels open and close depending on a time-varying membrane voltage file generated from a neuron model (described below).
+CaMKII subunits (typically $\beta$-subunits) are uniformly present at a concentration of \SI{10}{\umol}.
+They are also immobilized so they are presumed as attached to actins~\cite{Li:2016cq}.
+Freely diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}.
+This is consistent with the notion that at the resting level, freely diffusible CaM molecules are considerably limited in number compared to their binding proteins~\cite{Tran:2003fs,2008BpJ....95.6002S,LubyPhelps:1995kl}.
+Table S2 lists all the reactions with corresponding kinetic parameters involved in the network.
+Kinetic parameters are integrated from various sources as noted in Table S2 and are adjusted to satisfy microscopic reversibility.
 
 \subsection*{Ca\textsuperscript{2+} input conditions}
-We used a previously built multi-compartment neuron model to generate the somatic action potential trains at frequencies of 5 Hz and 10 Hz~\cite{Li:2014fv}.
+
+We used a previously constructed multi-compartment neuron model to generate the somatic action potential trains at frequencies of 5 Hz and 10 Hz~\cite{Li:2014fv}.
 The action potentials are activated by synaptic stimulations, following a series of presynaptic release of neurotransmitters which was modeled with a probability profile measured in experiments~\cite{Grover:2009hb}.
-Presynaptic stimulations activate spine AMPA receptors, elevate the membrane potential and open NMDA receptors, allowing Ca\textsuperscript{2+} ions to enter the spines.
-Accumulated Ca\textsuperscript{2+} influx further depolarizes the dendrite tree and rapidly propagate toward the soma, initiating somatic action potentials.
+Presynaptic stimulations activate spine AMPA receptors, elevate the membrane potential and open NMDA receptors, allowing Ca\textsuperscript{2+} ions to enter the spines and depolarize the dendrite tree.
+The depolarization rapidly propagates toward the soma, initiating somatic action potentials.
 The pattern of action potentials follows the synaptic stimulation.
-We varied the intervals of synaptic release to generate action potentials in the form of theta-burst with different inter-bursts intervals, which provide different Ca\textsuperscript{2+} influx frequencies for the signaling network.
+We varied the intervals of synaptic release to generate action potentials in the form of theta-burst with different inter-bursts intervals, which provide different Ca\textsuperscript{2+} influx frequencies for the reaction network (Fig~\ref{fig15}A,C).
 
-The cell level neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances.
-Its membrane voltage output was applied to the Ca\textsuperscript{2+} signaling model (Fig~\ref{fig10}A,C).
-Since L-type Ca\textsuperscript{2+} channels are relatively fast, the membrane potentials are unaffected by the channel activities, and the neuron model and the signaling model can be safely decoupled.
-The L-type channels in the signaling model are implemented as single channels with voltage-depdendent stochastic opening and closing.
-The rates of open and close are functions of membrane voltage and calculated using varialbles of $n$ and $\tau$ (Fig~\ref{fig9}C) similar to Hudgkin-Huxley equations~\cite{Tuckwell:2012tt}.
+\begin{figure}[!h]
+	\caption{{\bf Patterns of membrane voltage and Ca\textsuperscript{2+} influx. }
+	A: A 5 Hz action potential trains voltage file generated from the NEURON model. 
+	B: Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file.
+Ca\textsuperscript{2+} influx is stochastic.
+There are spontaneous influx in the absence of action potentials.
+	C: 10 Hz action potential trains voltage file generated from the NEURON model. 
+	D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. 
+	E: Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} influx files.
+	}
+\label{fig15}
+\end{figure}
+
+The neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances.
+Its membrane voltage output was applied to the Ca\textsuperscript{2+} reaction network model.
+Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast, the membrane potentials are unaffected by the channel activities.
+Thus the neuron model and the reaction network model can be safely decoupled.
+In the reaction network model, the stochastic Ca\textsuperscript{2+} channels open and close in response to the voltage input to provide Ca\textsuperscript influx.
+The voltage-depdendent opening and closing of these channels are modeled in a Hudgkin-Huxley way~\cite{Tuckwell:2012tt}.
+The rates of open and close are functions of membrane voltage and calculated using varialbles of $n$ and $\tau_n$, where $n$ describes voltage-dependent activation and $\tau_n$ is the time constant of $n$ (Fig~\ref{fig15}B).
+The channel obeys a simple first order kinetics as follows.
 
@@
 \begin{equation}rate_{close}(V)=\frac{1-n_{\inf}}{\tau_n}\end{equation}
-\begin{equation}n_{\inf}=\frac{1}{1+exp(\frac{V+V_{half}}{slope})}\end{equation}
-\begin{equation}\tau_n=\tau_0+\frac{4\tau_a\sqrt{\tau_g(1-\tau_g)}}{exp(-D_V(1-\tau_g)k_f/s)+exp(D_V\tau_gk_f/s)}\end{equation}
+\begin{equation}n_{\inf}=\frac{1}{1+exp(\frac{V+V_{1/2}}{slope})}\end{equation}
+\begin{equation}\tau_n=0.06+3\times\frac{\sqrt{0.45\times0.55}}{exp((V+V_{1/2})\times0.55/slope)+exp(-(V+V_{1/2})\times0.45/slope)}\end{equation}
+
+In particular, $n_{\inf}$ is the steady state value of $n$.
+$V_{1/2}$ and $slope$ together describe the channel activation in response to voltage.
+In our model, $V_{1/2}$ equals \SI{-15}{\mV} and $slope$ equals 8.0.
 
@@
 Since the membrane surface is \SI{1}{\um\square}, the current density is converted to a total current $I_{ghk}$ for the area.
-Assuming a total number of N channels are present, for each channel $\frac{I_{ghk}}{N}$ should equal $i_{ca}$.
-By fitting the two expressions, we obtained a N of 16 and a $V_s$ of \SI{-1.90955}{\milli\volt} (Fig~\ref{fig9}B).
-Using the $i_{ca}$, the number of ions entered during one time step is calcuated as $\frac{i_{ca}}{2e}dt$, where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
-To speed up the simulation, Ca\textsuperscript{2+} influx patterns were generated in a separate simulation using the voltage file (Fig~\ref{fig10}B,D).
-Then the generated Ca\textsuperscript{2+} influx files were fed into the signaling network.
-To guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then matched them by total Ca\textsuperscript{2+}.
+On the other hand, assuming a total number of N channels are present, for each channel $\frac{I_{ghk}}{N}$ a single channel current is $i_{ca}$.
+By fitting the total current $I_{ghk}$ and $N\times i_{ca}$, we obtained a N of 16 and a $V_s$ of \SI{-1.90955}{\milli\volt} (Fig~\ref{fig15}C).
+Using the $i_{ca}$, the number of ions entered during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
+In order to guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig15}E).
 
-\begin{figure}[!h]
-	\caption{{\bf}
-	A: A 5 Hz action potential trains voltage file generated from the NEURON model.
-B: Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file.
-Ca\textsuperscript{2+} influx is stochastic.
-There are spontaneous firings in the absence of action potentials.
-C: 10 Hz action potential trains voltage file generated from the NEURON model.
-D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file.
-Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively.
-The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} influx files.}
-\label{fig10}
-\end{figure}
 \section*{Conclusion}

### 2017-06-25 paper.tex
methods updated

@@
 
-We used a previously constructed multi-compartment neuron model to generate the somatic action potential trains at frequencies of 5 Hz and 10 Hz~\cite{Li:2014fv}.
-The action potentials are activated by synaptic stimulations, following a series of presynaptic release of neurotransmitters which was modeled with a probability profile measured in experiments~\cite{Grover:2009hb}.
-Presynaptic stimulations activate spine AMPA receptors, elevate the membrane potential and open NMDA receptors, allowing Ca\textsuperscript{2+} ions to enter the spines and depolarize the dendrite tree.
-The depolarization rapidly propagates toward the soma, initiating somatic action potentials.
-The pattern of action potentials follows the synaptic stimulation.
-We varied the intervals of synaptic release to generate action potentials in the form of theta-burst with different inter-bursts intervals, which provide different Ca\textsuperscript{2+} influx frequencies for the reaction network (Fig~\ref{fig15}A,C).
+A neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances.
+Its membrane voltage output was applied to the Ca\textsuperscript{2+} reaction network model.
+Since the kinetics of L-type Ca\textsuperscri with detailed morphology of a CA1 pyramidal cell and ion channel conductances to generate a voltage response at the soma to various stimlation conditions.
+Theta-burst stimulation (5 pulses at 100 Hz, repeated 5 times at 5 Hz or 10 Hz intervals) was applied to synapses on spines taking into account a probability of release measured in experiments~\cite{Grover:2009hb}.
+This stimulation activated AMPA and NMDA receptor-channels on dendritic spines causing depolarization in the dendritic tree that propagate to the soma and initiated action potentials.
+Voltage profiles for 5 Hz and 10 Hz intervals stimulations are shown in Fig~\ref{fig15}A,C.
 
@@
 	\caption{{\bf Patterns of membrane voltage and Ca\textsuperscript{2+} influx. }
-	A: A 5 Hz action potential trains voltage file generated from the NEURON model. 
+	A: A 5 Hz action potential burst voltage file generated from the NEURON model. 
 	B: Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file.
 Ca\textsuperscript{2+} influx is stochastic.
-There are spontaneous influx in the absence of action potentials.
-	C: 10 Hz action potential trains voltage file generated from the NEURON model. 
+There is spontaneous influx in the absence of action potentials.
+	C: 10 Hz action potential burst voltage file generated from the NEURON model. 
 	D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. 
@@
 
-The neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances.
-Its membrane voltage output was applied to the Ca\textsuperscript{2+} reaction network model.
-Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast, the membrane potentials are unaffected by the channel activities.
+This membrane voltage output for neuron model was used to determine Ca\textsuperscript{2+} influx through L-type channels in our reaction network model.
+Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast and their density is low the membrane potential is not affected by their activities.
 Thus the neuron model and the reaction network model can be safely decoupled.
-In the reaction network model, the stochastic Ca\textsuperscript{2+} channels open and close in response to the voltage input to provide Ca\textsuperscript influx.
-The voltage-depdendent opening and closing of these channels are modeled in a Hudgkin-Huxley way~\cite{Tuckwell:2012tt}.
-The rates of open and close are functions of membrane voltage and calculated using varialbles of $n$ and $\tau_n$, where $n$ describes voltage-dependent activation and $\tau_n$ is the time constant of $n$ (Fig~\ref{fig15}B).
+L-type Ca\textsuperscript{2+} channels are modeled stochastically and they open and close in response to the voltage input to provide Ca\textsuperscript{2+} influx.
+The voltage-depdendent opening and closing of these channels are modeled with the Hudgkin-Huxley formation~\cite{Tuckwell:2012tt}.
+The rates of channel opening and closing are functions of membrane voltage and calculated using variables $n$ and $\tau_n$, where $n$ describes voltage-dependent activation and $\tau_n$ is the time constant of $n$ (Fig~\ref{fig16}B).
 The channel obeys a simple first order kinetics as follows.
@@
 \begin{equation}rate_{close}(V)=\frac{1-n_{\inf}}{\tau_n}\end{equation}
-\begin{equation}n_{\inf}=\frac{1}{1+exp(\frac{V+V_{1/2}}{slope})}\end{equation}
-\begin{equation}\tau_n=0.06+3\times\frac{\sqrt{0.45\times0.55}}{exp((V+V_{1/2})\times0.55/slope)+exp(-(V+V_{1/2})\times0.45/slope)}\end{equation}
+\begin{equation}n_{\inf}=\frac{1}{1+exp(-\frac{V-V_{1/2}}{slope})}\end{equation}
+\begin{equation}\tau_n=0.06+4\times\frac{0.75\times\sqrt{0.45\times0.55}}{exp((V-V_{1/2})\times0.55/slope)+exp(-(V-V_{1/2})\times0.45/slope)}\end{equation}
 
@@
 
-For each channel, at a given time, a probability is calculated based on the membrane voltage to decided whether a channel opens.
+For each channel, at a given time, a probability is calculated based on the membrane voltage to decide whether a channel opens.
 If it opens, a varying number of Ca\textsuperscript{2+} ions are generated.
@@
 A current density is calculated using GHK current equation~\cite{citeulike:306134} assuming an extracellular [Ca\textsuperscript{2+}] of \SI{2}{\m\Molar} and an intracellular concentration of \SI{50}{\nano\Molar}.
-Since the membrane surface is \SI{1}{\um\square}, the current density is converted to a total current $I_{ghk}$ for the area.
+Since the membrane surface is \SI{1}{\square\um}, the current density is converted to a total current $I_{ghk}$ for the area.
 On the other hand, assuming a total number of N channels are present, for each channel $\frac{I_{ghk}}{N}$ a single channel current is $i_{ca}$.
-By fitting the total current $I_{ghk}$ and $N\times i_{ca}$, we obtained a N of 16 and a $V_s$ of \SI{-1.90955}{\milli\volt} (Fig~\ref{fig15}C).
-Using the $i_{ca}$, the number of ions entered during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
+By fitting the total current $I_{ghk}$ and $N\times i_{ca}$, we obtained a N of 16 and a $V_s$ of \SI{-1.91}{\milli\volt} (Fig~\ref{fig15}C).
+Using the $i_{ca}$, the number of ions entering one channel during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
 In order to guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig15}E).

### 2017-07-02 paper.tex
july-01

@@
 In addition, the size of the holoenzyme shows a subtle effect on CaMKII activation.
-The more extended the subunits are organized, the easier for CaM molecules to access and activate the subunits.
+The more extended the subunits are organized, the easier for calmodulin molecules to access and activate the subunits.
 The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca\textsuperscript{2+} channel nanodomains and cytoplasm.
@@
 We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods.
-Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca\textsuperscript{2+} input conditions, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.
+Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence relies on Ca\textsuperscript{2+} input amount, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.
 
@@
 CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}.
-In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus.
+In iaddition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus.
 Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors.
@@
 Reactions are defined between binding sites.
-The reaction network is stored in a hashtable to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network.
+The reaction network is stored in a hash table to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network.
 We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}.
@@
 The size of this volume mimics that of a Ca\textsuperscript{2+} channel nanodomain or a small dendritic spine head.
-Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.867}{\micro\Molar} and 9.302\SI{}{\micro\Molar} respectively.
+Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.867}{\micro\Molar} and \SI{9.302}{\micro\Molar} respectively.
 The concentrations are chosen to produce an observable amount of 4-Ca\textsuperscript{2+} bound CaM at the steady state.
@@
 On the right-hand side, binding sites involved in the reaction are assigned to new values.
-Notice the ‘==‘ sign is used on the left-hand side, but the ‘=‘ sign is used on the right-hand side.
-The ‘==‘ represents True or False of equality, whereas the ‘=‘ denotes an assigning operation.
+Notice the "==" sign is used on the left-hand side, but the "=" sign is used on the right-hand side.
+The "==" represents True or False of equality, whereas the "=" denotes an assigning operation.
 	}
@@
 
-As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complex.
+As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complexes.
 As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system, assuming saturating amount of fully loaded CaM molecules are provided and subunit phosphorylations are allowed only from neighbors that are CaM bound but not those phosphorylated, at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme.
@@
 This preferred pathway is consistent with a hypothetical major pathway that leads to CaMKII autophosphorylation shown in Pepke et al.~\cite{Pepke:2010ju}.
-The observation is also consistent with Shifman et al.~\cite{Shifman:2006hw} in showing that partially loaded CaM molecules may be very important in activating CaMKII subunits.
+The observation is also consistent with Shifman et al.~\cite{Shifman:2006hw} in that partially loaded CaM molecules may be very important in activating CaMKII subunits.
 We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S2.
@@
 	For a CaM binding state, 3 types of reactions were shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or binding of CaM to an unphosphorylated CaMKII subunit.
-Green arrows indicate reactions participating in the predominant pathway starting from N0C0.
+Green arrows indicate reactions involved in the predominant pathway starting from N0C0.
 	}
@@
 Therefore, edges of the network are not equally involved in CaMKII phosphorylation.
-Phosphorylation from KNxC2 is especially important.
 
@@
 	\caption{{\bf Comparison of 10 Hz and 5 Hz on CaMKII activation.}
-	A: Number of bound CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	A: Number of bound CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input conditions.
 	B: Number of phosphorylated CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
@@
 
-We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylation (Fig~\ref{fig8}C,D).
+We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylations (Fig~\ref{fig8}C,D).
 Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
 The frequency effect can be reversed by providing saturating amount of Ca\textsuperscript{2+}.
-As demonstrated in Fig~\ref{fig9}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential.
+As demonstrated in Fig~\ref{fig9}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential pulse.
 The total Ca\textsuperscript{2+} influx was again comparable for 5 Hz and 10 Hz.
@@
 To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
-Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed that the 5 Hz input results in more phosphorylation reactions than 10 Hz(Fig~\ref{fig10}A,B).
+Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed that the 5 Hz input results in more phosphorylation reactions than 10 Hz (Fig~\ref{fig10}A,B).
 The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
-We set Ca\textsuperscript{2+} diffusion as \SI{0.0000011}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
+We set Ca\textsuperscript{2+} diffusion as \SI{1.1e-6}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
 Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequencies.
-Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition(Fig~\ref{fig10}C,D).
-In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided by $3\times$ becomes sufficiently saturating to change the frequency preference.
+Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition (Fig~\ref{fig10}C,D).
+In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided at $3\times$ becomes sufficiently saturating to change the frequency preference.
 It is also worth noting that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig10}E,F).
@@
 	C: The same as in A except that $3.5\times$ Ca\textsuperscript{2+} is provided.
-	D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{0.0000011}{\square\cm\per\s}. 
+	D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{1.1e-6}{\square\cm\per\s}. 
 	E: The same as in A except that $2\times$ Ca\textsuperscript{2+} is provided.
@@
 	B: Summary of phosphorylation levels in the presence of varying Ca\textsuperscript{2+} ions per pulse from 10000 ions/pulse to 50000 ions/pulse delivered at 5 Hz and 10 Hz.
+Data from 5 trials were included.
 	}
@@
 Recent work on CaMKII holoenzyme structure suggests that how subunits are organized can affect the activation of the holoenzyme.
-In particular, whether subunits are arranged in a compact or an extended way can change the accessibility of Ca\textsuperscript{2+}-CaM to CaMKII.
+In particular, whether subunits are arranged in a compact or an extended way can change the accessibility of CaM to CaMKII.
 It is known that the structural arrangement is related to a linker region between a subunit's kinase domain and the holoenzyme central hub.
@@
 Another study by Chao et al.~\cite{Chao:2011iw} also indicated that the linker length affects the capability of a subunit's kinase domain to undock from the central hub.
-Undocking helps the subunit to release from autoinhibited state.
+Undocking helps the subunit to release from an autoinhibited state.
 Thus a longer linker can keep the kinase domain further away from the central hub, allowing a better chance for Ca\textsuperscript{2+}-CaM to access the binding site.
@@
 To examine such a structural effect on CaMKII activation, we varied the radius of an one-ring shaped holoenzyme from \SI{5}{\nm} to \SI{15}{\nm}.
-The holoenzyme consists of 6 euqally spaced subunits.
+The holoenzyme consists of 6 equally spaced subunits.
 Neighbor subunits are spaced the same length as the holoenzyme radius.
@@
 	\caption{{\bf Effects of holoenzyme size on CaMKII phosphorylation.}
-	Phophorylation levels as a function of the holoenzyme radius presented in mean+standard deviation. Each radius condition is summarized from 15 trials.
+	Phophorylation levels as a function of the holoenzyme radius presented in mean + standard deviation. Each radius condition is summarized from 15 trials.
 	}
@@
 In the study, we presented an efficient approach to simulate multi-subunit molecules and reaction network in detailed kinetics, as well as the effects of diffusion on the reaction network.
+Our adaption of the Smodlyn preserves its particle-based simulation feature, allowing us to include diffusion and spatial heterogeneity in models.
+Also, we added new data structures to describe reactions at the binding sites level, leading to an efficient solution of combinatorial explosion.
 We also introduced an intuitive approach to analyzing the pathway choices of a network based on the reaction history of a simulation, allowing us to grasp insights of a large network quickly.
+In traditional analysis, the number of molecule species are examined to resolve the dynamics of the network.
+However, the method is indirect when it comes to draw insights of the network behaviors.
+Also, being able to track the reaction history of the system is not possible in deterministic simulations such as ODE based ones.
+Here we demonstrated that the reaction history can be highly informative and allows us to see clearly a favored pathway of a network under different input conditions.
 
-Our findings agree with observations from previous studies.
+This approach allows us to obtain several insights into the Ca\textsuperscript{2+}-CaM-CaMKII netowrk.
 First, under physiological conditions when Ca\textsuperscript{2+} influx is moderate, CaM molecules partially loaded with Ca\textsuperscript{2+} are important for CaMKII activation.
-In particular, the pathway through N0C1, N0C2, and KN0C2 plays a crucial role.
-Second, factors such as CaM availablility~\cite{DeKoninck:1998wh} and Ca\textsuperscript{2+} diffusion can affect the CaMKII frequency response.
-Experiment studies suggest that free diffusible CaM molecules are higly limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}.
-The limited CaM further implies a regulatory role of of many endogenous CaM binding proteins ~\cite{Skene:1990kf}.
-Third, intracellular crowding and spatial homogeneity can slow down molecule diffusion.
+In particular, CaM molecules that have 2 Ca\textsuperscript{2+} attached on the C lobe (including N0C2, N1C2 and N2C2) are preferentially bound to CaMKII subunits.
+This is consistent with the predominant pathway hypothesis suggested in Pepke et al.~\cite{Pepke:2010ju}.
+Second, we found that factors such as CaM availablility~\cite{DeKoninck:1998wh} and Ca\textsuperscript{2+} diffusion can affect the CaMKII frequency response.
+Limited amount of CaM makes the given Ca\textsuperscript{2+} input more likely to saturate the available CaM.
+Likewise, slow Ca\textsuperscript{2+} diffusion allows more extensive interactions between Ca\textsuperscript{2+} and CaM, thus more likely for the given amount of Ca\textsuperscript{2+} input to become saturating.
+Experimental studies suggest that the number of freely diffusible CaM molecules is higly limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}.
+The limited CaM further implies a regulatory role for the many endogenous CaM binding proteins.
+For example, Skene et al.~\cite{Skene:1990kf} suggested that a CaM binding protein invovled in axonal grwoth cone GAP-43 can act as a "CaM sponge" by sequesting a fair amount of CaM under submembrane area and releasing these CaM molecules in response to protein kinase C activity, thereby modulating the development of axonal growth cones.
+Third, it is known that intracellular crowding and spatial homogeneity can slow down molecule diffusion.
 For example, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.
-This complication in Ca\textsuperscript{2+} diffusion may have substantial effects on CaMKII phosphorylation and frequency response.
+We believe that such a complication in Ca\textsuperscript{2+} diffusion may have substantial effects on CaMKII phosphorylation and frequency response.
+For example, depending on the size of the nandomain, slow Ca\textsuperscript{2+} diffusion in a nanodomain can potentially result in a localized Ca\textsuperscript{2+} signal with sufficient strength to activate downstream cascading proteins.
 Fourth, we demonstrated that the holoenzyme size can affect the level of phosphorylation.
-This is consistent with the idea that the configuration of a holoenzyme, whether compact or extended, affects CaM molecules accessing the CaMKII subunits.
+We increased the size of the holoenzyme by changing the distance between neighbor subunits and found that the phosphorylation level of the netowrk increased slightly accordingly.
+This is consistent with the idea that the configuration of a holoenzyme, whether compact or extended, affects CaM molecules accessing the CaMKII subunits~\cite{Stratton:2013el}.
 
 %chao etal, autoinhibited through dimerization;
-%stratton etal, subunits exchange 
-Our model does not contain Thr305/Thr306 phosphorylation and some newly discovered mechanisms of CaMKII, which may lead to more complicated activation pattern of CaMKII holoenzymes.
+%stratton etal, subunit exchange 
+Our model does not contain Thr305/Thr306 phosphorylation (few subunits would have phosphoryated during the timeframe simulated here) and also lacks some newly discovered mechanisms of CaMKII structural features, which may lead to a more complicated activation pattern of CaMKII holoenzymes.
 For example, it has been found that there exists a compact autoinhibition state, which occurs through dimerization of adjacent subunits from top and bottom rings~\cite{Chao:2010fn}.
-Once Ca\textsuperscript{2+} bound to a dimerized subunit, the dimer dissembles and the two subunits are released.
-Another recent study indicated that phosphorylated CaMKII subunits can undergo subunits exchange to faciliate proprogating the activation triggered by Ca\textsuperscript{2+}-CaM~\cite{Stratton:2014ct}.
+Once Ca\textsuperscript{2+} is bound to a dimerized subunit, the dimer disassembles and the two subunits are popped away from the center of the holoenzyme.
+Another recent study indicated that phosphorylated CaMKII subunits can undergo subunit exchange to faciliate propagating activation triggered by Ca\textsuperscript{2+}-CaM~\cite{Stratton:2014ct}.
+The significance of these additional features of CaMKII activation awaits future study.
 
-One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of highly concetrated molecules.
-One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM are acting very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in modeling studies.
-These binding kinetics are in the diffusion-limited regime, rendering traditional mass action based methods inaccurate.
-However for these fast kinetics, there currently lacks an efficient method to simulate.
-If using the original Smoldyn algorithm, the simulation time step has to be considerablly reduced to obtain the correct steady state~\cite{Andrews:2015}.
+One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of highly concentrated molecules.
+One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM act very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in modeling studies (although see ~\cite{Mironov:2013} for a critique of their estimates).
+These binding kinetics, if accurate, would place these reactions in the diffusion-limited regime, rendering traditional mass action based methods inaccurate.
+However for kinetics this fast, adequte simulation options are limited and not efficient.
+If using the original Smoldyn algorithm, the simulation time step has to be considerably reduced to obtain the correct steady state~\cite{Andrews:2015}; alternatively, we may increase the geminiate recombination probability~\cite{Andrews:2004fs}.
 Another software package using an enhanced Green's Function Algorithm~\cite{vanZon:2005jd} can handle the high concentration diffusion-limited reactions accurately, but it takes an impractically long time to simulate.
-It is interesting to examine the Ca\textsuperscript{2+}-CaM-CaMKII using the fast kinetics parameters and we anticipate to address this question in future development.
+If diffusion is slowed considerably in local nanodomains such that Ca\textsuperscript{2+}-CaM-CaMKII interactions become diffusion-limited, it will be necessary to develop different algorithms to handle these interactions accurately.
 
@@
 All reactions are stored in a hash table with reactants and their binding states as entry keys.
-A hash table is a data structure to store association arrays as well as rapid lookup.
-In our case, the reaction network can be considered as associations between reactants, therefore ideal to be impemented using a hash table.
+A hash table is a data structure that stores association arrays and allows rapid lookup.
+In our case, the reaction network can be considered as associations between reactants, therefore are ideal to be implemented using a hash table.
 CaM is an example of a molecule with multiple binding sites.
@@
 A CaMKII holoenzyme is an example of a complex composed of two 6-subunits rings.
-Each subunit is a molecule containing binding sites for calmodulin and phosphorylation.
+Each subunit is a molecule containing binding sites for CaM and phosphorylation.
 Each ring has a radius of \SI{8}{\nm} and is separated from its direct neighbors at a fixed distance of \SI{8}{\nm} (estimated from~\cite{Gaertner:2004jk}).
@@
 Therefore, during reactions we do not remove molecules, but merely change molecule binding states and positions.
-Molecules bound together physically overlap and synchronize their locations automatically.
-When molecules diffuse, the diffusion coefficient is determined by the dominant molecule.
-For example, when Ca\textsuperscript{2+} and CaM are bound, the attached Ca\textsuperscript{2+} molecule diffuse with the CaM diffusion rate; similarly, CaM bound to a CaMKII subunit will diffuse with the CaMKII diffusion rate.
+Molecules bound together physically overlap, ynchronize their locations automatically and diffuse together.
+The diffusion coefficient is determined by the dominant molecule.
+For example, when Ca\textsuperscript{2+} and CaM are bound, the attached Ca\textsuperscript{2+} molecule diffuse with the CaM diffusion rate; similarly, CaM bound to a CaMKII subunit will diffuse with the CaMKII.
 
@@
 Rxn1 has a forward rate constant $k_{f1}$ in \SI{}{\per\micro\Molar\per\s} and a backward rate constant $k_{b1}$ \SI{}{\per\s}.
-Rxn2 has a similar rate constants $k_{f2}$ and $k_{b2}$.
+Rxn2 has similar rate constants $k_{f2}$ and $k_{b2}$.
 The two reactions can be viewed together as an equivalent rxn3, which has overall kinetics rates $k_{f3}$ and $k_{b3}$.
@@
 \begin{figure}[!h]
-	\caption{{\bf Model geometry and Ca\textsuperscript{2+} parameters.}
+	\caption{{\bf Model geometry and Ca\textsuperscript{2+} channel parameters.}
 	A: Dimensions of the cube-shaped model. Ca\textsuperscript{2+} channels are located on the top surface. 
-	B: $n_{inf}(v)$ and $\tau_n(v)$ used in the voltage-gated Ca\textsuperscript{2+} channels.
-	C: The number channels N by fitting the single channel current to the GHK equation. N equals 16 by fitting. 
+	B: $n_{inf}(V)$ and $\tau_n(V)$ used in the voltage-gated Ca\textsuperscript{2+} channels.
+	C: The number channels N by fitting the single channel current to the GHK equation. N was found to be 16. 
 	}
@@
 Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type) are located on the top surface to provide Ca\textsuperscript{2+} influx.
-For simplicity, these channels are located at the center of the membrane.
+For simplicity, these channels are placed together at the center of the membrane.
 The channels open and close depending on a time-varying membrane voltage file generated from a neuron model (described below).
 CaMKII subunits (typically $\beta$-subunits) are uniformly present at a concentration of \SI{10}{\umol}.
-They are also immobilized so they are presumed as attached to actins~\cite{Li:2016cq}.
+They are also immobilized so presumlably they attached to actins~\cite{Li:2016cq}.
 Freely diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}.
@@
 Its membrane voltage output was applied to the Ca\textsuperscript{2+} reaction network model.
-Since the kinetics of L-type Ca\textsuperscri with detailed morphology of a CA1 pyramidal cell and ion channel conductances to generate a voltage response at the soma to various stimlation conditions.
+Since the kinetics of L-type Ca\textsuperscript{2+} with detailed morphology of a CA1 pyramidal cell and ion channel conductances to generate a voltage response at the soma to various stimulation conditions.
 Theta-burst stimulation (5 pulses at 100 Hz, repeated 5 times at 5 Hz or 10 Hz intervals) was applied to synapses on spines taking into account a probability of release measured in experiments~\cite{Grover:2009hb}.
-This stimulation activated AMPA and NMDA receptor-channels on dendritic spines causing depolarization in the dendritic tree that propagate to the soma and initiated action potentials.
+This stimulation activated AMPA and NMDA receptor-channels on dendritic spines causing depolarizations in the dendritic tree, which propagated to the soma and initiated action potentials.
 Voltage profiles for 5 Hz and 10 Hz intervals stimulations are shown in Fig~\ref{fig15}A,C.
@@
 	D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. 
-	E: Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} influx files.
+	E: Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} input files.
 	}
@@
 This membrane voltage output for neuron model was used to determine Ca\textsuperscript{2+} influx through L-type channels in our reaction network model.
-Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast and their density is low the membrane potential is not affected by their activities.
+Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast and their density is low, the membrane potential is not affected by their activities.
 Thus the neuron model and the reaction network model can be safely decoupled.
-L-type Ca\textsuperscript{2+} channels are modeled stochastically and they open and close in response to the voltage input to provide Ca\textsuperscript{2+} influx.
+L-type Ca\textsuperscript{2+} channels are modeled stochastically and they open and close in response to the voltage input.
 The voltage-depdendent opening and closing of these channels are modeled with the Hudgkin-Huxley formation~\cite{Tuckwell:2012tt}.
@@
 $V_{1/2}$ and $slope$ together describe the channel activation in response to voltage.
-In our model, $V_{1/2}$ equals \SI{-15}{\mV} and $slope$ equals 8.0.
+In our model, $V_{1/2}$ equals \SI{-15}{\mV} and $slope$ equals 8.
 
@@
 A current density is calculated using GHK current equation~\cite{citeulike:306134} assuming an extracellular [Ca\textsuperscript{2+}] of \SI{2}{\m\Molar} and an intracellular concentration of \SI{50}{\nano\Molar}.
+The maximum channel permeation is \SI{0.241e-3}{\cm\per\s}.
 Since the membrane surface is \SI{1}{\square\um}, the current density is converted to a total current $I_{ghk}$ for the area.
@@
 By fitting the total current $I_{ghk}$ and $N\times i_{ca}$, we obtained a N of 16 and a $V_s$ of \SI{-1.91}{\milli\volt} (Fig~\ref{fig15}C).
-Using the $i_{ca}$, the number of ions entering one channel during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
-In order to guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig15}E).
+Using the $i_{ca}$, the number of ions entering one channel during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{fig15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
+To guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig15}E).
 
-\section*{Conclusion}
+%\section*{Conclusion}
 \section*{Supporting information}

### 2017-07-02 paper.tex
july-01 revised

@@
 \section*{Abstract}
-Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons and understanding how this occurs has been the focus of many modeling studies.
-However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled.
+Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons.
+Understanding how this occurs has been the focus of many modeling studies.
+However, CaMKII is notoriously difficult simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled.
 To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion.
@@
 Depending on whether Ca\textsuperscript{2+} influx is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns.
-In addition, the size of the holoenzyme shows a subtle effect on CaMKII activation.
+In addition, the size of the holoenzyme produces a subtle effect on CaMKII activation.
 The more extended the subunits are organized, the easier for calmodulin molecules to access and activate the subunits.
@@
 CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}.
-In iaddition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus.
+In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus.
 Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors.
@@
 
-The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings.
-Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase.
-Activities such as CaM binding and auto-phosphorylation result in various subunits states.
+The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings~\cite{}.
+Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase~\cite{}.
+Activities such as CaM binding and auto-phosphorylation result in various subunit states.
 For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site.
@@
 It would be extremely time-consuming and not practical for BioNetGen to generate this network.
-To overcome this problem, most previous modeling studies of CaMKII simplified the Ca\textsuperscript{2+}-CaM-CaMKII reaction network by either modeling CaMKII as monomers~\cite{Bhalla:2004cu,Pepke:2010ju} or allowing only CaM fully-loaded with Ca\textsuperscript{2+} to interact with CaMKII subunits~\cite{Zhabotinsky:2000fp,Kubota:2001ul,Dupont:2003vq,Michalski:2012ds}.
+To overcome this problem, most previous modeling studies of CaMKII have simplified the Ca\textsuperscript{2+}-CaM-CaMKII reaction network by either modeling CaMKII as monomers~\cite{Bhalla:2004cu,Pepke:2010ju} or allowing only CaM fully-loaded with Ca\textsuperscript{2+} to interact with CaMKII subunits~\cite{Zhabotinsky:2000fp,Kubota:2001ul,Dupont:2003vq,Michalski:2012ds}.
 
@@
 
-In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus.
+In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (Ca\textsubscript{V}) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus.
 Specifically, we examined Ca\textsuperscript{2+}-CaM-CaMKII network activity near L-type Ca\textsuperscript{2+} channels.
@@
 If the encountering step takes a much longer time to occur than the reacting step, the reaction is diffusion-limited; otherwise, the reaction is activation-limited.
-Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{Bamford198505},
+Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{Rice:198505},
 \begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
@@
 Given the diffusion coefficients for Ca\textsuperscript{2+} and CaM and the parameter values for Ca\textsuperscript{2+} and CaM reaction kinetics in Table S2, we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime.
-This was also true if diffusion constants were reduced 10 fold.
+This was also true if diffusion constants were reduced 10-fold.
 Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.
@@
 We coded a deterministic ODE system in COPASI~\cite{Hoops:2006gy} and compared results with our stochastic model using the same kinetics.
-We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{fig2}.
+We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as shown in Fig~\ref{fig2}.
 Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used.
@@
 ‘+’ represents a binding reaction; ‘$\sim$’ represents molecules that are bound.
-‘\{\}’ on the left-hand side specifies the states of binding sites of reactants or the conditions for a reaction to occur.
+‘\{\}’ on the left-hand side specifies the binding states of the reactant binding sites or the conditions for a reaction to occur.
 On the right-hand side, binding sites involved in the reaction are assigned to new values.
@@
 As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complexes.
-As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system, assuming saturating amount of fully loaded CaM molecules are provided and subunit phosphorylations are allowed only from neighbors that are CaM bound but not those phosphorylated, at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme.
-In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numebrs, the fraction approaches $1-e^{-1}$.
-We built a model using a closed system with $\SI{1}{\um}\times\SI{1}{\um}\times\SI{1}{\um}$ geomemtry.
-6000 CaMKII subunits and 6020 fully loaded CaM were included in the system.
+As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system with a saturating amount of fully loaded CaM molecules assuming subunit phosphorylation is allowed only from a neighbor that is CaM bound but not from one that is phosphorylated, then at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme.
+In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numbers, the fraction approaches $1-e^{-1}$.
+We built a model using a closed system with $\SI{1}{\um}\times\SI{1}{\um}\times\SI{1}{\um}$ geometry.
+Six thousand CaMKII subunits and 6020 fully loaded CaM were included in the system.
 Only the following reactions were allowed: bindings and unbindings between CaM and CaMKII, as well as phosphorylations by neighbor subunits that were bound but not phosphorylated.
-As expected, we obtained steady state phosphorylation levels that match the Michalski and Loew's formula.
-Finally, using the same closed system, we tested the complete network and compared with results obtained from a simulation implemented with spatial Gillespie algorithm by Zeng and Holmes~\cite{Zeng:2010bq}.
-The two simulations showed comparable results.
+As expected, we obtained steady state phosphorylation levels that match the Michalski and Loew formula.
+
+Finally, using the same closed system, we tested the complete network and compared results with those obtained from a simulation implemented with a spatial Gillespie algorithm~\cite{Zeng:2010bq}.
+The two methods showed comparable results.
 
@@
 CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in a $\SI{1}{\um}\times\SI{1}{\um}\times\SI{2}{\um}$ box.
-As an initial condition, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII.
+As an initial condition, the box contained 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII.
 The initial state is at an equilibrium, which we carefully computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules.
-For simplicity, we do not consider the resting level intracellular Ca\textsuperscript{2+} contributed significantly to the steady state initial condition.
-Extracellular Ca\textsuperscript{2+} ions were released as a single source from the top center of box ($\SI{1}{\um}\times$\SI{1}{\square\um}).
+For simplicity, we did not consider resting intracellular Ca\textsuperscript{2+} in the steady state initial condition, as calculations showed its effect to be minor.
+Ca\textsuperscript{2+} influx was modeled as entry from a single source at the top center of the box.
 A previously generated input file having a total of 6 Ca\textsuperscript{2+} bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods).
@@
 "K-NxCy\_rxn" means a CaMKII subunit binds a CaM that has x Ca\textsuperscript{2+} ions bound on the N-lobe and y Ca\textsuperscript{2+} ions bound on the C-lobe (x and y range from 0 to 2).
-"KpNxCy\_rxn" means a phosphorylated CaMKII subunit binds a CaM with x and y Ca\textsuperscript{2+} ions bound on the N and C lobes respectively.
+"Kp-NxCy\_rxn" means a phosphorylated CaMKII subunit binds a CaM with x and y Ca\textsuperscript{2+} ions bound on the N and C lobes respectively.
+"KpNxCy\_rxn" means phosphorylation of a KNxCy subunit.
 Red edges mark the predominant pathway.
@@
 Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{fig5}.
-This preferred pathway is consistent with a hypothetical major pathway that leads to CaMKII autophosphorylation shown in Pepke et al.~\cite{Pepke:2010ju}.
-The observation is also consistent with Shifman et al.~\cite{Shifman:2006hw} in that partially loaded CaM molecules may be very important in activating CaMKII subunits.
-We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S2.
+We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S2 chosen to be consistent with experimental or modeling studies~\cite{Pepke:2010ju,Zeng:2010bq,Lucic:2008gt}.
 For example, even though N sites bind Ca\textsuperscript{2+} faster, the C sites have higher affinity making Ca\textsuperscript{2+} binding to C sites preferred.
@@
 	\caption{{\bf Accumulated reaction occurrences over time for Layer1 reactions.}
-	For a CaM binding state, 3 types of reactions were shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or binding of CaM to an unphosphorylated CaMKII subunit.
+	For a CaM binding state, 3 types of reactions are shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or binding of CaM to an unphosphorylated CaMKII subunit.
 Green arrows indicate reactions involved in the predominant pathway starting from N0C0.
@@
 	\caption{{\bf Accumulated reaction occurrences over time for Layer2 reactions.}
-	For a CaM binding state, 3 types of reactions were shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or phosphorylation of the CaMKII that has the CaM bound.
+	For a CaM binding state, 3 types of reactions are shown: binding of Ca\textsuperscript{2+} at an N site, binding of Ca\textsuperscript{2+} at a C site or phosphorylation of the CaMKII that has the CaM bound.
 	}
@@
 	\caption{{\bf Reaction scheme 1 and 2 adapted from the complete network.}
-	A: Reaction scheme 1 and 2 to verify that CaMKII bound with CaM NxC2 are more invovled in phosphorylation than those bound with N2Cx.
+	A: In reaction scheme 1, phosphorylation was allowed only from CaMKII bound with NxC2 (blue paths). In reaction scheme 2, phosphorylation was allowed only from CaMKII bound with N2Xc (green paths).
 	B.
-Comparison of phosphorylation levels under complete network, scheme 1 and scheme 2.
-The results of the complete network were summarized from 15 trials of simulations and presented in mean phosphorylation +/- standard deviation.
+Comparison of phosphorylation levels with the complete network, scheme 1 and scheme 2.
+The results of the complete network were summarized from 15 trials of simulations and presented as mean phosphorylation +/- standard deviation.
 	} 
@@
 	\caption{{\bf Comparison of 10 Hz and 5 Hz on CaMKII activation.}
-	A: Number of bound CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input conditions.
-	B: Number of phosphorylated CaMKII subunits (including bound and phosphorylated) under 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	A: Number of bound CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input conditions.
+	B: Number of phosphorylated CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
 	}
@@
 
-We examined the reaction occurrences over time and noticed that the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylations (Fig~\ref{fig8}C,D).
+We examined the reaction occurrences over time and noticed that compared to 5 Hz input the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylations (Fig~\ref{fig8}C,D).
 Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
-The frequency effect can be reversed by providing saturating amount of Ca\textsuperscript{2+}.
+However, the frequency effect can be reversed by providing a saturating amount of Ca\textsuperscript{2+}.
 As demonstrated in Fig~\ref{fig9}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential pulse.
@@
 As Ca\textsuperscript{2+} influx was increased, the network produced more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminished.
-Eventually, the 10 Hz input becomes saturating and the network generates less phosphorylation under 10 Hz than the 5 Hz input (Fig~\ref{fig9}F).
+Eventually, the 10 Hz input became saturating and the network generated less phosphorylation with 10 Hz than with 5 Hz input (Fig~\ref{fig9}F).
 \begin{figure}[!h]
@@
 
-Equivalently, the saturating Ca\textsuperscript{2+} effects on frequency preference can also be facilitated by limiting the available CaM and slowing Ca\textsuperscript{2+} diffusion (Fig~\ref{fig10}).
-To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default amount of CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
-Even though the Ca\textsuperscript{2+} input does not change, by limiting CaM we noticed that the 5 Hz input results in more phosphorylation reactions than 10 Hz (Fig~\ref{fig10}A,B).
+Equivalently, a reversal in frequency preference is also seen when available CaM is limited or Ca\textsuperscript{2+} diffusion is slowed (Fig~\ref{fig10}).
+To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
+Even though the Ca\textsuperscript{2+} input did not change, limiting CaM results in more phosphorylation reactions with 5 Hz than with 10 Hz Ca\textsuperscript{2+} input (Fig~\ref{fig10}A,B).
 The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
 We set Ca\textsuperscript{2+} diffusion as \SI{1.1e-6}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
-Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequencies.
-Also, the frequency preference for 10 Hz is reversed at $3\times$ Ca\textsuperscript{2+} influx condition (Fig~\ref{fig10}C,D).
+Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequency.
+Also, the frequency preference for 10 Hz is reversed at the $3\times$ Ca\textsuperscript{2+} influx condition (Fig~\ref{fig10}C,D).
 In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided at $3\times$ becomes sufficiently saturating to change the frequency preference.

### 2017-07-18 paper.tex
07-17-2017 update

@@
 In our study, we focus on one particular pathway that involves calmodulin and Ca\textsuperscript{2+}/calmodulin-dependent protein kinase II (CaMKII).
-We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods.
+We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway, and the method allows us to study the pathway with more details than conventional methods.
 Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence relies on Ca\textsuperscript{2+} input amount, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.
@@
 
-The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings~\cite{}.
-Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase~\cite{}.
+The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings~\cite{Lisman:2002ki,Coultrap:2012ip,Stratton:2013el,Hell:2014bd}.
+Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase~\cite{Lisman:2002ki,Coultrap:2012ip,Stratton:2013el,Hell:2014bd}.
 Activities such as CaM binding and auto-phosphorylation result in various subunit states.
 For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site.
-Once phosphorylated, the CaM unbinding rate decreases dramatcially, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}.
+Once phosphorylated, the CaM unbinding rate decreases dramatically, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}.
 Even when the CaM molecule unbinds, the CaMKII subunit stays activated, entering an autonomous state.
 Finally, Thr305/Thr306 can also be phosphorylated.
-There theronine sites overlap with the CaM binding site and its phosphorylation blocks the binding of CaM molecules.
+There the threonine site overlap with the CaM binding site and its phosphorylation blocks the binding of CaM molecules.
 In this case, the subunit becomes "capped".
@@
 Each subunit would potentially have 20 states.
-Then for the holoenzyme, the total number of unique species reaches over $\frac{6^{20}}{6}=609.36\times10^{12}$.
+Then for the holoenzyme, the total number of unique species reaches over $\frac{6^{20}}{6} \approx 609.36\times10^{12}$.
 It would be extremely time-consuming and not practical for BioNetGen to generate this network.
@@
 
-In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (Ca\textsubscript{V}) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus.
+In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (Ca\textsubscript{V}) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM signaling through the cytoplasm.
 Specifically, we examined Ca\textsuperscript{2+}-CaM-CaMKII network activity near L-type Ca\textsuperscript{2+} channels.
-This is an area less studied previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}.
+This is an area less studied in previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}.
 To study this Ca\textsuperscript{2+}-CaM-CaMKII signaling network and deal with the problem of combinatorial explosion, we modified the published and freely available simulator Smoldyn~\cite{Andrews:2004fs}.
@@
 Reactions are defined between binding sites.
-The reaction network is stored in a hash table to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network.
+The reaction network is stored in a hash table to allow for lookup during simulations when reactants collide.
+Therefore expanding and loading a complete network is not required.
 We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}.
@@
 \subsection*{Testing of the simulator modifications}
-We first tested the modifications to Smoldyn used in this paper (see Methods) using a Ca\textsuperscript{2+}-CaM network (Fig~\ref{fig1}A).
+We first tested the modifications to Smoldyn used in this paper (see Methods) using a Ca\textsuperscript{2+}-CaM network (Fig~\ref{fig1}).
 The 4 Ca\textsuperscript{2+} binding sites on CaM gives rise to a total of 9 different binding states of a CaM molecule.
@@
 \begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
-For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx k_{on}$ as diffusion is sufficiently fast.
-Given the diffusion coefficients for Ca\textsuperscript{2+} and CaM and the parameter values for Ca\textsuperscript{2+} and CaM reaction kinetics in Table S2, we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime.
-This was also true if diffusion constants were reduced 10-fold.
+For diffusion-limited reactions, $k_a \gg k_{enc}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_{on} \approx k_a$ as diffusion is sufficiently fast.
+Given the diffusion coefficients for Ca\textsuperscript{2+} and CaM and the parameter values for Ca\textsuperscript{2+} and CaM reaction kinetics in Table S1, we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime.
+This was also true if diffusion constants were reduced 10 fold.
 Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.
-
-We coded a deterministic ODE system in COPASI~\cite{Hoops:2006gy} and compared results with our stochastic model using the same kinetics.
-We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as shown in Fig~\ref{fig2}.
-Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used.
-These results suggest that our modifications of Smoldyn are working properly.
-
 \begin{figure}[!h]
 	\caption{{\bf Ca\textsuperscript{2+}-CaM interaction network.}
-	A: Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM.
+	(A)Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM.
 CaM has two lobes, an N lobe and a C lobe, each of which has 2 Ca\textsuperscript{2+} binding sites.
@@
 Thus site N2 cannot bind Ca\textsuperscript{2+} unless N1 has already bound Ca\textsuperscript{2+} and a similar rule applies to sites C2 and C1.
-In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" implies binding sites N1 and N2 are both bound with Ca\textsuperscript{2+} and "C2" implies similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" imply only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound.
+In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" represents binding sites N1 and N2 are both bound with Ca\textsuperscript{2+} and "C2" represents similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" represent only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound.
 Arrows represent the direction of binding reactions.
-	B: The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator.
+	(B)The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator.
 The binding and unbinding rate constants are labeled at the end of reactions.
@@
 Notice the "==" sign is used on the left-hand side, but the "=" sign is used on the right-hand side.
-The "==" represents True or False of equality, whereas the "=" denotes an assigning operation.
+The "==" represents True or False of an equality, whereas the "=" denotes an assigning operation.
 	}
@@
 
-\begin{figure}[!h]
-	\caption{{\bf Test of Smoldyn modifications with a simple Ca\textsuperscript{2+}-CaM reaction network.}
-	Plots compare the number of CaM molecules in each of the 9 possible states as computed with the modified stochastic simulator and COPASI.
+We coded a deterministic ODE system in COPASI~\cite{Hoops:2006gy} and compared results with our stochastic model using the same kinetics.
+We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species.
+Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used.
+These results suggest that our modifications of Smoldyn are working properly.
+%	 
+%\begin{figure}[!h]
+%	\caption{{\bf Test of Smoldyn modifications with a simple Ca\textsuperscript{2+}-CaM reaction network.}
+%	Plots compare the number of CaM molecules in each of the 9 possible states as computed with the modified stochastic simulator and COPASI.
 Thick smooth lines are results using COPASI and thin black and red lines are results with the modified simulator for two different values of the calcium diffusion coefficient.
-	}
-\label{fig2}
-\end{figure}
+%	}
+%\label{fig2}
+%\end{figure}
 
 As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complexes.
-As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system with a saturating amount of fully loaded CaM molecules assuming subunit phosphorylation is allowed only from a neighbor that is CaM bound but not from one that is phosphorylated, then at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme.
+As Michalski and Loew~\cite{Michalski:2012ds} indicated that for a closed system, different subunits number of a holoenzyme results in distinct the steady state fraction of phosphorylated subunits, given a saturating amount of fully loaded CaM molecules and assuming that subunit phosphorylation is allowed only from a neighbor that is CaM bound but not from one that is phosphorylated.
 In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numbers, the fraction approaches $1-e^{-1}$.
@@
 We set up a prototype model to identify the major network branches that contribute to the phosphorylation of CaMKII subunits.
-The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig3}A.
-In this network diagram molecule species are represented as vertices and reactions are represented as the network edges.
-CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in a $\SI{1}{\um}\times\SI{1}{\um}\times\SI{2}{\um}$ box.
+The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig2}A.
+In this network diagram molecule species are represented as vertices and reactions are represented as edges.
+CaM molecules and CaMKII holoenzymes with 6-subunit rings were initially uniformly distributed in a $\SI{1}{\um}\times\SI{1}{\um}\times\SI{2}{\um}$ box.
 As an initial condition, the box contained 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII.
 The initial state is at an equilibrium, which we carefully computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules.
-For simplicity, we did not consider resting intracellular Ca\textsuperscript{2+} in the steady state initial condition, as calculations showed its effect to be minor.
-Ca\textsuperscript{2+} influx was modeled as entry from a single source at the top center of the box.
+For simplicity, we did not consider resting intracellular Ca\textsuperscript{2+} are part of the steady state initial condition, as calculations showed its effect to be minor.
+Ca\textsuperscript{2+} influx was modeled as an entry from a single source at the top center of the box.
 A previously generated input file having a total of 6 Ca\textsuperscript{2+} bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods).
@@
 	\caption{{\bf The complete Ca\textsuperscript{2+}-CaM-CaMKII network.}
-	A: Species are represented by vertices.
+	(A)Species are represented by vertices.
 Edges are labeled to indicate binding reactions.
@@
 Red edges mark the predominant pathway.
-B: Layers shaded with blue colors.
-The back layer(Layer1) represents interactions between Ca\textsuperscript{2+} and CaM.
-The front layer(Layer2) represents interactions between Ca\textsuperscript{2+} and CaM attached to a CaMKII subunit.
+	(B)Layers shaded with blue colors.
+The back layer (Layer1) represents interactions between Ca\textsuperscript{2+} and CaM.
+The front layer (Layer2) represents interactions between Ca\textsuperscript{2+} and CaM attached to a CaMKII subunit.
 Layer3 represents interactions between Ca\textsuperscript{2+} and CaM attached to a phosphorylated CaMKII subunit.
 	}
-\label{fig3}
+\label{fig2}
 \end{figure}
 
-We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{fig3}A).
-The network consists of three layers (Fig~\ref{fig3}B).
+We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{fig2}A).
+The network consists of three layers (Fig~\ref{fig2}B).
 The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2).
@@
 For each CaM state, we counted the accumulated occurrences of the possible reaction types.
-Results are shown in Fig~\ref{fig4}.
-The plot reveals a predominant pathway for CaM and CaMKII state transitions as labeled in Fig~\ref{fig3}A.
+Results are shown in Fig~\ref{fig3}.
+The plot reveals a predominant pathway for CaM and CaMKII state transitions as labeled in Fig~\ref{fig2}A.
 Starting as apoCaM, a CaM molecule tends to bind a Ca\textsuperscript{2+} ion on the C1 site and then on the C2 site, entering the N0C2 state.
-CaM in the N0C2 state has a strong preference to enter layer 2 by binding to a CaMKII subunit, after which it quickly binds Ca\textsuperscript{2+} ions to the N1 and N2 sites to become KN2C2.
-Also evident from Fig~\ref{fig4} is that once CaM has two Ca\textsuperscript{2+} ions on either lobe, the preference is to bind to a CaMKII subunit before binding additional Ca\textsuperscript{2+} ions.
+CaM in the N0C2 state has a strong preference to enter Layer2 by binding to a CaMKII subunit, after which it quickly binds Ca\textsuperscript{2+} ions to the N1 and N2 sites to become KN2C2.
+Also evident from Fig~\ref{fig3} is that once CaM has two Ca\textsuperscript{2+} ions on either lobe, the preference is to bind to a CaMKII subunit before binding additional Ca\textsuperscript{2+} ions.
 Once bound to a CaMKII subunit, additional Ca\textsuperscript{2+} ions bind to CaM more quickly.
 In this scenario, it is rare for CaM to become fully bound with Ca\textsuperscript{2+} ions prior to binding with a CaMKII subunit.
-Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{fig5}.
-We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S2 chosen to be consistent with experimental or modeling studies~\cite{Pepke:2010ju,Zeng:2010bq,Lucic:2008gt}.
+Nevertheless, phosphorylation of CaMKII subunits occurs most often when subunits are bound with fully loaded CaM, but still often with CaM having C sites loaded, KN0C2 and KN1C2, as shown in Fig~\ref{fig4}.
+We note that the way the predominant pathway is chosen at each vertex is a consequence of reaction affinities as given in Table S1 chosen to be consistent with experimental or modeling studies~\cite{Pepke:2010ju,Zeng:2010bq,Lucic:2008gt}.
 For example, even though N sites bind Ca\textsuperscript{2+} faster, the C sites have higher affinity making Ca\textsuperscript{2+} binding to C sites preferred.
@@
 	}
-\label{fig4}
+\label{fig3}
 \end{figure}
@@
 	}
-\label{fig5}
+\label{fig4}
 \end{figure}
 
-To confirm the critical role of NxC2 (x=0,1,2) CaM in activation and phosphorylation of CaMKII, we set up two modified reaction schemes (Fig~\ref{fig6}A).
+To confirm the critical role of NxC2 (x=0,1,2) CaM in activation and phosphorylation of CaMKII, we set up two modified reaction schemes (Fig~\ref{fig5}A).
 In Scheme1, only CaMKII subunits bound with NxC2 are allowed to become phosphorylated.
@@
 Note the phosphorylation rates in the two schemes are equivalent, i.e., $k_{on}$ of reaction KpN2C0\_rxn is the same as that of KpN0C2\_rxn, and the $k_{on}$ of KpN2C1\_rxn equals that of KpN1C2\_rxn.
-Not surprisingly, the two schemes give rise to different phosphorylation levels as shown in Fig~\ref{fig6}B.
-Scheme 1 performs slightly worse than the whole network, whereas Scheme 2 produces much lower phosphorylation levels.
+Not surprisingly, the two schemes give rise to different phosphorylation levels as shown in Fig~\ref{fig5}B.
+Scheme 1 performs slightly worse than the whole network, whereas Scheme 2 produces much lower phosphorylation level.
 Therefore, edges of the network are not equally involved in CaMKII phosphorylation.
@@
 	\caption{{\bf Reaction scheme 1 and 2 adapted from the complete network.}
-	A: In reaction scheme 1, phosphorylation was allowed only from CaMKII bound with NxC2 (blue paths). In reaction scheme 2, phosphorylation was allowed only from CaMKII bound with N2Xc (green paths).
-	B.
-Comparison of phosphorylation levels with the complete network, scheme 1 and scheme 2.
+	(A)In reaction scheme 1, phosphorylation was allowed only from CaMKII bound with NxC2 (blue paths). In reaction scheme 2, phosphorylation was allowed only from CaMKII bound with N2Cx (green paths).
+	(B)Comparison of phosphorylation levels with the complete network, scheme 1 and scheme 2.
 The results of the complete network were summarized from 15 trials of simulations and presented as mean phosphorylation +/- standard deviation.
 	} 
-\label{fig6}
+\label{fig5}
 \end{figure}
@@
 To confirm this CaMKII activation pattern, we tested our 6-subunit CaMKII model using previously generated 10 Hz and 5 Hz Ca\textsuperscript{2+} influx files.
-The total number of Ca\textsuperscript{2+} ions entering was comparable (40433 ions with 5 Hz vs 40435 ions with 10 Hz).
-As shown in Fig~\ref{fig7}, we observed much higher phosphorylation levels with 10 Hz than with 5 Hz input.
+The total number of Ca\textsuperscript{2+} ions entering was comparable in the two cases (40433 ions with 5 Hz vs 40435 ions with 10 Hz).
+As shown in Fig~\ref{fig6}, we observed much higher phosphorylation levels with 10 Hz than with 5 Hz input.
 
@@
 	\caption{{\bf Comparison of 10 Hz and 5 Hz on CaMKII activation.}
-	A: Number of bound CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input conditions.
-	B: Number of phosphorylated CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	(A)Number of bound CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input conditions.
+	(B)Number of phosphorylated CaMKII subunits (including bound and phosphorylated) with 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	}
+\label{fig6}
+\end{figure}
+
+We examined the reaction occurrences over time and noticed that compared to 5 Hz input the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig7}A,B), and consequently more autophosphorylation (Fig~\ref{fig7}C,D).
+Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
+However, the frequency effect can be reversed by providing a saturating amount of Ca\textsuperscript{2+}.
+In Fig~\ref{fig8}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential pulse.
+The total Ca\textsuperscript{2+} influx was again comparable for 5 Hz and 10 Hz.
+As Ca\textsuperscript{2+} influx was increased, the network produced more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminished.
+Eventually, the 10 Hz input became saturating and the network generated less phosphorylation with 10 Hz than with 5 Hz input (Fig~\ref{fig8}F).
+\begin{figure}[!h]
+	\caption{{\bf Effects of 10 Hz and 5 Hz on reaction occurrences.} 
+	(A)Accumulated reaction occurrences for CaM CaMKII binding reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
+	(B)The same as in A but for 10 Hz Ca\textsuperscript{2+} input.
+	(C)Accumulated reaction occurrences for CaMKII phosphorylation reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
+	(D)The same as in C but for 10 Hz Ca\textsuperscript{2+} input.
 	}
@@
 
-We examined the reaction occurrences over time and noticed that compared to 5 Hz input the 10 Hz input results in more bindings between N0C2 CaM and CaMKII (Fig~\ref{fig8}A,B), and consequently more autophosphorylations (Fig~\ref{fig8}C,D).
-Therefore the frequency effect is inherent to the binding between Ca\textsuperscript{2+}, CaM and CaMKII.
-However, the frequency effect can be reversed by providing a saturating amount of Ca\textsuperscript{2+}.
-As demonstrated in Fig~\ref{fig9}, we increased the Ca\textsuperscript{2+} channel number to allow more Ca\textsuperscript{2+} influx per action potential pulse.
-The total Ca\textsuperscript{2+} influx was again comparable for 5 Hz and 10 Hz.
-As Ca\textsuperscript{2+} influx was increased, the network produced more phosphorylated CaMKII subunits for both input frequencies, but the phosphorylation level difference between 5 Hz and 10 Hz diminished.
-Eventually, the 10 Hz input became saturating and the network generated less phosphorylation with 10 Hz than with 5 Hz input (Fig~\ref{fig9}F).
 \begin{figure}[!h]
-	\caption{{\bf Effects of 10 Hz and 5 Hz on reaction occurrences.} 
-	A: Accumulated reaction occurrences for CaM CaMKII binding reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
-	B: The same as in A but for 10 Hz Ca\textsuperscript{2+} input.
-	C: Accumulated reaction occurrences for CaMKII phosphorylation reactions in the presence of 5 Hz Ca\textsuperscript{2+} input. 
-	D: The same as in C but for 10 Hz Ca\textsuperscript{2+} input.
+	\caption{{\bf Effects of the various Ca\textsuperscript{2+} influx on frequency dependence.}
+	(A-E)Accumulated reaction occurrences for phosphorylation of CaMKII subunits bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) with different levels of Ca\textsuperscript{2+} influx.
+	(F)Summary of phosphorylation levels in the presence of varying amount of Ca\textsuperscript{2+} delivered at 5 Hz and 10 Hz.
+The advantage of 10 Hz on phosphorylation is reversed as Ca\textsuperscript{2+} influx increases.
 	}
@@
 
+Equivalently, a reversal in frequency preference is also seen when available CaM is limited or Ca\textsuperscript{2+} diffusion is slowed (Fig~\ref{fig9}).
+To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
+Limiting CaM reduced phosphorylation for input at both frequencies.
+However even though the Ca\textsuperscript{2+} input did not change, limiting CaM resulted in more phosphorylation reactions with 5 Hz than with 10 Hz Ca\textsuperscript{2+} input (Fig~\ref{fig9}A,B).
+A similar effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
+We set Ca\textsuperscript{2+} diffusion as \SI{1.1e-6}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
+Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequency.
+However, the frequency preference for 10 Hz is reversed at the $3\times$ Ca\textsuperscript{2+} influx condition (Fig~\ref{fig9}C,D).
+In other words, with slow diffusion, Ca\textsuperscript{2+} influx increased 3 fold becomes sufficiently saturating to change the frequency preference.
+It is also worth noting that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice the amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig9}E,F).
+
 \begin{figure}[!h]
-	\caption{{\bf Effects of various Ca\textsuperscript{2+} influx on frequency dependence.}
-	A-E: Accumulated reaction occurrences for phosphorylation of CaMKII subunits bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) with different levels of Ca\textsuperscript{2+} influx.
-	F: Summary of phosphorylation levels in the presence of varying amount of Ca\textsuperscript{2+} delivered at 5 Hz and 10 Hz.
-The advantage of 10 Hz on phosphorylation is reversed as Ca\textsuperscript{2+} influx increases.
+	\caption{{\bf Effects of CaM availability and Ca\textsuperscript{2+} diffusion on frequency dependence.}
+	(A)With $3\times$ Ca\textsuperscript{2+} influx, accumulated reaction occurrences for CaMKII phosphorylation bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
+	(B)The same as in A except that the total CaM available in the system is decreased to half. 
+	(C)The same as in A except that $3.5\times$ Ca\textsuperscript{2+} influx is provided.
+	(D)The same as in C except that Ca\textsuperscript{2+} diffusion is slowed to \SI{1.1e-6}{\square\cm\per\s}. 
+	(E)The same as in A except that $2\times$ Ca\textsuperscript{2+} influx is provided.
+	(F)The same as in E except with both slowed Ca\textsuperscript{2+} diffusion and half the amount of CaM.
 	}
@@
 
-Equivalently, a reversal in frequency preference is also seen when available CaM is limited or Ca\textsuperscript{2+} diffusion is slowed (Fig~\ref{fig10}).
-To demonstrate, we compared simulations using $3.5\times$ Ca\textsuperscript{2+} input with the default CaM (\SI{5}{\micro\Molar}) and limited CaM (\SI{2.5}{\micro\Molar}).
-Even though the Ca\textsuperscript{2+} input did not change, limiting CaM results in more phosphorylation reactions with 5 Hz than with 10 Hz Ca\textsuperscript{2+} input (Fig~\ref{fig10}A,B).
-The same effect can be achieved with slowed Ca\textsuperscript{2+} diffusion.
-We set Ca\textsuperscript{2+} diffusion as \SI{1.1e-6}{\square\cm\per\s}, which is half of the default speed and does not change the activation-limited regime of the network.
-Slow diffusion results in a dramatic increase of CaMKII phosphorylation regardless of Ca\textsuperscript{2+} amount and input frequency.
-Also, the frequency preference for 10 Hz is reversed at the $3\times$ Ca\textsuperscript{2+} influx condition (Fig~\ref{fig10}C,D).
-In other words, with slow diffusion, Ca\textsuperscript{2+} influx provided at $3\times$ becomes sufficiently saturating to change the frequency preference.
-It is also worth noting that by combining limited CaM and slow diffusion, reversing the frequency preference is possible with merely twice amount of Ca\textsuperscript{2+} influx (Fig~\ref{fig10}E,F).
+To understand why frequency dependence changed, we examined the reaction occurrences over time on Layer1 edges (Fig~\ref{fig10}).
+We noticed that regardless of the input frequency, the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} levels become saturating.
+In the presence of moderate Ca\textsuperscript{2+} influx (Fig~\ref{fig10}A,B), CaM molecules follow the predominant pathway by choosing reaction paths according to affinities: bindings between N0C2 and CaMKII are more likely to occur than bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules.
+CaMKII binding with N0C2 dominates all types of CaMKII-CaM binding reactions (Fig~\ref{fig10}D,E,G).
+However, when Ca\textsuperscript{2+} influx becomes saturating, bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules start to dominate over the bindings between N0C2 and CaMKII (Fig~\ref{fig10}C).
+CaM molecules tend to stay in Layer1 longer until they get fully loaded with Ca\textsuperscript{2+}.
+As a result, CaMKII binding with N2C2 increases and becomes dominant over all other CaM CaMKII binding reactions (Fig~\ref{fig10}F,H,I).
+This decision change gives rise to an altered pathway and eventually a slower rise phosphorylation level.
+Given the same total amount of Ca\textsuperscript{2+}, 10 Hz influx saturates CaM in a shorter time than 5 Hz.
+Correspondingly, the switch of pathway choice is more dramatic for 10 Hz at $4.5\times$ Ca\textsuperscript{2+}, resulting in the reversed frequency preference as observed previously.
 
 \begin{figure}[!h]
-	\caption{{\bf Effects of CaM availability and Ca\textsuperscript{2+} diffusion on frequency dependence.}
-	A: With $3\times$ Ca\textsuperscript{2+} influx, accumulated reaction occurrences for CaMKII phosphorylation bound with NxC2 in the presence of 5 Hz (thin line) and 10 Hz (thick line) Ca\textsuperscript{2+} input.
-	B: The same as in A except that the total CaM available in the system is decreased to half. 
-	C: The same as in A except that $3.5\times$ Ca\textsuperscript{2+} is provided.
-	D: The same as in C except that Ca\textsuperscript{2+} diffusion is slowed  to \SI{1.1e-6}{\square\cm\per\s}. 
-	E: The same as in A except that $2\times$ Ca\textsuperscript{2+} is provided.
-	F: The same as in E except with slowed Ca\textsuperscript{2+} diffusion and half amount of CaM.
+	\caption{{\bf Change of pathway choice as Ca\textsuperscript{2+} influx increases}
+	(A)With $2\times$ Ca\textsuperscript{2+} input, reaction occurrences of CaM N0C2 binding Ca\textsuperscript{2+} at the N1 site (blue) and directly binding with CaMKII (black).
+Thick lines are for 10 Hz input, and thin lines are for 5 Hz input.
+	(B)The same as in A except that $3\times$ Ca\textsuperscript{2+} influx is given.
+	(C)The same as in A except that $4.5\times$ Ca\textsuperscript{2+} influx is given.
+	(D-F)Accumulated reaction occurrences for CaM CaMKII binding reactions with varying amount of total Ca\textsuperscript{2+} influx from $2\times$, $3\times$ to $4.5\times$.
+With Ca\textsuperscript{2+} input delivered at 5 Hz.
+	(G-I)The same as in D-F except that Ca\textsuperscript{2+} input is delivered at 10 Hz.
 	}
@@
 
-To understand why frequency dependence changed, we examined the reaction occurrences over time on Layer1 edges (Fig~\ref{fig11}).
-We noticed that regardless of the input frequency, the reaction pathway choice deviates from the predominant pathway when Ca\textsuperscript{2+} levels become saturating.
-In the presence of moderate Ca\textsuperscript{2+} influx (Fig~\ref{fig11}A,B), CaM molecules follow the predominant pathway by choosing reaction paths according to affinities: bindings between N0C2 and CaMKII are more likely to occur than bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules.
-CaMKII binding with N0C2 dominates all types of CaMKII-CaM binding reactions (Fig~\ref{fig11}D,E,G).
-However, when Ca\textsuperscript{2+} influx becomes saturating, bindings of Ca\textsuperscript{2+} ions on N1 sites of CaM molecules dominate over the bindings between N0C2 and CaMKII (Fig~\ref{fig11}C).
-CaM molecules tend to stay in the Layer1 untill they get fully loaded by Ca\textsuperscript{2+}.
-As a result, CaMKII binding with N2C2 becomes dominant over all other CaM CaMKII binding reactions (Fig~\ref{fig11}F,H,I).
-This decision change gives rise to an altered pathway and eventually a suboptimal phosphorylation level.
-Given the same total amount of Ca\textsuperscript{2+}, 10 Hz influx saturates CaM in a shorter time than 5 Hz.
-Correspondingly, the switch of pathway choice is more dramatic for 10 Hz at $4.5\times$ Ca\textsuperscript{2+}, resulting in the reversed frequency perference as observed previously.
+To validate our conclusion about the pathway decision change, we used a reduced reaction network to capture the observed frequency preference reversal (Fig~\ref{fig11}).
+The reduced network is derived from the initial steps in the whole Ca\textsuperscript{2+}-CaM-CaMKII network.
+We used the original Smoldyn to simulate the simple network, in order to demonstrate that the reversal of frequency preference is not an artifact of our modified simulator but is inherent to the network itself.
+We used 5 pulses of instantaneous Ca\textsuperscript{2+} release as the input and varied the amount of Ca\textsuperscript{2+} influx per pulse.
+The CaMKII molecules are modeled as monomers with an arbitrary phosphorylation rate of \SI{1}{\per\s}.
+As expected, the reversal of frequency preference can be qualitatively captured by the simple reaction scheme (Fig~\ref{fig11}A).
+The 10 Hz stimulus becomes saturating and fails to generate more phosphorylation than 5 Hz when Ca\textsuperscript{2+} influx reaches 40000 ions per pulse (Fig~\ref{fig11}B).
 
 \begin{figure}[!h]
-	\caption{{\bf Change of pathway choice as Ca\textsuperscript{2+} influx increases}
-	A: With $2\times$ Ca\textsuperscript{2+} input, reaction occurrences of CaM N0C2 binding Ca\textsuperscript{2+} at the N1 site and directly binding with CaMKII.
-Thick lines are for 10 Hz input, and thin lines are for 5 Hz input.
-B: The same as in A except that $3\times$ Ca\textsuperscript{2+} is given.
-C: The same as in C except that $4.5\times$ Ca\textsuperscript{2+} is given.
-D-F: Accumulated reaction occurrences for CaM CaMKII binding reactions under varying amount of total Ca\textsuperscript{2+} influx from $2\times$, $3\times$ to $4.5\times$.
-Ca\textsuperscript{2+} inputs are delivered at 5 Hz.
-G-I: The same as in D-F except that Ca\textsuperscript{2+} inputs are delivered at 10 Hz.
+	\caption{{\bf Change of pathway choice demonstrated using a reduced network}
+	(A)The reduced reaction network derived from the complete network. CaMKII is modeled as a monomer and undergoes phosphorylation at an arbitrary rate \SI{1}{\per\s} as long as bound with a CaM N1C2. 
+	(B)Summary of phosphorylation levels in the presence of varying Ca\textsuperscript{2+} ions per pulse from 10000 ions/pulse to 50000 ions/pulse delivered at 5 Hz and 10 Hz.
+Data from 5 trials are summarized.
 	}
@@
 
-To validate our conclusion about the pathway decision change, we used a reduced reaction network to capture the observed frequency preference reversal (Fig~\ref{fig12}).
-The reduced network is derived from the initial steps in the whole Ca\textsuperscript{2+}-CaM-CaMKII network.
-We used the original Smoldyn to simulate the simple network, in order to demonstrate that the reversal of frequency preference is not an artifact of our modified simulator but is inherent to the network itself.
-We used 5 pulses of instantaneous Ca\textsuperscript{2+} release as the input and varied the amount of Ca\textsuperscript{2+} influx per pulse.
-The CaMKII molecules are modeled as monomers with an arbirary phosphorylation rate of \SI{1}{\per\s}.
-As expected, the reversal of frequency preference can be qualitatively captured by the simple reaction scheme (Fig~\ref{fig12}A).
-The 10 Hz stimulus becomes saturating and fails to generate more phosphorylation than 5 Hz when Ca\textsuperscript{2+} influx reaches 40000 ions per pulse.
+\subsection*{The structural organization of CaMKII subunits}
+
+Recent work on CaMKII holoenzyme structure suggests that how subunits are organized can affect the activation of the holoenzyme.
+In particular, whether subunits are arranged in a compact or an extended way can change the accessibility of CaM to CaMKII.
+It is known that the structural arrangement is related to a linker region between a subunit's kinase domain and the holoenzyme central hub.
+Bayer et al.~\cite{Bayer:2002er} examined splice variants of $\beta$-CaMKII, which have identical kinase domains yet different linker lengths.
+These variants responded to Ca\textsuperscript{2+} oscillations differently, even though they showed no response difference to prolonged Ca\textsuperscript{2+}-CaM input.
+In particular, for Ca\textsuperscript{2+} pulse input, the variants with longer linker length exhibited a higher autophosphorylation rate.
+Another study by Chao et al.~\cite{Chao:2011iw} also indicated that the linker length affects the capability of a subunit's kinase domain to undock from the central hub.
+Undocking helps the subunit to release from an autoinhibited state.
+Thus a longer linker is expected to keep the kinase domain further away from the central hub, allowing a better chance for Ca\textsuperscript{2+}-CaM to access the binding site.
+
+In our prototype model, holoenzyme subunits have distinct physical locations and are separated by \SI{8}{\nm} from its neighbors.
+To examine the effect of a longer linker length on CaMKII activation, we varied the radius of a one-ring 6-subunit holoenzyme from \SI{5}{\nm} to \SI{15}{\nm}.
+Neighbor subunits are equally spaced at the same distance as the holoenzyme radius.
+Studies suggest that a typical one-ring holoenzyme radius is within \SI{5}{\nm} to \SI{8}{\nm}~\cite{Chao:2011iw,Gaertner:2004jk}.
+For each radius condition, we ran 15 simulation trials using the default Ca\textsuperscript{2+} influx delivered at 5 Hz.
+As shown in Fig~\ref{fig12}, there is a slight but distinguishable increase in the phosphorylation level as the radius of the holoenzyme increases.
 
 \begin{figure}[!h]
-	\caption{{\bf Change of pathway choice demonstracted using a reduced network}
-	A: The reduced reaction network derived from the complete network.
-CaMKII are modeled as monomers and can undergo phosphorylation at an arbitrary rate \SI{1}{\per\s} as long as bound with a CaM N1C2.
-	B: Summary of phosphorylation levels in the presence of varying Ca\textsuperscript{2+} ions per pulse from 10000 ions/pulse to 50000 ions/pulse delivered at 5 Hz and 10 Hz.
-Data from 5 trials were included.
+	\caption{{\bf Effects of holoenzyme size on CaMKII phosphorylation.}
+	Phosphorylation levels as a function of the holoenzyme radius presented as mean + standard deviation. Each radius condition is summarized from 15 trials.
 	}
@@
 
-\subsection*{The structural orgnization of CaMKII subunits}
+\section*{Discussion}
+In the study, we present an efficient approach to simulate multi-subunit molecules with detailed kinetics in reaction-diffusion networks.
+Our approach is an adaptation of Smoldyn~\cite{Andrews:2004fs}, a particle-based stochastic simulator that allows reaction and diffusion to be simulated in spatially heterogenous environments.
+Our adaptation adds new data structures to Smoldyn to describe reactions at the binding site level, leading to an efficient solution of the problem of combinatorial explosion inherent in models with multi-subunit molecules~\cite{Stefan:2014gl}.
+We also introduced an intuitive approach to analyzing the pathway choices of a network based on the reaction history of a simulation, allowing us to grasp insights quickly about how reactions procced in a large network.
+In traditional analyses, the numbers of each molecular species are examined to resolve the dynamics of the network.
+However this procedure is indirect at best when it comes to drawing insights about network behavior.
+Furthermore, being able to track reaction history of a system is not possible in deterministic simulations.
+Here we were able to count the number of times a reaction occurs easily and we demonstrated that this reaction history can be highly informative and can allow us to see clearly how the favored network pathway may change with different input conditions.
 
-Recent work on CaMKII holoenzyme structure suggests that how subunits are organized can affect the activation of the holoenzyme.
-In particular, whether subunits are arranged in a compact or an extended way can change the accessibility of CaM to CaMKII.
-It is known that the structural arrangement is related to a linker region between a subunit's kinase domain and the holoenzyme central hub.
-Bayer et al.~\cite{Bayer:2002er} examined splice variants of $\beta$-CaMKII.
-These variants have identical kinase domains yet different linker lengths.
-They responded to Ca\textsuperscript{2+} oscillations differently, even though they showed no difference to prolonged Ca\textsuperscript{2+}-CaM input.
-In particular, for Ca\textsuperscript{2+} pulse input, the variants with longer-linker length exhibited a higher autophosphorylation rate.
-Another study by Chao et al.~\cite{Chao:2011iw} also indicated that the linker length affects the capability of a subunit's kinase domain to undock from the central hub.
-Undocking helps the subunit to release from an autoinhibited state.
-Thus a longer linker can keep the kinase domain further away from the central hub, allowing a better chance for Ca\textsuperscript{2+}-CaM to access the binding site.
+We used the modified Smoldyn simulator and reaction history information to obtain several insights into the Ca\textsuperscript{2+}-CaM-CaMKII reaction network.
+First, under physiological conditions when Ca\textsuperscript{2+} influx is low to moderate, CaM molecules partially loaded with Ca\textsuperscript{2+} are important for CaMKII activation.
+In particular, reaction history shows that CaM molecules that have 2 Ca\textsuperscript{2+} ions attached on the C lobe (in particular species N0C2, but also N1C2 and N2C2) preferentially bind to CaMKII subunits before adding additional Ca\textsuperscript{2+} ions to the N lobe.
+This is consistent with the predominant pathway hypothesis suggested by Pepke et al.~\cite{Pepke:2010ju} as well as with experimental work by Shifman et al.~\cite{Shifman:2006hw}.
+Nevertheless, phosphorylation was found to occur primarily from CaMKII bound with CaM fully loaded with Ca\textsuperscript{2+}(KN2C2) and higher frequencies of Ca\textsuperscript{2+} input resulted in more CaMKII phosphorylation.
 
-In our prototype model, holoenzyme subunits have distinct physical locations and are separated by \SI{8}{\nm} from its neighbors.
-To examine such a structural effect on CaMKII activation, we varied the radius of an one-ring shaped holoenzyme from \SI{5}{\nm} to \SI{15}{\nm}.
-The holoenzyme consists of 6 equally spaced subunits.
-Neighbor subunits are spaced the same length as the holoenzyme radius.
-According to the literature~\cite{Chao:2011iw,Gaertner:2004jk}, a typical one-ring holoenzyme radius is within \SI{5}{\nm} to \SI{8}{\nm}.
-For each radius condition, we tested 15 simulation trials using the default Ca\textsuperscript{2+} influx delivered at 5 Hz.
-As shown in Fig~\ref{fig13}, there is a slight but distinguishable increase in phosphorylation level as the radius of the holoenzyme increases.
+Second, while CaMKII activation is known to be sensitive to the frequency of Ca\textsuperscript{2+} signals~\cite{DeKoninck:1998wh}, we found that the frequency dependence is reversed with stronger Ca\textsuperscript{2+} signals with more CaMKII subunit phosphorylation seen at 5 Hz input than at 10 Hz.
+Reaction history shows that this occurs because of a change in the predominant pathway for CaMKII activation in which CaM with 2 Ca\textsuperscript{2+} ions on the C lobe now becomes more likely to bind Ca\textsuperscript{2+} on the N lobe than to bind with CaMKII.
+This is significant because many experiments that study frequency dependence are done in conditions where CaM is saturated with Ca\textsuperscript{2+} and this is rarely the situation encountered by the cell.
 
-\begin{figure}[!h]
-	\caption{{\bf Effects of holoenzyme size on CaMKII phosphorylation.}
-	Phophorylation levels as a function of the holoenzyme radius presented in mean + standard deviation. Each radius condition is summarized from 15 trials.
-	}
-\label{fig13}
-\end{figure}
+Third, we found that factors such as CaM availability and Ca\textsuperscript{2+} diffusion can also affect the frequency dependence of CaMKII activation by Ca\textsuperscript{2+} signals, also by changing the predominant pathway for CaMKII activation.
+A limited amount of CaM makes the given Ca\textsuperscript{2+} input more likely to saturate available CaM on both lobes before binding to CaMKII.
+Similarly, slow Ca\textsuperscript{2+} diffusion allows more extensive interactions between Ca\textsuperscript{2+} and CaM thus making it more likely for a given amount of Ca\textsuperscript{2+} input to become saturating.
+Experimental studies suggest that the number of freely diffusible CaM molecules is highly limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl} and limited CaM further implies a regulatory role for the many endogenous CaM binding proteins~\cite{Skene:1990kf}.
+A limited amount of CaM or a slowed Ca\textsuperscript{2+} diffusion may permit enough Ca\textsuperscript{2+} to bind to available CaM and allow CaMKII activation to occur at lower frequency Ca\textsuperscript{2+} signals.
 
-\section*{Discussion}
-In the study, we presented an efficient approach to simulate multi-subunit molecules and reaction network in detailed kinetics, as well as the effects of diffusion on the reaction network.
-Our adaption of the Smodlyn preserves its particle-based simulation feature, allowing us to include diffusion and spatial heterogeneity in models.
-Also, we added new data structures to describe reactions at the binding sites level, leading to an efficient solution of combinatorial explosion.
-We also introduced an intuitive approach to analyzing the pathway choices of a network based on the reaction history of a simulation, allowing us to grasp insights of a large network quickly.
-In traditional analysis, the number of molecule species are examined to resolve the dynamics of the network.
-However, the method is indirect when it comes to draw insights of the network behaviors.
-Also, being able to track the reaction history of the system is not possible in deterministic simulations such as ODE based ones.
-Here we demonstrated that the reaction history can be highly informative and allows us to see clearly a favored pathway of a network under different input conditions.
+Fourth, it is known that intracellular crowding and spatial homogeneity can slow down molecule diffusion.
+For example, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.
+We believe that such a restriction in Ca\textsuperscript{2+} diffusion may have substantial effects on CaMKII phosphorylation and the frequency dependence.
+For example, depending on the size of the nandomain, slow Ca\textsuperscript{2+} diffusion in a nanodomain can potentially result in a localized Ca\textsuperscript{2+} signal with sufficient strength to activate downstream cascading proteins.
 
-This approach allows us to obtain several insights into the Ca\textsuperscript{2+}-CaM-CaMKII netowrk.
-First, under physiological conditions when Ca\textsuperscript{2+} influx is moderate, CaM molecules partially loaded with Ca\textsuperscript{2+} are important for CaMKII activation.
-In particular, CaM molecules that have 2 Ca\textsuperscript{2+} attached on the C lobe (including N0C2, N1C2 and N2C2) are preferentially bound to CaMKII subunits.
-This is consistent with the predominant pathway hypothesis suggested in Pepke et al.~\cite{Pepke:2010ju}.
-Second, we found that factors such as CaM availablility~\cite{DeKoninck:1998wh} and Ca\textsuperscript{2+} diffusion can affect the CaMKII frequency response.
-Limited amount of CaM makes the given Ca\textsuperscript{2+} input more likely to saturate the available CaM.
-Likewise, slow Ca\textsuperscript{2+} diffusion allows more extensive interactions between Ca\textsuperscript{2+} and CaM, thus more likely for the given amount of Ca\textsuperscript{2+} input to become saturating.
-Experimental studies suggest that the number of freely diffusible CaM molecules is higly limited \textit{in vivo}~\cite{2008BpJ....95.6002S,Persechini:2002tb,LubyPhelps:1995kl}.
-The limited CaM further implies a regulatory role for the many endogenous CaM binding proteins.
-For example, Skene et al.~\cite{Skene:1990kf} suggested that a CaM binding protein invovled in axonal grwoth cone GAP-43 can act as a "CaM sponge" by sequesting a fair amount of CaM under submembrane area and releasing these CaM molecules in response to protein kinase C activity, thereby modulating the development of axonal growth cones.
-Third, it is known that intracellular crowding and spatial homogeneity can slow down molecule diffusion.
-For example, a recent biophysical study~\cite{2013PNAS..11015794T} suggests that the diffusion of Ca\textsuperscript{2+} ions can be reduced by ten times in a nanodomain around the Ca\textsuperscript{2+} channel mouth.
-We believe that such a complication in Ca\textsuperscript{2+} diffusion may have substantial effects on CaMKII phosphorylation and frequency response.
-For example, depending on the size of the nandomain, slow Ca\textsuperscript{2+} diffusion in a nanodomain can potentially result in a localized Ca\textsuperscript{2+} signal with sufficient strength to activate downstream cascading proteins.
-Fourth, we demonstrated that the holoenzyme size can affect the level of phosphorylation.
-We increased the size of the holoenzyme by changing the distance between neighbor subunits and found that the phosphorylation level of the netowrk increased slightly accordingly.
-This is consistent with the idea that the configuration of a holoenzyme, whether compact or extended, affects CaM molecules accessing the CaMKII subunits~\cite{Stratton:2013el}.
+Finally, we demonstrated that holoenzyme size can affect the level of phosphorylation.
+We increased the size of the holoenzyme by changing the distance between neighbor subunits and found that the phosphorylation level of the network increased slightly accordingly.
+This is consistent with the idea that the configuration of a holoenzyme, whether compact or extended, affects the ability of CaM molecules to access CaMKII subunits~\cite{Stratton:2013el}.
+The extension may allow a subunit to sample a volume away from other subunits, increasing the possibility of a reaction.
 
-%chao etal, autoinhibited through dimerization;
-%stratton etal, subunit exchange 
-Our model does not contain Thr305/Thr306 phosphorylation (few subunits would have phosphoryated during the timeframe simulated here) and also lacks some newly discovered mechanisms of CaMKII structural features, which may lead to a more complicated activation pattern of CaMKII holoenzymes.
+Our model does not contain Thr305/Thr306 phosphorylation (few subunits would have become phosphorylated during the time frame simulated here) and also lacks some newly discovered CaMKII structural feature mechanisms, which may lead to a more complicated activation pattern of CaMKII holoenzyme subunits.
 For example, it has been found that there exists a compact autoinhibition state, which occurs through dimerization of adjacent subunits from top and bottom rings~\cite{Chao:2010fn}.
-Once Ca\textsuperscript{2+} is bound to a dimerized subunit, the dimer disassembles and the two subunits are popped away from the center of the holoenzyme.
-Another recent study indicated that phosphorylated CaMKII subunits can undergo subunit exchange to faciliate propagating activation triggered by Ca\textsuperscript{2+}-CaM~\cite{Stratton:2014ct}.
+Once Ca\textsuperscript{2+} is bound to a dimerized subunit, the dimer disassembles and the two subunits swing away from the center of the holoenzyme.
+Another recent study indicated that phosphorylated CaMKII subunits can undergo subunit exchange to facilitate propagating activation triggered by Ca\textsuperscript{2+}-CaM~\cite{Stratton:2014ct}.
 The significance of these additional features of CaMKII activation awaits future study.
@@
 One technical challenge for particle-based simulation is to handle diffusion-limited reactions, especially in the presence of highly concentrated molecules.
-One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM act very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in modeling studies (although see ~\cite{Mironov:2013} for a critique of their estimates).
-These binding kinetics, if accurate, would place these reactions in the diffusion-limited regime, rendering traditional mass action based methods inaccurate.
-However for kinetics this fast, adequte simulation options are limited and not efficient.
-If using the original Smoldyn algorithm, the simulation time step has to be considerably reduced to obtain the correct steady state~\cite{Andrews:2015}; alternatively, we may increase the geminiate recombination probability~\cite{Andrews:2004fs}.
+One recent experimental study~\cite{Faas:2011fna} estimated that the N sites of CaM act very fast to bind Ca\textsuperscript{2+}, much faster than previously cited for CaM-N lobe binding kinetics in experimental or modeling studies (although see ~\cite{Mironov:2013} for a critique of these estimates).
+If accurate, these fast binding kinetics would place these reactions in the diffusion-limited regime, rendering traditional mass action based methods inaccurate.
+However for kinetics this fast, adequate simulation options are limited and not efficient.
+If using the original Smoldyn algorithm, the simulation time step would have to be considerably reduced to obtain the correct steady state~\cite{Andrews:2015}; alternatively, one might increase the geminate recombination probability~\cite{Andrews:2004fs}.
 Another software package using an enhanced Green's Function Algorithm~\cite{vanZon:2005jd} can handle the high concentration diffusion-limited reactions accurately, but it takes an impractically long time to simulate.
@@
 A hash table is a data structure that stores association arrays and allows rapid lookup.
-In our case, the reaction network can be considered as associations between reactants, therefore are ideal to be implemented using a hash table.
+In our case, the reaction network can be considered as associations between reactants, and therefore are ideal to be implemented using a hash table.
 CaM is an example of a molecule with multiple binding sites.
 The binding reactions involving the N and C lobes of a CaM can be coded as in Fig~\ref{fig1}B.
-A CaMKII holoenzyme is an example of a complex composed of two 6-subunits rings.
+A CaMKII holoenzyme is an example of a complex composed of two 6-subunit rings.
 Each subunit is a molecule containing binding sites for CaM and phosphorylation.
 Each ring has a radius of \SI{8}{\nm} and is separated from its direct neighbors at a fixed distance of \SI{8}{\nm} (estimated from~\cite{Gaertner:2004jk}).
-For simplicity, we modeled CaMKII holoenzymes as one ring of 6 subunits.
+For simplicity, we usually modeled CaMKII holoenzymes as one ring of 6 subunits.
 
-In the original version of Smoldyn, each reaction generates new molecules as products and reactant molecules are removed.
+In the original version of Smoldyn, each reaction generates new molecules and reactant molecules are removed.
 In our case, since one molecule can have multiple binding sites and is potentially associated with multiple partners, entirely removing a molecule is not practical because other attached molecules would also be affected.
@@
 Therefore, during reactions we do not remove molecules, but merely change molecule binding states and positions.
-Molecules bound together physically overlap, ynchronize their locations automatically and diffuse together.
+Molecules bound together physically overlap, synchronize their locations automatically and diffuse together.
 The diffusion coefficient is determined by the dominant molecule.
-For example, when Ca\textsuperscript{2+} and CaM are bound, the attached Ca\textsuperscript{2+} molecule diffuse with the CaM diffusion rate; similarly, CaM bound to a CaMKII subunit will diffuse with the CaMKII.
+For example, when Ca\textsuperscript{2+} and CaM are bound, the attached Ca\textsuperscript{2+} molecule diffuses with the CaM diffusion rate; similarly, CaM bound to a CaMKII subunit will diffuse with CaMKII.
 
@@
 Since the N and C sites act independently, the N1 and C1 sites compete for Ca\textsuperscript{2+} and an apoCaM N0C0 can become either N1C0 or N0C1, resulting in a branching reaction scheme.
-Thus a decision process is needed to choose a reaction path when a binding event occurs.
+Thus a decision process is needed to choose a reaction path when such a binding event occurs.
 
@@
 Rxn2 has similar rate constants $k_{f2}$ and $k_{b2}$.
-The two reactions can be viewed together as an equivalent rxn3, which has overall kinetics rates $k_{f3}$ and $k_{b3}$.
+The two reactions can be viewed together as an equivalent rxn3, which has overall kinetic rates $k_{f3}$ and $k_{b3}$.
 According to the law of mass action, the reactions can be written as differential equations and we can obtain
@@
 If the distance between a molecule pair is smaller than $r_3$, binding happens.
-To make a reaction choice, we generated a uniformly distributed random number from 0 to 1.
+To make a reaction choice, we generate a uniformly distributed random number from 0 to 1.
 If the number falls in the range $(0,\frac{kf_1}{kf_3}]$, then rxn1 is chosen; instead if the number falls in the range $(\frac{kf_1}{kf_3}, 1]$, we pick rxn2.
@@
 We first used a Ca\textsuperscript{2+}-CaM network for testing to confirm that modifications to the simulator were working properly.
-Then we added CaMKII holoenzymes to study the reaction network in details.
-We set up a cube-shaped model to represent a portion of a cell body (Fig~\ref{fig14}A).
-The cube has dimensions of \SI{1}{\um} in width and \SI{2}{\um} in depth.
+Then we added CaMKII holoenzymes to study the reaction network in detail.
+We set up a cube-shaped model to represent a portion of a cell body.
+The cube has dimensions of \SI{1}{\um} in width and length, as well as \SI{2}{\um} in depth.
 The top surface of the cube represents the cell membrane, reflective to all molecules.
 The four sides are also reflective.
-The bottom surface is partially absorbing to Ca\textsuperscript{2+} ions but reflective to CaM.
+The bottom surface is partially absorbing to Ca\textsuperscript{2+} ions but reflective to CaM and CaMKII.
 This conservation of CaM molecules and CaMKII subunits guarantees a steady state initial condition.
@@
 
+Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type) are located on the top surface to provide Ca\textsuperscript{2+} influx.
+For simplicity, these channels are placed together at the center of the membrane.
+The channels open and close depending on a time-varying membrane voltage file generated from a neuron model (described below).
+CaMKII subunits are uniformly present at a concentration of \SI{10}{\micro\Molar}.
+They are also immobilized, presumably are attached to actin~\cite{Li:2016cq}.
+Freely diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}.
+This is consistent with the notion that at the resting level, freely diffusible CaM molecules are considerably limited in number compared to their binding proteins~\cite{Tran:2003fs,2008BpJ....95.6002S,LubyPhelps:1995kl}.
+Table S1 lists all the reactions with corresponding kinetic parameters involved in the network.
+Kinetic parameters are integrated from various sources as noted in Table S1 and are adjusted to satisfy microscopic reversibility.
+
+\subsection*{Ca\textsuperscript{2+} input conditions}
+
+A neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology of a CA1 pyramidal cell and ion channel conductances to generate a voltage response at the soma to various stimulation conditions.
+Theta-burst stimulation (5 pulses at 100 Hz, repeated 5 times at 5 Hz or 10 Hz intervals) was applied to synapses on spines taking into account a probability of release measured in experiments~\cite{Grover:2009hb}.
+This stimulation activated AMPA and NMDA receptor-channels on dendritic spines causing depolarization in the dendritic tree, which propagated to the soma and initiated action potentials.
+Voltage profiles for 5 Hz and 10 Hz interval stimulation are shown in Fig~\ref{fig13}A,C.
+
+\begin{figure}[!h]
+	\caption{{\bf Patterns of membrane voltage and Ca\textsuperscript{2+} influx. }
+	(A)A 5 Hz action potential burst voltage file generated from the NEURON model. 
+	(B)Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file.
+Ca\textsuperscript{2+} influx is stochastic.
+There is a spontaneous influx in the absence of action potentials.
+	(C)10 Hz action potential burst voltage file generated from the NEURON model. 
+	(D)Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. 
+	(E)Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} input files.
+	}
+\label{fig13}
+\end{figure}
+
+This membrane voltage output from the neuron model was used to determine Ca\textsuperscript{2+} influx through L-type channels in our reaction network model.
+Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast and their density is low, the membrane potential is little affected by their activity.
+Thus the neuron model and the reaction network model can be safely decoupled.
+L-type Ca\textsuperscript{2+} channels are modeled stochastically.
+They open and close in response to the voltage input.
+The voltage-dependent opening and closing of these channels are modeled with the Hodgkin-Huxley formalism~\cite{Tuckwell:2012tt}.
+The rates of channel opening and closing are functions of membrane voltage and are calculated using variables $n$ and $\tau_n$, where $n$ describes voltage-dependent activation and $\tau_n$ is the time constant of $n$ (Fig~\ref{fig14}A).
+The channel is assumed to obey a simple first order kinetics as follows.
+
 \begin{figure}[!h]
 	\caption{{\bf Model geometry and Ca\textsuperscript{2+} channel parameters.}
-	A: Dimensions of the cube-shaped model. Ca\textsuperscript{2+} channels are located on the top surface. 
-	B: $n_{inf}(V)$ and $\tau_n(V)$ used in the voltage-gated Ca\textsuperscript{2+} channels.
-	C: The number channels N by fitting the single channel current to the GHK equation. N was found to be 16. 
+	(A)$n_{inf}(V)$ and $\tau_n(V)$ used in the voltage-gated Ca\textsuperscript{2+} channels.
+	(B)The number channels N by fitting the single channel current to the GHK equation. N was found to be 16. 
 	}
@@
 
-Voltage-gated Ca\textsuperscript{2+} channels (presumably L-type) are located on the top surface to provide Ca\textsuperscript{2+} influx.
-For simplicity, these channels are placed together at the center of the membrane.
-The channels open and close depending on a time-varying membrane voltage file generated from a neuron model (described below).
-CaMKII subunits (typically $\beta$-subunits) are uniformly present at a concentration of \SI{10}{\umol}.
-They are also immobilized so presumlably they attached to actins~\cite{Li:2016cq}.
-Freely diffusible CaM molecules are uniformly distributed at a concentration of \SI{5}{\micro\Molar}.
-This is consistent with the notion that at the resting level, freely diffusible CaM molecules are considerably limited in number compared to their binding proteins~\cite{Tran:2003fs,2008BpJ....95.6002S,LubyPhelps:1995kl}.
-Table S2 lists all the reactions with corresponding kinetic parameters involved in the network.
-Kinetic parameters are integrated from various sources as noted in Table S2 and are adjusted to satisfy microscopic reversibility.
-
-\subsection*{Ca\textsuperscript{2+} input conditions}
-
-A neuron model was constructed using NEURON~\cite{Carnevale:2006iv} with a detailed morphology and ion channel conductances.
-Its membrane voltage output was applied to the Ca\textsuperscript{2+} reaction network model.
-Since the kinetics of L-type Ca\textsuperscript{2+} with detailed morphology of a CA1 pyramidal cell and ion channel conductances to generate a voltage response at the soma to various stimulation conditions.
-Theta-burst stimulation (5 pulses at 100 Hz, repeated 5 times at 5 Hz or 10 Hz intervals) was applied to synapses on spines taking into account a probability of release measured in experiments~\cite{Grover:2009hb}.
-This stimulation activated AMPA and NMDA receptor-channels on dendritic spines causing depolarizations in the dendritic tree, which propagated to the soma and initiated action potentials.
-Voltage profiles for 5 Hz and 10 Hz intervals stimulations are shown in Fig~\ref{fig15}A,C.
-
-\begin{figure}[!h]
-	\caption{{\bf Patterns of membrane voltage and Ca\textsuperscript{2+} influx. }
-	A: A 5 Hz action potential burst voltage file generated from the NEURON model. 
-	B: Ca\textsuperscript{2+} influx generated using the 5 Hz action potentials voltage file.
-Ca\textsuperscript{2+} influx is stochastic.
-There is spontaneous influx in the absence of action potentials.
-	C: 10 Hz action potential burst voltage file generated from the NEURON model. 
-	D: Ca\textsuperscript{2+} influx generated using the 10 Hz action potential file. 
-	E: Accumulated Ca\textsuperscript{2+} influx for 5Hz, 10 Hz, 8 and 16 channels respectively. The total Ca\textsuperscript{2+} amount is matched between Ca\textsuperscript{2+} input files.
-	}
-\label{fig15}
-\end{figure}
-
-This membrane voltage output for neuron model was used to determine Ca\textsuperscript{2+} influx through L-type channels in our reaction network model.
-Since the kinetics of L-type Ca\textsuperscript{2+} channels are relatively fast and their density is low, the membrane potential is not affected by their activities.
-Thus the neuron model and the reaction network model can be safely decoupled.
-L-type Ca\textsuperscript{2+} channels are modeled stochastically and they open and close in response to the voltage input.
-The voltage-depdendent opening and closing of these channels are modeled with the Hudgkin-Huxley formation~\cite{Tuckwell:2012tt}.
-The rates of channel opening and closing are functions of membrane voltage and calculated using variables $n$ and $\tau_n$, where $n$ describes voltage-dependent activation and $\tau_n$ is the time constant of $n$ (Fig~\ref{fig16}B).
-The channel obeys a simple first order kinetics as follows.
-
 \begin{quote}
@@
 \end{quote}
-Then the following set of equations are used to decribe voltage-gated Ca\textsuperscript{2+} channels:
+%\begin{equation}dn(V)=\frac{dn-n_{\inf}}{dt}\end{equation}
+
+Then the following set of equations are used to describe voltage-gated Ca\textsuperscript{2+} channel kinetics:
 \begin{equation}rate_{open}(V)=\frac{n_{\inf}}{\tau_n}\end{equation}
 \begin{equation}rate_{close}(V)=\frac{1-n_{\inf}}{\tau_n}\end{equation}
-\begin{equation}n_{\inf}=\frac{1}{1+exp(-\frac{V-V_{1/2}}{slope})}\end{equation}
-\begin{equation}\tau_n=0.06+4\times\frac{0.75\times\sqrt{0.45\times0.55}}{exp((V-V_{1/2})\times0.55/slope)+exp(-(V-V_{1/2})\times0.45/slope)}\end{equation}
+\begin{equation}n_{\inf}=\frac{1}{1+\exp(-\frac{V-V_{1/2}}{slope})}\end{equation}
+\begin{equation}\tau_n=0.06+4\times\frac{0.75\times\sqrt{0.45\times0.55}}{\exp((V-V_{1/2})\times0.55/slope)+\exp(-(V-V_{1/2})\times0.45/slope)}\end{equation}
 
-In particular, $n_{\inf}$ is the steady state value of $n$.
-$V_{1/2}$ and $slope$ together describe the channel activation in response to voltage.
+where $n_{\inf}$ is the steady state value of $n$, and $V_{1/2}$ and $slope$ together describe the channel activation in response to voltage.
 In our model, $V_{1/2}$ equals \SI{-15}{\mV} and $slope$ equals 8.
@@
 The $rate_{open}$ and $rate_{close}$ are used to calculate conditional probabilities to determine the state of a channel for the next time step in the following way
-\begin{equation}P(C|O)=1-exp(-rate_{close}(V)dt)\end{equation}
-\begin{equation}P(O|C)=1-exp(-rate_{open}(V)dt)\end{equation}
+\begin{equation}P(C|O)=1-\exp(-rate_{close}(V)dt)\end{equation}
+\begin{equation}P(O|C)=1-\exp(-rate_{open}(V)dt)\end{equation}
 \begin{equation}P(O|O)=1-P(C|O)\end{equation}
@@
 To calculate how many ions, we used the following equation~\cite{2013PNAS..11015794T} to obtain the unitary current for a single channel
-\begin{equation}i_{ca}=-g(V-V_s)\frac{\exp\frac{-(V-V_s)}{RT/zF}}{1-\exp\frac{-(V-V_s)}{RT/zF}}\end{equation}
-where $g$ is chosen as \SI{5}{\pico\siemens} and $RT/zF$ equals \SI{12}{\milli\volt}.
-A current density is calculated using GHK current equation~\cite{citeulike:306134} assuming an extracellular [Ca\textsuperscript{2+}] of \SI{2}{\m\Molar} and an intracellular concentration of \SI{50}{\nano\Molar}.
-The maximum channel permeation is \SI{0.241e-3}{\cm\per\s}.
-Since the membrane surface is \SI{1}{\square\um}, the current density is converted to a total current $I_{ghk}$ for the area.
-On the other hand, assuming a total number of N channels are present, for each channel $\frac{I_{ghk}}{N}$ a single channel current is $i_{ca}$.
-By fitting the total current $I_{ghk}$ and $N\times i_{ca}$, we obtained a N of 16 and a $V_s$ of \SI{-1.91}{\milli\volt} (Fig~\ref{fig15}C).
-Using the $i_{ca}$, the number of ions entering one channel during one time step is calcuated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{fig15}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
-To guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig15}E).
+\begin{equation}i_{ca}=-g(V-V_s)\frac{\exp(\frac{-(V-V_s)}{RT/zF})}{1-\exp(\frac{-(V-V_s)}{RT/zF})}\end{equation}
+where $g$ is chosen as \SI{5}{\pico\siemens} and $RT/zF$ equals \SI{12}{\milli\volt}, and $V_s$ is determined as described below.
+A current density is calculated using GHK current equation as follows,
+\begin{equation}I=P\frac{Vz^2F^2}{RT}\frac{[Ca^{2+}]_i-[Ca^{2+}]_o\exp(-zVF/RT)}{1-\exp(-zVF/RT)}\end{equation}
+
+where an extracellular [Ca\textsuperscript{2+}]\textsubscript{o} equals \SI{2}{\m\Molar}, an intracellular concentration [Ca\textsuperscript{2+}]\textsubscript{i} equals \SI{50}{\nano\Molar} and a maximum membrane permeability to Ca\textsuperscript{2+} P is \SI{0.241e-3}{\cm\per\s}.
+Since the membrane surface is \SI{1}{\square\um}, the current density is converted to a total current $I_{ghk}$ for this area.
+If a total of N channels are present, in this membrane surface the single channel current $i_{ca}$ equals $\frac{I_{ghk}}{N}$.
+By fitting the total current $I_{ghk}$ with $N\times i_{ca}$, we obtained an N of 16 and a $V_s$ of \SI{-1.91}{\milli\volt} (Fig~\ref{fig14}B).
+From $i_{ca}$, the number of ions entering each open channel during one time step is calculated as $\frac{i_{ca}}{2e}dt$ (Fig~\ref{fig13}B,D), where 2 is the valence of Ca\textsuperscript{2+} and $e$ is the elementary charge.
+To guarantee a consistent amount of total Ca\textsuperscript{2+} for a given set of 5 Hz and 10 Hz voltage files, we generated 40 trials of Ca\textsuperscript{2+} influx files for each frequency and then selected the ones with equivalent total Ca\textsuperscript{2+} influx (Fig~\ref{fig13}E).
 
@@
 \label{S1_Table}
-{\bf The total amount of Ca\textsuperscript{2+} influx of selected files for 5 Hz and 10 Hz.} 
-\paragraph*{S2 Table.}
-\label{S2_Table}
 {\bf Kinetic parameters of all reactions.}
-\paragraph*{S3 Table.}
-\label{S3_Table}
-{\bf Reaction characterizations.}
+\paragraph*{S2 File.}
+\label{S2_File}
+{\bf Reaction network configuration file.}
 

### 2017-07-19 paper.tex
pg8

@@
 {\Large
-\textbf\newline{Simulating detailed Ca\textsuperscript{2+}-calmodulin-CaMKII network in a new way: biophysical attributes that affect CaMKII activation pattern} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Simulating a detailed Ca\textsuperscript{2+}-calmodulin-CaMKII network in a new way: biophysical attributes that affect the CaMKII activation pattern} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 Ximing Li\textsuperscript{1},
-William Holmes\textsuperscript{1}
+William R. Holmes\textsuperscript{1}
 \\
@@
 Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons.
-Understanding how this occurs has been the focus of many modeling studies.
-However, CaMKII is notoriously difficult simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled.
-To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion.
+Understanding how this occurs has been the focus of numerous studies including many that use models.
+However, CaMKII is notoriously difficult to simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled.
+To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator, Smoldyn, to manage the problem of combinatorial explosion.
 With this new method, spatial and temporal aspects of the signaling network can be studied without compromising biochemical details.
 We used this new method to examine how calmodulin molecules, both partially loaded and fully loaded with Ca\textsuperscript{2+}, choose pathways to interact with and activate CaMKII under various Ca\textsuperscript{2+} input conditions.
-We found that the frequency dependence of the CaMKII phosphorylation is intrinsic to the network kinetics and the activation pattern can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion.
+We found that the CaMKII phosphorylation dependence on Ca\textsuperscript{2+} signal frequency is intrinsic to the network kinetics and the activation pattern can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion.
 Depending on whether Ca\textsuperscript{2+} influx is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns.
@@
 In our study, we focus on one particular pathway that involves calmodulin and Ca\textsuperscript{2+}/calmodulin-dependent protein kinase II (CaMKII).
-We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway, and the method allows us to study the pathway with more details than conventional methods.
+We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway, and the method allows us to study the pathway with more detail than conventional methods.
 Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence relies on Ca\textsuperscript{2+} input amount, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.
@@
 \section*{Introduction}
-Calcium/calmodulin-dependent protein kinase II (CaMKII) is an important enzyme widely distributed in the central nervous system and other tissues including cardiac muscles~\cite{Wayman:2008gla,Lisman:2002ki,Erickson:2014fs}.
+Calcium/calmodulin-dependent protein kinase II (CaMKII) is an important enzyme widely distributed in the central nervous system and other tissues including cardiac muscle~\cite{Wayman:2008gla,Lisman:2002ki,Erickson:2014fs}.
 CaMKII is involved in intracellular Ca\textsuperscript{2+} signaling as an effector of the Ca\textsuperscript{2+} sensor protein calmodulin (CaM)~\cite{Herring:2016bh,Hell:2014bd,Coultrap:2012ip}.
-In neurons, CaMKII molecules are essential in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}.
+In neurons, CaMKII molecules play an essential role in long-term potentiation (LTP), our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}.
 In spines, activated CaMKII molecules interact with postsynaptic density proteins, facilitating actins to reorganize, leading to spine enlargement and upregulation of $alpha$-amino-3-hydroxy-5-methyl-4-isoxazolepropionic acid (AMPA) receptor numbers~\cite{Herring:2016bh}.
 CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}.
-In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus.
-Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors.
+In addition to synaptic functions, recent studies suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus~\cite{Ma:2014dr,Li:2016cq}.
+Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, and participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors.
 Therefore CaMKII molecules play a key role in excitation-transcription coupling.
@@
 The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings~\cite{Lisman:2002ki,Coultrap:2012ip,Stratton:2013el,Hell:2014bd}.
-Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase~\cite{Lisman:2002ki,Coultrap:2012ip,Stratton:2013el,Hell:2014bd}.
+Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with a CaM binding site and phosphorylation sites, and a catalytic domain to act as a kinase~\cite{Lisman:2002ki,Coultrap:2012ip,Stratton:2013el,Hell:2014bd}.
 Activities such as CaM binding and auto-phosphorylation result in various subunit states.
-For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site.
+For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286 site (or Thr287 depending on the subtype).
 Once phosphorylated, the CaM unbinding rate decreases dramatically, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}.
@@
 Finally, Thr305/Thr306 can also be phosphorylated.
-There the threonine site overlap with the CaM binding site and its phosphorylation blocks the binding of CaM molecules.
+There the threonine sites overlap with the CaM binding site and their phosphorylation blocks the binding of CaM molecules.
 In this case, the subunit becomes "capped".
@@
 Software that adopts a rule-based approach such as BioNetGen~\cite{Hlavacek:2006iq} can be used to expand the network based on a set of given reaction rules.
-But for CaMKII holoenzyme, using BioNetGen to generate the network is still computationally intensive.
-After expansion with BioNetGen, a 6-subunit holoenzyme with four states for each subunit contains 700 unique species and 12192 unique reactions.
+But for a CaMKII holoenzyme, using BioNetGen to generate the network is still computationally intensive.
+After expansion with BioNetGen, a 6-subunit holoenzyme with four states for each subunit has 700 unique species and 12192 unique reactions.
 The size of the network grows substantially when more detailed kinetics are involved.
@@
 
-In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (Ca\textsubscript{V}) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM signaling through the cytoplasm.
+In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (Ca\textsubscript{V}) provide the Ca\textsuperscript{2+} influx, as this will serve our future work on modeling Ca\textsuperscript{2+}-CaM signaling through the cytoplasm.
 Specifically, we examined Ca\textsuperscript{2+}-CaM-CaMKII network activity near L-type Ca\textsuperscript{2+} channels.
@@
 Therefore expanding and loading a complete network is not required.
-We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}.
-Moreover, we added detailed components to the network and examined the frequency dependence of CaMKII activation.
+We tested and verified these modifications and then created a detailed Ca\textsuperscript{2+}-CaM-CaMKII network to examine the frequency dependence of CaMKII activation.
 We found that slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation.
@@
 This was also true if diffusion constants were reduced 10 fold.
-Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.
+Thus this simple model resembles a well-mixed system and an ODE model will provide a good test standard.
 \begin{figure}[!h]
 	\caption{{\bf Ca\textsuperscript{2+}-CaM interaction network.}
-	(A)Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM.
+	(A) Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM.
 CaM has two lobes, an N lobe and a C lobe, each of which has 2 Ca\textsuperscript{2+} binding sites.
@@
 Thus site N2 cannot bind Ca\textsuperscript{2+} unless N1 has already bound Ca\textsuperscript{2+} and a similar rule applies to sites C2 and C1.
-In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" represents binding sites N1 and N2 are both bound with Ca\textsuperscript{2+} and "C2" represents similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" represent only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound.
+In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" represents binding sites N1 and N2 both bound with Ca\textsuperscript{2+} and "C2" represents similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" represent only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound.
 Arrows represent the direction of binding reactions.
-	(B)The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator.
+	(B) The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator.
 The binding and unbinding rate constants are labeled at the end of reactions.
 ‘+’ represents a binding reaction; ‘$\sim$’ represents molecules that are bound.
-‘\{\}’ on the left-hand side specifies the binding states of the reactant binding sites or the conditions for a reaction to occur.
+‘\{\}’ on the left-hand side specifies the states of the reactant binding sites or the conditions for a reaction to occur.
 On the right-hand side, binding sites involved in the reaction are assigned to new values.
@@
 Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used.
-These results suggest that our modifications of Smoldyn are working properly.
 %	 
@@
 As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complexes.
-As Michalski and Loew~\cite{Michalski:2012ds} indicated that for a closed system, different subunits number of a holoenzyme results in distinct the steady state fraction of phosphorylated subunits, given a saturating amount of fully loaded CaM molecules and assuming that subunit phosphorylation is allowed only from a neighbor that is CaM bound but not from one that is phosphorylated.
-In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numbers, the fraction approaches $1-e^{-1}$.
+Michalski and Loew~\cite{Michalski:2012ds} indicated that for a closed system, given that CaMKII holoenzymes exposed to a saturating amount of fully loaded CaM molecules and assuming that subunit phosphorylation is allowed only from a neighbor that is CaM bound but not from one that is phosphorylated, the steady state fraction of phosphorylated subunits depends on the number of subunits in a holoenzyme.
+In particular, for holoenzymes as dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numbers, the fraction approaches $1-e^{-1}$.
 We built a model using a closed system with $\SI{1}{\um}\times\SI{1}{\um}\times\SI{1}{\um}$ geometry.
-Six thousand CaMKII subunits and 6020 fully loaded CaM were included in the system.
-Only the following reactions were allowed: bindings and unbindings between CaM and CaMKII, as well as phosphorylations by neighbor subunits that were bound but not phosphorylated.
-As expected, we obtained steady state phosphorylation levels that match the Michalski and Loew formula.
+A total of 6000 CaMKII subunits and 6020 fully loaded CaM were included in the system.
+CaMKII holoenzymes all had either 2, 3, 6 or 12 subunits.
+Only the following reactions were allowed: bindings and unbindings between CaM and CaMKII subunits, as well as phosphorylations by neighbor subunits that were CaM bound but not phosphorylated.
+As expected, we obtained steady state phosphorylation levels that depended on the number of the subunits in a holoenzyme
+he Michalski and Loew formula.
 
@@
 The two methods showed comparable results.
+The results of all these tests suggest that our modifications of Smoldyn are working properly.
 
@@
 The initial state is at an equilibrium, which we carefully computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules.
-For simplicity, we did not consider resting intracellular Ca\textsuperscript{2+} are part of the steady state initial condition, as calculations showed its effect to be minor.
-Ca\textsuperscript{2+} influx was modeled as an entry from a single source at the top center of the box.
+For simplicity, we did not consider resting intracellular Ca\textsuperscript{2+} as part of the steady state initial condition, as calculations showed its effect to be minor.
+Ca\textsuperscript{2+} influx was modeled as entry from a single source at the top center of the box.
 A previously generated input file having a total of 6 Ca\textsuperscript{2+} bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods).
@@
 	\caption{{\bf The complete Ca\textsuperscript{2+}-CaM-CaMKII network.}
-	(A)Species are represented by vertices.
+	(A) Species are represented by vertices.
 Edges are labeled to indicate binding reactions.
@@
 Red edges mark the predominant pathway.
-	(B)Layers shaded with blue colors.
+	(B) Layers shaded with blue colors.
 The back layer (Layer1) represents interactions between Ca\textsuperscript{2+} and CaM.
@@
 However, binding to a phosphorylated CaMKII subunit requires a bound subunit to be phosphorylated first and then lose the bound CaM.
-This type of reaction rarely occurs during the early stage of the simulation since phosphorylation is slow and CaM unbinding from a phosphorylated subunit is uncommon.
+This type of reaction rarely occurs during the time period of the simulation since phosphorylation is slow and CaM unbinding from a phosphorylated subunit is uncommon.
 Thus we focus on the first 3 types of reactions.
